---
title: "Why We Do Matching"
subtitle: "A causal-inference deck on confounding, comparability, and design"
author: "Luke Heley"
date: 9 March 2026
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: why_matching-revealjs.html
    theme: [white, workshop-theme.scss]
    slide-level: 2
    slide-number: c/t
    progress: true
    controls: true
    overview: true
    transition: fade
    center: false
    width: 1400
    height: 850
    footer: '<a href="../index.html">Back to site</a>'
    preview-links: auto
execute:
  echo: false
  eval: false
---

## Session Promise

::: {.kicker}
This short deck explains why matching exists before it explains any specific matching algorithm.
:::

- start from the counterfactual problem
- show how confounding distorts raw comparisons
- explain why matching is a design-stage attempt to improve comparability
- keep the limits explicit throughout [@cunningham2021mixtape; @huntingtonklein_matching]

## 20-Minute Arc

| Time | Segment | Purpose |
|---|---|---|
| `0-4` | Counterfactual problem | Why causal inference needs a comparison group |
| `4-8` | Benchmark comparison | Why randomization is the clearest comparison |
| `8-12` | Confounding | Why raw observational comparisons fail |
| `12-16` | Matching logic | What matching is trying to repair |
| `14-18` | Assumptions and limits | What matching cannot do |
| `18-20` | Practical takeaway | How to use matching responsibly |

## Start With The Fundamental Problem

- each unit has a treated potential outcome and an untreated potential outcome
- we only observe one of them for any given unit
- the missing outcome is the counterfactual
- causal inference is the problem of constructing a credible stand-in for that missing outcome [@cunningham2021mixtape]

## Why We Need A Comparison Group

::: {.card-grid}
::: {.mini-card}
**Observed outcome**

What happened under the treatment state actually received.
:::

::: {.mini-card}
**Missing outcome**

What would have happened under the alternative treatment state.
:::

::: {.mini-card}
**Design task**

Build a comparison group that is credible enough to stand in for the missing outcome.
:::
:::

## Best Benchmark: Randomization

- if treatment is randomized, the treated and untreated groups are comparable in expectation
- that is why `RCT`s are the clearest benchmark for causal comparison quality
- but `RCT`s are a benchmark, not magic: attrition, noncompliance, and external validity still matter [@heiss2026_randomization_matching_content]

## Why Matching Enters The Story

- in many policy settings, treatment was not randomized
- the comparison group is therefore not automatically credible
- matching matters because it is a second-best design repair when stronger design features are unavailable
- the job is to recover the best comparison the data can still defend [@cunningham2021mixtape; @huntingtonklein_matching]

## Dataset We Use: `lalonde`

- the lab and report use `MatchIt::lalonde`, built from the National Supported Work training study with a non-experimental comparison group
- this dataset became a benchmark because LaLonde showed that many non-experimental estimators missed the experimental training effect [@lalonde1986evaluating]
- Dehejia and Wahba made it a matching case study by showing that better design choices could recover more credible estimates [@dehejia1999causal]
- Abadie and Imbens kept it central in the matching literature because it is a disciplined test of how matching estimators behave when the benchmark is known [@abadie2011biascorrected]

::: {.small-note}
We use `lalonde` because it is not just convenient. It is a benchmark dataset for asking whether an observational design can get back to something close to the experimental answer.
:::

## Why Raw Comparisons Mislead

- in observational data, treatment is usually related to prior characteristics
- the treated group may differ from the untreated group before treatment starts
- a raw difference in outcomes therefore combines:
  - the effect of treatment
  - the effect of pre-existing differences
- that mixture is the basic confounding problem

## The Confounder Problem

- a confounder is a common cause of treatment and outcome
- confounders create backdoor paths between treatment and outcome
- if those paths remain open, the observed association is not a clean causal effect [@cunningham2021mixtape; @huntingtonklein_matching]

```{r}
#| label: setup-ggdag
#| include: false
#| eval: true
suppressPackageStartupMessages({
  library(dagitty)
  library(dplyr)
  library(ggdag)
  library(ggplot2)
  library(grid)
})

matching_palette <- c(
  Confounder = "#A44A1A",
  `Adjusted confounder` = "#7F6000",
  Treatment = "#275D8A",
  Outcome = "#4B7F52",
  `Background factor` = "#6B5C7A",
  Collider = "#A43D63",
  Mediator = "#B65A1E"
)

slide_plot_background <- "#f7f1e8"
dag_edge_color <- "#112620"

plot_matching_dag <- function(dag, roles, label_x, label_y,
                              xlim = c(0.7, 3.3), ylim = c(0.6, 2.4)) {
  td <- tidy_dagitty(dag)
  edges <- dplyr::filter(td$data, !is.na(to)) |>
    dplyr::mutate(
      dx = xend - x,
      dy = yend - y,
      dist = sqrt(dx^2 + dy^2),
      pad = pmin(0.18, dist / 3),
      x_plot = x + dx / dist * pad,
      y_plot = y + dy / dist * pad,
      xend_plot = xend - dx / dist * pad,
      yend_plot = yend - dy / dist * pad
    )
  nodes <- dplyr::distinct(td$data, name, .keep_all = TRUE) |>
    dplyr::mutate(
      role = unname(roles[name]),
      label_x = unname(label_x[name]),
      label_y = unname(label_y[name])
    )

  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_plot, y = y_plot, xend = xend_plot, yend = yend_plot),
      arrow = arrow(type = "closed", angle = 22, length = unit(14, "pt")),
      linewidth = 2.2,
      color = dag_edge_color,
      lineend = "round"
    ) +
    geom_dag_point(
      data = nodes,
      aes(x = x, y = y, color = role),
      size = 18,
      show.legend = FALSE
    ) +
    geom_label(
      data = nodes,
      aes(x = label_x, y = label_y, label = label, fill = role),
      color = "white",
      fontface = "bold",
      linewidth = 0,
      label.padding = unit(0.18, "lines"),
      show.legend = FALSE
    ) +
    scale_color_manual(values = matching_palette) +
    scale_fill_manual(values = matching_palette) +
    coord_cartesian(xlim = xlim, ylim = ylim, clip = "off") +
    theme_void() +
    theme(
      panel.background = element_rect(fill = slide_plot_background, color = NA),
      plot.background = element_rect(fill = slide_plot_background, color = NA),
      plot.margin = margin(10, 20, 10, 20)
    )
}

backdoor_dag <- dagify(
  outcome ~ treatment + prior_need,
  treatment ~ prior_need,
  exposure = "treatment",
  outcome = "outcome",
  labels = c(
    prior_need = "Prior need",
    treatment = "Treatment",
    outcome = "Outcome"
  ),
  coords = list(
    x = c(prior_need = 1, treatment = 2, outcome = 3),
    y = c(prior_need = 2, treatment = 1, outcome = 1)
  )
)

blocked_backdoor_dag <- dagify(
  outcome ~ treatment + prior_need,
  treatment ~ prior_need,
  exposure = "treatment",
  outcome = "outcome",
  labels = c(
    prior_need = "Condition on prior need",
    treatment = "Treatment",
    outcome = "Outcome"
  ),
  coords = list(
    x = c(prior_need = 1, treatment = 2, outcome = 3),
    y = c(prior_need = 2, treatment = 1, outcome = 1)
  )
)

collider_dag <- dagify(
  earnings ~ programme + admin_capacity,
  application_complete ~ motivation + admin_capacity,
  programme ~ motivation,
  exposure = "programme",
  outcome = "earnings",
  labels = c(
    motivation = "Motivation",
    programme = "Programme",
    application_complete = "Application complete",
    admin_capacity = "Administrative capacity",
    earnings = "Earnings"
  ),
  coords = list(
    x = c(motivation = 1, programme = 2, application_complete = 2, admin_capacity = 3, earnings = 4),
    y = c(motivation = 2, programme = 2, application_complete = 1, admin_capacity = 2, earnings = 2)
  )
)

mediator_dag <- dagify(
  earnings ~ programme + prior_need + skill_gain,
  programme ~ prior_need,
  skill_gain ~ programme,
  exposure = "programme",
  outcome = "earnings",
  labels = c(
    prior_need = "Prior need",
    programme = "Programme",
    skill_gain = "Skill gain after programme",
    earnings = "Earnings"
  ),
  coords = list(
    x = c(prior_need = 1, programme = 2, skill_gain = 3, earnings = 4),
    y = c(prior_need = 2, programme = 2, skill_gain = 1, earnings = 2)
  )
)
```

## Backdoor Path

```{r}
#| label: fig-backdoor-path
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7.6
#| fig-height: 3.9
#| out-width: 78%
#| fig-align: center
plot_matching_dag(
  backdoor_dag,
  roles = c(
    prior_need = "Confounder",
    treatment = "Treatment",
    outcome = "Outcome"
  ),
  label_x = c(prior_need = 1.02, treatment = 2.22, outcome = 2.98),
  label_y = c(prior_need = 1.76, treatment = 1.18, outcome = 0.80)
)
```

- this is the core confounding case: `D <- X -> Y`
- `Prior need` is a common cause of treatment and outcome
- the raw treated-control gap mixes treatment effects with selection [@mixtape_sessions_unconfoundedness]

## What This Means For Design

- if treated units differ from controls on `Prior need`^[`Prior need` means baseline disadvantage or need that existed before treatment, such as poorer health, lower prior earnings, or higher case severity.], the raw outcome gap is confounded
- part of the observed gap then reflects pre-treatment selection, not just treatment
- the direction of bias depends on whether `Prior need` raises or lowers the outcome
- a credible design must block the backdoor path by conditioning on the right pre-treatment variables
- this is the bridge from causal logic to matching logic [@cunningham2021mixtape]

## Block The Backdoor

```{r}
#| label: fig-block-backdoor
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7.6
#| fig-height: 3.9
#| out-width: 78%
#| fig-align: center
plot_matching_dag(
  blocked_backdoor_dag,
  roles = c(
    prior_need = "Adjusted confounder",
    treatment = "Treatment",
    outcome = "Outcome"
  ),
  label_x = c(prior_need = 1.12, treatment = 2.22, outcome = 2.98),
  label_y = c(prior_need = 1.76, treatment = 1.18, outcome = 0.80)
)
```

- once the confounder is conditioned on, the remaining open path is the causal path from treatment to outcome
- regression, matching, subclassification, and weighting all aim at this same design goal
- the methods differ in implementation, not in the underlying causal target [@mixtape_sessions_unconfoundedness]

## Matching In One Sentence

- matching tries to compare treated units with untreated units that are similar on relevant observed pre-treatment covariates
- the aim is to make treatment status less entangled with baseline differences
- matching is therefore a design tool for reducing observed confounding, not a guarantee of identification

## Why Match?

::: {.signal}
Matching is useful because imbalance creates model dependence.
:::

- imbalance -> model dependence -> researcher discretion -> bias
- raw mismatch forces the outcome model to extrapolate
- extrapolation creates analyst discretion about specification
- analyst discretion can move the estimate even when the causal story is unchanged
- matching trims or down-weights the least credible comparisons first [@heiss2026_randomization_matching_content; @ho2007matching]

## Why Matching Appeals In Practice

::: {.card-grid}
::: {.mini-card}
**Visible comparisons**

The analyst has to show who is being compared to whom.
:::

::: {.mini-card}
**Less hidden extrapolation**

The design leans less on functional-form assumptions alone.
:::

::: {.mini-card}
**Explicit diagnostics**

Balance, overlap, retention, and weights become reportable design costs.
:::
:::

## Design In Two Steps

::: {.lab-grid}
::: {.lab-card}
**1. Preprocess the design**

- define the estimand
- choose covariates from the assignment story
- inspect imbalance and overlap
- match or weight before estimating
:::

::: {.lab-card}
**2. Estimate on the redesigned data**

- use the matched sample or weights as the design
- estimate the treatment effect after preprocessing
- report uncertainty that reflects the adjusted design
:::
:::

## From Subclassification To Matching

- subclassification is the clearest starting point: compare treated and untreated units within the same covariate-defined strata
- exact matching is the strict version of that logic
- approximate matching relaxes exact equality when exact strata become too sparse
- the underlying idea stays the same: compare like with like [@cunningham2021subclassification]

## The Curse Of Dimensionality

- once many covariates matter at once, exact strata proliferate
- some cells become tiny or empty
- overlap disappears for parts of the treated group
- matching methods exist because the causal logic is simple, but the data rarely support simple exact comparisons everywhere [@cunningham2021subclassification]

## Good Controls, Not All Controls

- matching should target plausible pre-treatment confounders
- not every available baseline variable belongs in the design
- the right question is causal: which variables help block confounding without opening new problems?
- this is where causal reasoning matters most
- a pre-treatment common cause belongs in the design
- a collider does not become safe just because it was measured before treatment
- a post-treatment mediator does not belong in the matching design [@mixtape_sessions_unconfoundedness]

## Collider Warning

```{r}
#| label: fig-collider-warning
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7.6
#| fig-height: 3.9
#| out-width: 78%
#| fig-align: center
plot_matching_dag(
  collider_dag,
  roles = c(
    motivation = "Background factor",
    programme = "Treatment",
    application_complete = "Collider",
    admin_capacity = "Background factor",
    earnings = "Outcome"
  ),
  label_x = c(
    motivation = 1.00,
    programme = 2.05,
    application_complete = 2.08,
    admin_capacity = 3.00,
    earnings = 4.02
  ),
  label_y = c(
    motivation = 2.32,
    programme = 2.30,
    application_complete = 0.76,
    admin_capacity = 2.32,
    earnings = 2.30
  ),
  xlim = c(0.7, 4.3),
  ylim = c(0.55, 2.55)
)
```

- a collider is different from a confounder
- if you condition on `Application complete`, you can open a non-causal path between treatment and outcome through upstream factors
- this is why not every pre-treatment variable belongs in the design [@mixtape_sessions_unconfoundedness]

## Post-Treatment Mediator Warning

```{r}
#| label: fig-mediator-warning
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7.6
#| fig-height: 3.9
#| out-width: 78%
#| fig-align: center
plot_matching_dag(
  mediator_dag,
  roles = c(
    prior_need = "Confounder",
    programme = "Treatment",
    skill_gain = "Mediator",
    earnings = "Outcome"
  ),
  label_x = c(prior_need = 1.00, programme = 2.06, skill_gain = 3.02, earnings = 4.00),
  label_y = c(prior_need = 2.32, programme = 2.30, skill_gain = 0.74, earnings = 2.30),
  xlim = c(0.7, 4.3),
  ylim = c(0.55, 2.55)
)
```

- if you condition on a post-treatment mediator, you block part of the treatment effect
- that changes the estimand away from the total programme effect
- mediators therefore do not belong in the design stage when the target is a total causal effect

## What Matching Is Buying

| Problem | What matching tries to improve |
|---|---|
| Baseline imbalance | More comparable treated and control units |
| Opaque regression dependence | A clearer comparison rule |
| Weak visibility on overlap | Explicit evidence on where support fails |
| Hidden design costs | Retention and weighting costs that can be reported plainly |

## Matching And Weighting Are Siblings

- both are design-stage responses to the same comparison problem
- matching changes who remains in the analytic sample
- weighting changes how much each unit counts in the final comparison
- many matching procedures can also be understood as weighted averages rather than as mysterious algorithms [@huntingtonklein_matching; @heiss2026_randomization_matching_content]

## One Logic, Several Implementations

::: {.card-grid}
::: {.mini-card}
**Regression**

Compares outcomes conditional on observed covariates.
:::

::: {.mini-card}
**Matching**

Builds more comparable treated-control pairs or samples first.
:::

::: {.mini-card}
**Stratification and weights**

Take within-cell differences and aggregate them with weights.
:::
:::

::: {.small-note}
All are trying to estimate causal contrasts after blocking backdoor paths with observed covariates.
:::

## Matching Is Still About Assumptions

::: {.metric-grid}
::: {.metric-card}
**Conditional independence**

No important unmeasured confounders remain after conditioning.
:::

::: {.metric-card}
**Common support**

Comparable treated and untreated units must exist.
:::

::: {.metric-card}
**Balance**

Observed covariates should actually look similar after adjustment.
:::

::: {.metric-card}
**Stable estimand**

The design should still answer the question you started with.
:::
:::

## What Matching Cannot Do

- it cannot fix omitted confounding
- it cannot rescue a design with no overlap
- it cannot justify bad controls
- it cannot turn observational data into randomized data
- it should not be the default when a stronger design is available

## When Not To Match

- do not default to matching when a credible `RDD`, `DiD`, or randomized design is available
- do not force matching when overlap is so weak that only severe pruning or extreme weights make the comparison work
- do not pretend a matched design answers the original question if the estimand has drifted to a narrower subgroup

## Why We Still Do It

- many important programme and policy questions are observational
- regression alone can hide the comparison problem rather than expose it
- matching forces the analyst to defend the counterfactual more directly
- even when imperfect, it can make a weak design more transparent and less model-dependent

## Practical Standard

- define the estimand first
- choose covariates from the assignment story
- inspect raw imbalance and overlap
- match or weight as a design step
- estimate only after the adjusted design is credible enough to defend
- follow the main design with uncertainty checks and sensitivity checks
- report the remaining limitations directly

## What The Lab Focuses On

::: {.lab-grid}
::: {.lab-card}
**We focus on**

- exact matching as the transparency benchmark
- coarsened exact matching as the main approximate matching method
- entropy balancing as the weighting comparison
:::

::: {.lab-card}
**We do not center**

- nearest-neighbor propensity-score matching in the core lab
- optimal, full, genetic, or other advanced matching variants
- the full `MatchIt` method menu on a first pass
:::
:::

::: {.small-note}
This narrower scope keeps the design rule visible. For the broader implementation menu, see the [MatchIt documentation](https://kosukeimai.github.io/MatchIt/).
:::

## Closing Line

::: {.signal}
We do matching to reduce observed confounding by building a more credible comparison group, while remaining honest that causality still depends on assumptions the data cannot prove.
:::
