---
title: "Synthetic Control Solver Equivalence"
subtitle: "Quality-assurance report for the CMake-free implementation"
date: "2026-09-23"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-summary: "Show analysis code"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
results <- read.csv("results/acceptance.csv", check.names = FALSE)
weights <- read.csv("results/weights.csv", check.names = FALSE)
paths <- read.csv("results/paths.csv", check.names = FALSE)
inputs <- read.csv("results/inputs.csv", check.names = FALSE)
build_tools <- read.csv("results/build-tools.csv", check.names = FALSE)

fmt <- function(x, digits = 6) formatC(x, digits = digits, format = "fg", flag = "#")
overall_pass <- all(results$overall_pass)
```

## QA conclusion

::: {.callout-tip}
## `r if (overall_pass) "PASS" else "FAIL"`: numerical equivalence demonstrated

For the two benchmark designs, the base-R implementation and `Synth` produce
the same synthetic-control result within the stated tolerances. Both
implementations satisfy the simplex constraints, the base-R result is exactly
repeatable, and every acceptance check passes.
:::

The strongest applied test is Proposition 99. Its largest difference between
the two synthetic paths is only
`r fmt(results$max_abs_path_difference[results$scenario == "Proposition 99"])`
cigarette-sales units. The mean post-treatment gap differs by
`r fmt(results$abs_mean_post_gap_difference[results$scenario == "Proposition 99"])`,
and pre-treatment MSPE differs by
`r fmt(100 * results$relative_pre_mspe_difference[results$scenario == "Proposition 99"], 4)`%.
These differences are numerical solver tolerances, not a change in the
estimated counterfactual or substantive conclusion.

## What was compared

The comparison holds the statistical specification fixed and changes only the
solver:

1. The treated and donor design matrices are identical.
2. Each predictor and pre-treatment outcome row is divided by its standard
   deviation across the treated unit and donors.
3. Every standardized row receives equal importance.
4. Donor weights must be non-negative and sum to one.
5. The base-R implementation uses `stats::constrOptim()`; the reference uses
   `Synth::synth()` with the same fixed predictor-importance vector.

This is the exact objective implemented in the book. The comparison does not
mix it with `Synth`'s optional nested search for a different predictor-importance
vector. That would be a different estimator specification, not a solver test.

The reference solve used `Synth` 1.1-10 with `Margin.ipop = 1e-6`,
`Sigf.ipop = 8`, and `Bound.ipop = 100`. Raising the reference solver's accuracy
above its defaults makes the comparison commensurate with the tight convergence
tolerance used by `constrOptim()`.

## Acceptance criteria

The automated runner fails unless every scenario meets all of these conditions:

| Check | Acceptance threshold |
|---|---:|
| Base-R optimizer convergence | code `0` |
| Donor-weight sum error | less than `1e-10` |
| Minimum donor weight | greater than `-1e-10` |
| Exact-repeat maximum weight difference | less than `1e-12` |
| Relative pre-treatment MSPE difference | less than 1% |
| Maximum synthetic-path difference | less than 0.25 outcome units |
| Mean post-treatment gap difference | less than 0.05 outcome units |

## Results

```{r}
display <- data.frame(
  Scenario = results$scenario,
  Donors = results$donor_count,
  `Maximum weight difference` = fmt(results$max_abs_weight_difference),
  `Relative pre-MSPE difference` = paste0(fmt(100 * results$relative_pre_mspe_difference, 4), "%"),
  `Maximum path difference` = fmt(results$max_abs_path_difference),
  `Mean post-gap difference` = fmt(results$abs_mean_post_gap_difference),
  `All checks` = ifelse(results$overall_pass, "PASS", "FAIL"),
  check.names = FALSE
)
knitr::kable(display, align = c("l", "r", "r", "r", "r", "r", "c"))
```

The Proposition 99 test includes 38 donors and therefore exercises a materially
larger simplex than the six-donor teaching example. The largest donor-weight
difference in that test is
`r fmt(results$max_abs_weight_difference[results$scenario == "Proposition 99"])`;
the weight RMSE is
`r fmt(results$weight_rmse[results$scenario == "Proposition 99"])`.

### Synthetic paths

```{r, fig.width=9, fig.height=4.6}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
for (scenario_name in unique(paths$scenario)) {
  scenario <- paths[paths$scenario == scenario_name, ]
  y_range <- range(
    scenario$treated,
    scenario$reference_synthetic,
    scenario$package_free_synthetic
  )
  plot(
    scenario$time,
    scenario$treated,
    type = "l",
    lwd = 1.5,
    col = "grey35",
    ylim = y_range,
    xlab = "Time",
    ylab = "Outcome",
    main = scenario_name
  )
  lines(scenario$time, scenario$reference_synthetic, col = "#2468A2", lwd = 2, lty = 2)
  lines(scenario$time, scenario$package_free_synthetic, col = "#C43C39", lwd = 1.5, lty = 3)
  legend(
    "topright",
    legend = c("Treated", "Synth reference", "Base R"),
    col = c("grey35", "#2468A2", "#C43C39"),
    lty = c(1, 2, 3),
    lwd = c(1.5, 2, 1.5),
    bty = "n",
    cex = 0.8
  )
}
par(old_par)
```

The reference and package-free paths overlap visually. The numerical table is
the primary evidence; the plot is a reasonableness check rather than the basis
for the verdict.

### Pointwise numerical differences

```{r, fig.width=9, fig.height=4.6}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
for (scenario_name in unique(paths$scenario)) {
  scenario <- paths[paths$scenario == scenario_name, ]
  plot(
    scenario$time,
    scenario$absolute_path_difference,
    type = "h",
    lwd = 2,
    col = "#5B4B8A",
    xlab = "Time",
    ylab = "Absolute difference",
    main = scenario_name
  )
  abline(h = 0.25, col = "#C43C39", lty = 2)
}
par(old_par)
```

All pointwise differences remain far below the 0.25-unit acceptance threshold.

## Constraint and repeatability checks

```{r}
constraint_display <- data.frame(
  Scenario = results$scenario,
  `Weight-sum error` = fmt(results$package_free_weight_sum_error),
  `Minimum weight` = fmt(results$package_free_min_weight),
  `Repeat-run maximum difference` = fmt(results$repeatability_max_abs_weight_difference),
  `Optimizer code` = results$package_free_convergence,
  `Constraint check` = ifelse(results$pass_constraints, "PASS", "FAIL"),
  `Repeatability check` = ifelse(results$pass_repeatability, "PASS", "FAIL"),
  check.names = FALSE
)
knitr::kable(constraint_display, align = c("l", "r", "r", "r", "r", "c", "c"))
```

Both base-R fits converged, their weights are non-negative and sum to one, and
an immediate repeat produces bit-for-bit identical weights.

## Dependency boundary

The book implementation uses only base R's `stats` package for optimization.
It does not import `Synth`, `optimx`, or `nloptr`. `Synth` and its dependencies
exist only in the isolated QA environment used to establish the reference
result.

```{r}
tool_display <- data.frame(
  Tool = build_tools$tool,
  `Available during QA run` = ifelse(build_tools$available, "Yes", "No"),
  check.names = FALSE
)
knitr::kable(tool_display, align = c("l", "c"))
```

The reference QA run itself completed with CMake, Cargo, and Rust absent. The
reference package was installed against the system NLopt library; none of these
reference dependencies are included in the book.

## Data integrity and environment

```{r}
knitr::kable(inputs, align = c("l", "l"), col.names = c("Input", "SHA-256"))
```

The test ran under R 4.5.1 on Linux x86_64. Full package and platform details
are retained in [`results/session-info.txt`](results/session-info.txt).

## Reproduction

The complete test is implemented in
[`run-equivalence.R`](run-equivalence.R). It writes the machine-readable
acceptance results, donor weights, synthetic paths, input hashes, build-tool
inventory, serialized comparison objects, and session information to
[`results/`](results/).

Run it from the repository root in an isolated reference environment containing
`Synth`:

```bash
Rscript qa/synthetic-control-equivalence/run-equivalence.R .
```

The command exits with a non-zero status if any acceptance condition fails.

## Scope and limitations

- The tests cover the toy mechanics example and the full Proposition 99 design.
- They validate the estimator specification used in this book: equal importance
  across standardized predictor and pre-treatment outcome rows.
- They do not claim equivalence for every possible synthetic-control dataset,
  alternative constraint, or `Synth` optimizer option.
- They do not claim that equal row importance is universally preferable to a
  nested predictor-importance search. That is a design choice separate from the
  dependency and solver question tested here.

Within this stated scope, the CMake-free implementation reproduces the reference
synthetic controls to substantially tighter tolerances than the acceptance
limits and is suitable for the book.
