---
title: "How to Do Regression Discontinuity Design"
subtitle: "Sharp and fuzzy cutoff designs, from assignment rules to local effects in R"
format: html
execute:
  echo: true
  warning: false
  message: false
---

Regression discontinuity (RD) uses a discontinuity in treatment assignment at a
known threshold. Units just below and just above the cutoff can provide a local
comparison when their untreated potential outcomes would otherwise vary smoothly
with the assignment score. The design comes from the rule, not from finding a
break in an outcome plot.

This guide walks through a sharp design, its diagnostics, and a compact fuzzy
extension. It assumes basic regression and R. Every example uses
[data bundled in the repository](../../labs/data.qmd); no data are fetched during
execution. The [sharp RD lab](../../labs/regression-discontinuity-foundations-lab.qmd)
provides the companion Quarto document, R notebook and offline ZIP.

## 1. Establish the assignment mechanism

Identify the running variable, cutoff, eligibility rule, actual treatment receipt,
outcome, and population. Determine when the score was measured and who could
influence it. Verify which side receives treatment: a coefficient defined as
right minus left is not always a treated-minus-untreated effect.

In a **sharp** design, crossing the cutoff changes treatment deterministically.
In a **fuzzy** design, it changes the probability of treatment. A weak jump in
receipt is not a sharp design with inconvenient noncompliance; it changes the
estimand and adds instrumental-variable assumptions [@gertler2016ch6rdd;
@cunningham2021rdd; @huntingtonklein2025rdd].

The sharp example concerns Uruguay's PANES programme. The supplied score is
centered at the eligibility threshold; participation is on the **negative** side.
The outcome `Support` takes values 0, 0.5, and 1. It is a government-support score,
not a binary probability. The treatment is access to a programme package, not a
cleanly isolated cash payment [@manacorda2011transfers].

RD requires continuity of relevant potential-outcome means at the threshold and
no other discontinuous intervention that would explain the outcome jump. Sorting,
score manipulation, selective observation, and coincident eligibility rules can
undermine that comparison. A large sample or an attractive curve does not resolve
those threats.

## 2. Audit the score and treatment coding

```{r}
source("../../labs/data/load-data.R")
library(rdrobust)
library(rddensity)
library(ggplot2)
gt <- as.data.frame(qed_data("gov_transfers"))
stopifnot(nrow(gt) == 1948,
          !anyNA(gt[c("Income_Centered", "Participation", "Support")]),
          !any(gt$Income_Centered == 0),
          all(gt$Participation == as.integer(gt$Income_Centered < 0)))
with(gt, table(participation=Participation, below=Income_Centered < 0))
colSums(is.na(gt))
```

The outcome survey is already restricted to roughly 0.02 on either side. That
restriction is a feature of the supplied data, not evidence that 0.02 is an optimal
analysis bandwidth. Record the available range, number of observations, distinct
score values, and any heaping. Missing education values should not silently remove
otherwise usable outcomes from the main analysis.

The cutoff is fixed by the assignment mechanism. Do not search across candidate
cutoffs and then present the most pronounced break as a prespecified policy rule.
When the original score is not centered, subtract the known cutoff once and verify
that eligibility still agrees with its sign.

## 3. Plot before fitting the effect model

```{r}
gt$side <- as.integer(gt$Income_Centered >= 0)
gt$bin <- cut(gt$Income_Centered, breaks=seq(-0.02,0.02,by=0.001),
              include.lowest=TRUE)
binned <- aggregate(cbind(Income_Centered, Support, Participation) ~ bin + side,
                     data=gt, FUN=mean)
ggplot(binned, aes(Income_Centered, Support, color=factor(side))) +
  geom_point(size=2) + geom_vline(xintercept=0, linetype=2) +
  scale_color_manual(values=c("#177b72", "#245ca4"), guide="none") +
  labs(x="Centered assignment score", y="Mean government-support score") +
  theme_minimal(base_size=12)
```

Plot the receipt probability as well as the outcome if the assignment mechanism
is in doubt. Binning helps reveal the data; it is not the estimator. A visual bin
width and a regression bandwidth are separate choices. Keep bins from straddling
the cutoff, expose the available support, and avoid making an apparent jump depend
on a convenient axis range.

Do not fit one high-order global polynomial through the entire sample simply
because it bends towards both clouds of points. Local low-order fits make the
comparison near the threshold explicit and reduce dependence on distant scores.

## 4. Recover a local jump transparently

For a first calculation, use a fixed half-width $h=0.01$, a triangular kernel,
and a separate linear slope on each side. The right-minus-left jump is the
coefficient on the right-side indicator when score zero is the intercept.

```{r}
h <- 0.01
local <- subset(gt, abs(Income_Centered) < h)
local$weight <- 1 - abs(local$Income_Centered)/h
manual <- lm(Support ~ side * Income_Centered, data=local, weights=weight)
jump <- unname(coef(manual)["side"])
left <- lm(Support ~ Income_Centered, data=subset(local, side == 0), weights=weight)
right <- lm(Support ~ Income_Centered, data=subset(local, side == 1), weights=weight)
stopifnot(abs(jump - (coef(right)[1]-coef(left)[1])) < 1e-10,
          sum(local$side == 0) == 537, sum(local$side == 1) == 400)
data.frame(right_minus_left=jump, programme_effect=-jump)
```

This gives a conventional programme-effect estimate of about +0.0335 **score
units**, because treatment is on the left. It is not a 3.35-percentage-point
increase in the probability of supporting government: the outcome includes the
intermediate value 0.5. The coefficient is local to the threshold, not an average
effect for all programme participants.

Ordinary weighted-regression intervals do not account for the approximation bias
of a local regression at a boundary. Use appropriate RD inference for the reported
effect, and keep that inferential choice distinct from this algebraic check.

## 5. Use bias-aware inference and report bandwidths

```{r}
fit_rd <- function(y, x, h=NULL, treatment=NULL) {
  args <- list(y=y, x=x, c=0, p=1, q=2, kernel="triangular",
               vce="hc0", masspoints="adjust", stdvars=TRUE,
               bwselect="mserd", bwrestrict=TRUE, scaleregul=1, level=95)
  if (!is.null(h)) { args$h <- h; args$b <- h }
  if (!is.null(treatment)) args$fuzzy <- treatment
  do.call(rdrobust, args)
}
fixed <- fit_rd(gt$Support, gt$Income_Centered, h=0.01)
automatic <- fit_rd(gt$Support, gt$Income_Centered)
programme_row <- function(fit, label) {
  data.frame(model=label, conventional=-fit$coef[1,1],
             bias_corrected=-fit$coef[3,1],
             lower=-fit$ci[3,2], upper=-fit$ci[3,1],
             h_left=fit$bws[1,1], h_right=fit$bws[1,2],
             b_left=fit$bws[2,1], b_right=fit$bws[2,2],
             n_left=fit$N_h[1], n_right=fit$N_h[2])
}
stopifnot(abs(unname(fixed$coef[1,1]) - jump) < 1e-8)
knitr::kable(rbind(programme_row(fixed,"Fixed window"),
                   programme_row(automatic,"MSE-selected window")), digits=4)
```

`p=1` fits a local line and `q=2` supplies the higher-order bias calculation.
The estimation bandwidth $h$ and bias bandwidth $b$ serve different roles. We
first fix both at 0.01 to make the demonstration reproducible, then let the
MSE selector choose them. Automatic selection is not a certification that the
assignment design is valid.

The fixed-window bias-corrected programme effect is approximately -0.0416, with
a robust interval from -0.1880 to 0.1048. Its sign differs from the conventional
point estimate. Show that difference rather than attaching a bias-corrected interval
to an unlabeled conventional estimate. When reversing treatment direction, negate
the estimate **and swap the negated interval endpoints**.

The automatic conventional estimate also changes sign in this example. That is
substantive evidence of sensitivity, not a reason to hide the selected bandwidth.
Report the result as local and imprecise under these specifications, rather than
choosing the row most consistent with the desired conclusion.

## 6. Examine threats to continuity

Use the separate score dataset for the density check. The outcome survey's sample
selection could distort its score distribution, so it is not an interchangeable
substitute for the density sample.

```{r}
scores <- qed_data("gov_transfers_density")
xd <- scores$Income_Centered
xd <- xd[abs(xd) < 0.02]
stopifnot(nrow(scores) == 52549, length(xd) == 20463)
density <- rddensity(xd, c=0, p=2, q=3, fitselect="unrestricted",
                     kernel="triangular", vce="jackknife", massPoints=TRUE,
                     regularize=TRUE, bwselect="comb", bino=FALSE)
summary(density)
```

A discontinuity in density can be a warning about sorting or recording rules.
An insignificant result is not proof of no manipulation. Ask whether units knew
the threshold, whether the score could be changed precisely, and whether the
recorded score is the assignment score rather than a later revision.

Inspect substantively relevant covariates measured before treatment. The timing
of `Age` and `Education` in this teaching file is not established as pre-treatment,
so call their analyses observed-covariate checks, not confirmed baseline balance
tests. Education has 51 missing values; use variable-specific complete cases
and report their counts rather than dropping those records from every analysis.

## 7. Plan sensitivity checks that test a concern

```{r}
bandwidths <- c(0.005, 0.01, 0.015, 0.02)
sensitivity <- do.call(rbind, lapply(bandwidths, function(h) {
  programme_row(fit_rd(gt$Support, gt$Income_Centered, h=h), paste("h =", h))
}))
knitr::kable(sensitivity, digits=4)
```

A bandwidth grid exposes the bias/variance tradeoff and the role of more distant
units. A local quadratic comparison asks whether curvature matters. Placebo
cutoffs should lie within one treatment regime; their estimation and bias windows
must not cross the real threshold. For example, placebos at +/-0.01 with
$h=b=0.004$ remain within their own regimes.

A donut exclusion removes observations closest to the cutoff. That can address a
specific heaping concern, but it also removes the observations most directly
informing the limiting comparison and requires extrapolation back to zero. It is
not a generic cure for manipulation. Prespecify the reason and report the changed
sample. The [diagnostics lab](../../labs/regression-discontinuity-diagnostics-lab.qmd)
implements these comparisons and rejects insufficient support.

## 8. Extend to a fuzzy design

The second example uses birth-quarter variation in veteran status and homeownership
[@fetter2013mortgages]. Crossing zero changes the probability of veteran status;
it does not deterministically assign it. Under relevance, continuity, exclusion,
and an appropriately oriented monotonicity assumption, a local Wald ratio can
identify an effect for compliers near the cutoff:

$$
\tau_{FRD}=\frac{\lim_{x\downarrow0}E[Y\mid X=x]-\lim_{x\uparrow0}E[Y\mid X=x]}
{\lim_{x\downarrow0}E[D\mid X=x]-\lim_{x\uparrow0}E[D\mid X=x]}.
$$

Use identical observations, weights and score controls for the numerator and
denominator. Veteran status falls from left to right in the supplied score
orientation; do not reverse only one jump. Exclusion concerns every other channel
through which crossing the threshold can affect homeownership. Even a valid
effect of veteran status is not automatically an isolated mortgage-subsidy effect.

```{r}
raw <- as.data.frame(qed_data("mortgages"))
columns <- c("qob_minus_kw", "vet_wwko", "home_ownership")
vet <- raw[complete.cases(raw[columns]) & !is.na(raw$qob_minus_kw) &
             abs(raw$qob_minus_kw) < 12, columns]
stopifnot(nrow(raw) == 214144, nrow(vet) == 56901)
with(vet, c(left_scores=length(unique(qob_minus_kw[qob_minus_kw < 0])),
            right_scores=length(unique(qob_minus_kw[qob_minus_kw >= 0]))))
local_ratio <- function(h) {
  d <- subset(vet, abs(qob_minus_kw) < h)
  d$side <- as.integer(d$qob_minus_kw >= 0)
  d$weight <- 1 - abs(d$qob_minus_kw)/h
  first <- lm(vet_wwko ~ side*qob_minus_kw, data=d, weights=weight)
  reduced <- lm(home_ownership ~ side*qob_minus_kw, data=d, weights=weight)
  first_jump <- unname(coef(first)["side"])
  reduced_jump <- unname(coef(reduced)["side"])
  if (!is.finite(first_jump) || abs(first_jump) < 0.01) stop("Unusable first stage")
  data.frame(h=h, first_stage=first_jump, reduced_form=reduced_jump,
             conventional_ratio=reduced_jump/first_jump)
}
knitr::kable(do.call(rbind, lapply(c(6,9,12), local_ratio)), digits=4)
fuzzy <- fit_rd(vet$home_ownership, vet$qob_minus_kw, h=12,
                treatment=vet$vet_wwko)
summary(fuzzy)
```

At 12 quarters, the first-stage jump is about -0.1213 and the outcome jump about
-0.0226, yielding a conventional ratio near 0.1863, or 18.63 percentage points of
homeownership probability. The fuzzy bias-corrected estimate is around 0.3093
with an interval of approximately [0.1057, 0.5130]. The conventional ratio changes
materially at 6 and 9 quarters. These are not interchangeable estimates.

The first-stage cutoff of 0.01 in the code is a numerical safeguard, **not** a
general weak-instrument test. Report first-stage uncertainty and use inference
appropriate to weak identification when needed. Do not obtain IV standard errors
by treating fitted treatment as an observed regressor in an ordinary second stage.
The [fuzzy lab](../../labs/regression-discontinuity-fuzzy-lab.qmd) checks the ratio
against weighted IV with the same design matrices.

## 9. Respect coarse support and locality

There are only 12 distinct birth-quarter scores on each side of zero in the widest
window, despite tens of thousands of people. More records at the same score do
not create observations arbitrarily close to the cutoff. The extrapolation and
functional-form assumptions therefore deserve explicit discussion. Do not apply
a continuous-score density test to these discrete quarters, or assume that
clustering by quarter automatically fixes approximation bias.

Finally, report the running-variable definition, treatment direction, cutoff,
kernel, polynomial orders, $h$ and $b$, effective sample sizes, distinct score
support, missingness rules, and both conventional and bias-aware results. Include
plots, diagnostic evidence and sensitivity checks without using a checklist of
large p-values as proof of validity. State who the local effect concerns and why
it may not generalize far from the threshold. The [RDD reading map](regression-discontinuity-sources.qmd)
connects these practical choices to the evaluation texts.

## References
