---
title: "How to Do Difference-in-Differences"
subtitle: "A four-cell comparison, followed by a modern staggered-adoption workflow in R"
format: html
execute:
  echo: true
  warning: false
  message: false
---

Difference-in-differences (DiD) asks whether an exposed group's outcome changed
more than an unexposed group's outcome over the same period. Its identifying
claim concerns the **untreated change** for the exposed group, which we cannot
observe after treatment. Regression makes the contrast convenient; it does not
make that missing counterfactual credible.

This guide starts with a two-group, two-period design and then extends to units
treated in different years. It assumes basic regression and R. The executable
examples use [bundled local data](../../labs/data.qmd), not runtime downloads.
Use the [foundations lab](../../labs/difference-in-differences-foundations-lab.qmd)
for a downloadable Quarto document, R notebook, and complete offline bundle.

## 1. Start with assignment and the target effect

Write down who becomes exposed, when exposure starts, what comparison units remain
unexposed, and the outcome scale. Specify whether the target is an effect on the
treated population, on a particular adoption cohort, or at a particular duration
since treatment. These are not interchangeable questions.

Our first example uses Kentucky workers' compensation claims around a 1980 increase
in the benefit cap. High earners were exposed to the change in the cap; low earners
were not exposed in the same way. The outcome is log weeks receiving benefits.
The data are repeated cross-sections of claims, not a panel that follows the same
injured people in both periods [@meyer1995workerscomp; @gertler2016ch7did].

For an ATT interpretation, the key assumptions are parallel untreated trends,
no anticipation, no relevant spillovers, and stable enough composition and
measurement to compare the groups over time. Parallel trends does **not** mean
equal baseline levels. It means the exposed group's average outcome would have
changed like the comparison group's average outcome without the reform.

Different selection into claims after the reform can break this interpretation.
Before estimating anything, consider whether eligibility, reporting incentives,
industry composition, or injury severity changed differently for high earners.
A data dictionary cannot answer all those institutional questions.

## 2. Load and audit the analysis sample

```{r}
source("../../labs/data/load-data.R")
library(ggplot2)
library(fixest)
injury <- qed_data("injury")
ky <- subset(injury, ky == 1)
stopifnot(nrow(ky) == 5626, all(ky$durat > 0),
          all(is.finite(ky$ldurat)), !anyNA(ky[c("highearn", "afchnge")]))
with(ky, table(high_earner=highearn, after=afchnge))
means <- with(ky, tapply(ldurat, list(high=highearn, after=afchnge), mean))
knitr::kable(means, digits=4)
```

Confirm that every group-period cell exists, that treatment coding matches the
policy mechanism, and that sample restrictions were not selected by their effect
on the estimate. Report the number of observations excluded for missing outcomes
or covariates. In an unbalanced panel, attrition requires a separate audit; in
these repeated cross-sections, pay particular attention to who enters each cell.

The snapshot comes from `wooldridge::injury`, but loading it does not require
installing that package. The repository records its source revision and checksum,
so the same analysis uses the same bytes across machines.

## 3. Calculate the four-cell contrast

Let $\bar Y_{g,t}$ be the mean in group $g$ and period $t$, with group one exposed
and time one after the reform. Then

$$
\widehat{ATT}_{DiD}=(\bar Y_{1,1}-\bar Y_{1,0})
                   -(\bar Y_{0,1}-\bar Y_{0,0}).
$$

The counterfactual for exposed claims after the reform is their own baseline
plus the change observed among low earners. That is the design claim to defend,
not a mathematical identity that can be established by fitting a model.

```{r}
comparison_change <- means["0", "1"] - means["0", "0"]
counterfactual <- means["1", "0"] + comparison_change
did_manual <- means["1", "1"] - counterfactual
data.frame(comparison_change, exposed_counterfactual=counterfactual,
           did_log_points=did_manual, geometric_percent=100*expm1(did_manual))
```

```{r}
plot_data <- aggregate(ldurat ~ afchnge + highearn, data=ky, FUN=mean)
plot_data$group <- ifelse(plot_data$highearn == 1, "High earners", "Low earners")
cf <- data.frame(afchnge=c(0,1), ldurat=c(means["1","0"], counterfactual))
ggplot(plot_data, aes(afchnge, ldurat, color=group)) +
  geom_line(linewidth=0.8) + geom_point(size=2) +
  geom_line(data=cf, aes(afchnge, ldurat), inherit.aes=FALSE,
            linetype=2, color="#177b72", linewidth=0.8) +
  scale_x_continuous(breaks=c(0,1), labels=c("Before", "After")) +
  scale_color_manual(values=c("High earners"="#245ca4", "Low earners"="#545454")) +
  labs(x=NULL, y="Mean log benefit duration", color=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
```

The dashed line represents the exposed group's assumed untreated change. With
only one observed pre-period in this design, it is not an estimated pre-treatment
trend. There is no pre-trend test hiding in this four-cell graph.

## 4. Recover the contrast with regression

```{r}
fit <- feols(ldurat ~ highearn * afchnge, data=ky, vcov="hetero")
coeftable(fit)
interaction <- unname(coef(fit)["highearn:afchnge"])
stopifnot(abs(interaction - did_manual) < 1e-10,
          abs(did_manual - 0.1906012007) < 1e-8)
confint(fit, "highearn:afchnge")
```

The interaction is about 0.191 log points. Exponentiating gives a contrast of
roughly 21% on the geometric scale. This is not automatically a 21% effect on
arithmetic mean weeks, nor a 0.191-week change. Outcome transformations affect
both interpretation and the scale on which parallel trends is assumed. Parallel
trends in levels does not generally imply parallel trends in logs.

The heteroskedasticity-robust covariance used here describes sampling uncertainty
under an observation-level working model. It does not create many independent
policy assignments. Clustering on the two earnings groups is not a reliable
repair: two clusters do not support conventional large-cluster inference.
In a real multi-jurisdiction evaluation, start with the level of assignment and
plausible within-group dependence, not whichever standard errors are smallest
[@bertrand2004trustdid].

## 5. Treat covariates as a design choice

Covariates can make conditional parallel trends more plausible or improve
precision. They cannot repair arbitrary unobserved differential trends. In
repeated cross-sections, they may also reveal important composition changes.
Avoid controlling for variables affected by treatment when targeting the total
policy effect. Hospitalization, for example, should not automatically enter a
regression simply because it predicts duration.

Before fitting an adjusted model, draw a causal timeline for each proposed
covariate, inspect its distribution by group and period, and state the conditional
trend assumption. Use a common complete-case sample to distinguish a changed
estimand or changed sample from an estimator difference. Standard regression
adjustment and doubly robust DiD are not the same procedure; the latter combines
outcome and propensity-score components under specific conditions
[@santannazhao2020drdid]. The [foundations lab](../../labs/difference-in-differences-foundations-lab.qmd)
offers an adjusted-model comparison without treating adjustment as an automatic
upgrade.

## 6. Recognize when treatment timing changes the problem

When units adopt in different years, a conventional two-way fixed-effects (TWFE)
coefficient can mix comparisons with different interpretations. Already-treated
units may become controls for newly treated units. With heterogeneous or dynamic
effects, the resulting average need not equal the policy effect you intended to
estimate [@goodmanbacon2021timing; @sunabraham2021dynamic].

Map adoption cohorts before choosing an estimator. Verify that treatment is
absorbing if the method assumes it, distinguish never-treated from not-yet-treated
units, and determine which cohorts have a defensible comparison at each date.
An observation with no eligible comparator cannot be rescued by a longer formula.

For a compact extension, use `did::mpdta`: 500 counties observed in five years,
with log teen employment as the outcome and first treatment year as the cohort
variable. Zero means never treated, not treatment in year zero. This is a teaching
panel, not a full policy replication.

```{r}
library(did)
mpdta <- qed_data("mpdta")
stopifnot(nrow(mpdta) == 2500, length(unique(mpdta$countyreal)) == 500,
          !anyDuplicated(mpdta[c("countyreal", "year")]))
with(mpdta, table(cohort=first.treat, year=year))
```

## 7. Estimate group-time effects and name the comparison

```{r}
set.seed(20260809)
att <- att_gt(yname="lemp", tname="year", idname="countyreal",
              gname="first.treat", data=mpdta, xformla=~1,
              control_group="nevertreated", anticipation=0,
              base_period="universal", est_method="dr",
              bstrap=TRUE, biters=499, cband=TRUE, clustervars="countyreal")
summary(att)
```

This estimates $ATT(g,t)$ for cohort $g$ at calendar time $t$, using never-treated
counties as the comparison group. `xformla=~1` deliberately makes the example
unconditional; a doubly robust method name does not by itself justify conditional
parallel trends. The universal base period normalizes relative to the last
pre-treatment period, given zero anticipation, rather than silently changing the
baseline for every pre-period contrast [@callaway2021multipletime].

The fixed seed and 499 multiplier-bootstrap draws support a reproducible classroom
example. More demanding analysis should investigate simulation stability and
use enough draws for its inferential purpose. County clustering in this teaching
sample is not a defense of county-level inference for a state-level policy;
appropriate assignment identifiers and a defensible dependence model are needed
for a substantive evaluation.

## 8. Aggregate effects without hiding their support

```{r}
overall <- aggte(att, type="simple")
dynamic <- aggte(att, type="dynamic")
data.frame(att_log_points=overall$overall.att, standard_error=overall$overall.se)
stopifnot(abs(overall$overall.att - (-0.0399512752)) < 1e-8)
ggdid(dynamic) + labs(x="Years relative to treatment", y="ATT on log employment")
```

The simple aggregation is about -0.040 log points. Its weights reflect the
available post-treatment group-time cells; it is not necessarily an equal-weight
average over cohorts, calendar years, or treatment durations. A dynamic event
study organizes effects by time since adoption, but later horizons may be supported
only by early adopters. Compare the number and identity of contributing cohorts
before interpreting a changing curve as individual effects growing over time.

The plotted simultaneous bands address a family of event-time effects under the
specified estimation procedure. They are not the same as separate 95% intervals
around every coefficient. Normalized reference cells are fixed at zero by
construction, not exceptionally precise estimates of no effect.

## 9. Diagnose assumptions without declaring them proven

Plot raw group outcomes as well as adjusted effect estimates. Audit missingness,
cohort composition, anticipation windows, spillovers, and policy co-adoption.
Compare never-treated and not-yet-treated controls when each comparison has a
substantive defense. Differences across those estimates may reflect changed
support and identifying assumptions, not a contest in which one package wins.

Pre-treatment estimates can detect some violations, but failure to reject a
pre-trend test does not establish parallel trends. Low power can conceal meaningful
violations, and selecting a specification because it passes a pre-test changes
the inferential problem [@roth2022pretest]. A post-period departure consistent with
the policy may still reflect a concurrent shock affecting treated units differently.

The [staggered diagnostics lab](../../labs/difference-in-differences-staggered-diagnostics-lab.qmd)
adds the Bacon decomposition and Sun-Abraham event study. The
[modern estimators lab](../../labs/difference-in-differences-modern-estimators-lab.qmd)
adds conditional comparisons and an `HonestDiD` extension that asks how much
departure from parallel trends the conclusion can tolerate. Sensitivity bounds
depend on the restrictions supplied; they do not remove the need to defend those
restrictions [@rambachanroth2023parallel].

## 10. Report the design, not just the interaction

State the policy contrast, outcome scale, target population, assignment level,
sample exclusions, comparison group, and anticipation window. For staggered
adoption, report the group-time estimator, aggregation weights, event-time support,
and confidence-band method. Show the effect size and interval, then discuss the
strongest plausible alternative explanation.

For the Kentucky example, explicitly acknowledge that the four-cell sample
contains no multi-period pre-trend evidence. For the minimum-wage example,
acknowledge the teaching panel's inference limitations. Neither a small p-value
nor agreement between two estimators is a substitute for a credible untreated
counterfactual. Use the [DiD reading map](difference-in-differences-sources.qmd)
to match the next methodological step to the actual design problem.

## References
