---
title: "Mixtape Notes: NSW Benchmark From Section 5.3.4 Onward"
subtitle: "Observational versus experimental evidence, with R code"
author: "Luke Heley"
date: 9 March 2026
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
execute:
  echo: true
  eval: false
---

## Purpose

These notes condense section `5.3.4` onward of Cunningham's *Matching and Subclassification* chapter into a teaching sequence that is easy to run in R.

The goal is narrow:

- show why the NSW experimental estimate is the benchmark
- show why the observational NSW-treated-versus-CPS-controls comparison is biased
- show how propensity-score weighting, nearest-neighbor matching, and coarsened exact matching perform against that benchmark

This page is intentionally shorter and more didactic than the broader [Matching and Weighting](matching-and-weighting.qmd) overview and is meant to give the NSW benchmark its own runnable teaching note.

## What Section 5.3.4 Onward Is Doing

From `5.3.4` onward, the chapter stops talking about matching in the abstract and asks a harder question:

Can an observational design recover an effect that we already know from an experiment?

That is why the NSW example matters so much [@lalonde1986evaluating; @dehejia1999causal; @cunningham2021subclassification]:

- the NSW sample contains a randomized treatment-control comparison
- the CPS sample gives a non-experimental control pool
- the analyst can therefore compare observational estimates against an experimental reference rather than against intuition alone

This is the key teaching lesson:

- experimental data solve the comparison problem by design
- observational data do not
- matching and weighting are attempts to repair that problem using observed covariates only
- the repair should be judged by balance, overlap, retention, and closeness to the experimental benchmark

## Core Design Logic

The NSW benchmark exercise should be read as a sequence of four comparisons:

1. Experimental benchmark: NSW treated versus NSW controls.
2. Raw observational estimate: NSW treated versus CPS controls with no adjustment.
3. Adjusted observational estimates: weighting or matching on observed covariates.
4. Benchmark gap: how far each observational estimate remains from the randomized NSW result.

If the raw observational estimate is far away from the benchmark, that is evidence that the treated and untreated groups were not initially comparable.

If an adjusted observational estimate moves closer, that suggests the design repair helped.

If it still misses badly, that is a reminder that conditioning on observables is not the same thing as randomization.

## Step 1: Load Packages And Build The Two Samples

```{r}
required_packages <- c(
  "causaldata",
  "MatchIt",
  "WeightIt",
  "cobalt",
  "dplyr",
  "ggplot2"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

nsw <- causaldata::nsw_mixtape |>
  mutate(
    treat = as.integer(treat),
    outcome = re78
  )

cps <- causaldata::cps_mixtape |>
  mutate(
    treat = as.integer(treat),
    outcome = re78
  )

design_covariates <- c(
  "age", "educ", "black", "hisp", "marr", "nodegree", "re74", "re75"
)

benchmark_dat <- nsw

obs_dat <- bind_rows(
  nsw |> filter(treat == 1L) |> mutate(source = "NSW treated"),
  cps |> filter(treat == 0L) |> mutate(source = "CPS controls")
)

obs_dat |>
  count(source, treat)
```

Interpretation:

- `benchmark_dat` is the experimental comparison
- `obs_dat` is the observational comparison that creates the identification problem

## Step 2: Estimate The Experimental Benchmark

This is the reference effect. It uses only the randomized NSW sample.

```{r}
benchmark_summary <- benchmark_dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    mean_re78 = mean(outcome, na.rm = TRUE),
    mean_re74 = mean(re74, na.rm = TRUE),
    mean_re75 = mean(re75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "NSW treated", "NSW controls")) |>
  select(group, n, mean_re78, mean_re74, mean_re75)

benchmark_fit <- lm(outcome ~ treat, data = benchmark_dat)
benchmark_est <- unname(coef(benchmark_fit)["treat"])

benchmark_summary
summary(benchmark_fit)
benchmark_est
```

What this means:

- the `treat` coefficient is the benchmark `ATT`
- every later observational estimate should be compared back to this value

## Step 3: Show Why Raw Observational Data Are Problematic

Now switch to the harder comparison: NSW treated units against CPS controls.

```{r}
raw_summary <- obs_dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    mean_re78 = mean(outcome, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE),
    mean_educ = mean(educ, na.rm = TRUE),
    prop_black = mean(black == 1, na.rm = TRUE),
    prop_married = mean(marr == 1, na.rm = TRUE),
    mean_re74 = mean(re74, na.rm = TRUE),
    mean_re75 = mean(re75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "NSW treated", "CPS controls")) |>
  select(group, n, mean_re78, mean_age, mean_educ, prop_black, prop_married, mean_re74, mean_re75)

raw_fit <- lm(outcome ~ treat, data = obs_dat)
raw_est <- unname(coef(raw_fit)["treat"])

raw_summary
summary(raw_fit)
raw_est
```

Why this estimate is not credible on its own:

- CPS controls were not randomized against NSW treated units
- the groups differ on pre-treatment covariates
- they also differ on prior earnings, which are strong proxies for labor-market position before treatment
- the raw treated-control gap therefore mixes treatment effects with selection effects

## Step 4: Diagnose Imbalance And Overlap Before Any Adjustment

The most important design check is whether treated and control units look even remotely comparable.

```{r}
m_raw <- matchit(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  method = NULL,
  estimand = "ATT"
)

summary(m_raw, un = TRUE)

cobalt::bal.tab(
  m_raw,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

cobalt::love.plot(
  m_raw,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)
```

Quick overlap check:

```{r}
ps_fit <- glm(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  family = binomial()
)

obs_dat <- obs_dat |>
  mutate(ps = predict(ps_fit, type = "response"))

ggplot(obs_dat, aes(x = ps, fill = factor(treat))) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Estimated propensity score",
    y = "Density",
    fill = "Treat",
    title = "Propensity-score overlap: NSW treated versus CPS controls"
  )
```

What to look for:

- large standardized mean differences before adjustment
- weak common support in the propensity-score distribution
- evidence that some treated units have no close untreated analogs in CPS

This is the basic observational-data problem. The issue is not just omitted variables in theory. It is visible non-comparability in the realized sample.

## Step 5: Weighting On The Propensity Score

Section `5.3.5` moves from matching to weighting. The idea is to reweight controls so they better resemble the treated group on observed covariates [@greifer2025weightit].

```{r}
w_att <- weightit(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  method = "ps",
  estimand = "ATT"
)

summary(w_att)

cobalt::bal.tab(
  w_att,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

weighted_fit <- lm(outcome ~ treat, data = obs_dat, weights = w_att$weights)
weighted_est <- unname(coef(weighted_fit)["treat"])

summary(weighted_fit)
weighted_est
```

Teaching point:

- weighting often keeps more units than strict matching
- but poor overlap shows up as extreme weights
- if a few control units receive huge weight, the design may still be fragile

## Step 6: Nearest-Neighbor Matching

Section `5.3.6` turns to nearest-neighbor matching in its practical form. This is where operational choices such as replacement start to matter [@abadie2011biascorrected; @ho2011matchit].

```{r}
m_nearest <- matchit(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  method = "nearest",
  distance = "glm",
  estimand = "ATT",
  replace = TRUE
)

summary(m_nearest, un = TRUE)

cobalt::bal.tab(
  m_nearest,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

matched_nearest <- match.data(m_nearest)

nearest_fit <- lm(
  outcome ~ treat,
  data = matched_nearest,
  weights = weights
)

nearest_est <- unname(coef(nearest_fit)["treat"])

summary(nearest_fit)
nearest_est
```

What nearest-neighbor matching teaches:

- the label "nearest" does not guarantee a good design
- matching with replacement usually improves match quality for hard-to-match treated cases
- balance diagnostics matter more than the algorithm name

## Step 7: Coarsened Exact Matching

Section `5.3.7` presents coarsened exact matching as a more transparent compromise: exact matching, but on coarsened versions of the covariates [@iacus2012cem].

```{r}
m_cem <- matchit(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  method = "cem",
  estimand = "ATT",
  cutpoints = list(
    age = c(25, 35, 45),
    educ = c(10, 12, 14),
    re74 = c(0, 5000, 10000),
    re75 = c(0, 5000, 10000)
  )
)

summary(m_cem, un = TRUE)

cobalt::bal.tab(
  m_cem,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

matched_cem <- match.data(m_cem)

cem_fit <- lm(
  outcome ~ treat,
  data = matched_cem,
  weights = weights
)

cem_est <- unname(coef(cem_fit)["treat"])

summary(cem_fit)
cem_est
```

What CEM makes especially clear:

- coarsening is the design decision
- tighter bins improve comparability but can discard more treated units
- looser bins retain more data but may weaken balance

That visibility is why CEM is often useful for teaching. The design tradeoff is much easier to see than in more opaque distance-based approaches.

## Step 8: Compare Every Observational Design Back To The Experiment

The benchmark comparison is the whole point of the exercise.

```{r}
results_tbl <- tibble::tibble(
  method = c(
    "Experimental benchmark",
    "Raw observational difference",
    "Propensity-score weighting",
    "Nearest-neighbor matching",
    "Coarsened exact matching"
  ),
  estimate = c(
    benchmark_est,
    raw_est,
    weighted_est,
    nearest_est,
    cem_est
  )
) |>
  mutate(
    benchmark_gap = estimate - benchmark_est,
    abs_benchmark_gap = abs(benchmark_gap)
  )

results_tbl
```

How to interpret the table:

- the experimental row is the target
- the raw observational row shows how misleading untreated observational comparisons can be
- the adjusted rows show whether design repair moved the estimate toward the randomized answer
- no method should be judged only by closeness to the benchmark; you still need to inspect balance, overlap, and sample retention

## The Main Lesson About Observational Versus Experimental Data

The NSW example is useful because it separates two questions that often get blurred together.

Question 1:
Can I compute an adjusted observational estimate?

Yes. Matching and weighting make that easy.

Question 2:
Can I defend that estimate as a credible stand-in for the experimental answer?

Only sometimes, and only after checking:

- whether the treated and control groups overlap on relevant covariates
- whether balance actually improved after adjustment
- whether the adjusted design depends on a thin or unusual subset of controls
- whether the benchmark gap is small enough to inspire confidence

The broader lesson from section `5.3.4` onward is not that one matching method "wins." It is that observational estimators should be stress-tested against better evidence whenever a benchmark exists.

## Short Takeaways For Teaching

- Randomization is the benchmark because it creates comparability by design.
- Observational data start with a comparability problem, not just an estimation problem.
- Matching and weighting are design repairs based on observables, not substitutes for an experiment.
- A better-looking estimate is not enough; you need better balance and plausible overlap.
- The NSW example is powerful because it lets you compare observational methods against a known experimental reference rather than against vague plausibility.

## Next Step In This Repo

If you want the broader design framing around these benchmark comparisons, continue to the [Matching and Weighting](matching-and-weighting.qmd) overview.
