---
title: 'Lab: Seatbelt Law Design and Robustness'
format:
  html:
    toc: true
    embed-resources: true
engine: knitr
execute:
  echo: true
  eval: true
  error: false
references:
- URL: https://api.parliament.uk/historic-hansard/lords/1986/jan/20/motor-vehicles-wearing-of-seat-belts
  accessed:
    date-parts:
    - - 2026
      - 8
      - 4
  author:
  - literal: UK Parliament
  id: ukparliament1986seatbeltdebate
  issued:
    date-parts:
    - - 1986
  publisher: House of Lords Hansard, 20 January 1986, volume 470
  title: Motor vehicles (wearing of seat belts) regulations 1982
  type: ''
---

## How To Use This Page

This is the second interrupted time-series lab. Its guided core takes about 45–60 minutes; the placebo-date exercise is optional.



- Treat alternative outcomes and controls as design choices, not extra regressors to add automatically.
- Compare a short, prespecified set of defensible specifications.
- Use sensitivity results to identify what drives the answer.
- Finish with an attribution judgment, not a preferred p-value.

## Training Goal

By the end of the core lab, you should be able to:

1. distinguish directly targeted outcomes from candidate negative controls;
2. assess pre-intervention comparability before fitting a controlled ITS;
3. explain when exposure and contextual variables belong in a model;
4. compare a compact sensitivity set; and
5. state what the design can and cannot attribute to the law.

## Step 1: Recreate The Seatbelts Analysis Data

The columns do not all measure the same event:

| Variable | Meaning |
|---|---|
| `front` | Front-seat passengers killed or seriously injured |
| `drivers` | Car drivers killed or seriously injured |
| `DriversKilled` | Car drivers killed |
| `rear` | Rear-seat passengers killed or seriously injured |
| `VanKilled` | Van drivers killed |
| `kms` | Distance driven |
| `PetrolPrice` | Petrol price |

```{r}
data_helpers <- c("data/load-data.R", "../data/load-data.R", "docs/labs/data/load-data.R")
data_helpers <- data_helpers[file.exists(data_helpers)]
if (!length(data_helpers)) stop("Extract the complete lab ZIP, including its data folder, before running.")
source(data_helpers[[1]])
```

```{r}
required_packages <- c("ggplot2", "nlme")
missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]
if (length(missing_packages) > 0) {
  stop("Install the documented R environment first; missing: ", paste(missing_packages, collapse=", "), call.=FALSE)
}
invisible(lapply(required_packages, library, character.only = TRUE))

Seatbelts <- qed_data("Seatbelts")

seatbelts <- as.data.frame(Seatbelts)
seatbelts$time <- seq_len(nrow(seatbelts))
seatbelts$date <- seq(as.Date("1969-01-01"), by = "month", length.out = nrow(seatbelts))
seatbelts$post_time <- ifelse(
  seatbelts$law == 1,
  seatbelts$time - min(seatbelts$time[seatbelts$law == 1]),
  0
)
seatbelts$season_sin <- sin(2 * pi * seatbelts$time / 12)
seatbelts$season_cos <- cos(2 * pi * seatbelts$time / 12)
intervention_date <- min(seatbelts$date[seatbelts$law == 1])

stopifnot(
  intervention_date == as.Date("1983-02-01"),
  all(seatbelts[c("front", "drivers", "DriversKilled", "rear", "VanKilled")] > 0),
  all(is.finite(seatbelts$kms)),
  all(is.finite(seatbelts$PetrolPrice))
)
```

The primary outcome remains `front` because compulsory wearing directly covered front-seat occupants. Rear-seat wearing did not become compulsory for adults until 1991, outside this dataset [@ukparliament2012seatbelts].

## Step 2: Compare The Outcome Family

Plotting every outcome on its own scale avoids implying that a driver death and a passenger killed-or-seriously-injured observation are interchangeable.

```{r, fig.height=7.2, fig.width=10}
outcome_names <- c("front", "drivers", "DriversKilled", "rear", "VanKilled")
outcome_labels <- c(
  front = "Front-seat passengers: killed or seriously injured",
  drivers = "Car drivers: killed or seriously injured",
  DriversKilled = "Car drivers: killed",
  rear = "Rear-seat passengers: killed or seriously injured",
  VanKilled = "Van drivers: killed"
)

outcome_long <- do.call(rbind, lapply(outcome_names, function(name) {
  data.frame(
    date = seatbelts$date,
    outcome = unname(outcome_labels[[name]]),
    value = seatbelts[[name]]
  )
}))

ggplot(outcome_long, aes(date, value)) +
  geom_line(linewidth = 0.5, colour = "#24527a") +
  geom_vline(xintercept = intervention_date, linetype = "dashed", colour = "#a23b3b") +
  facet_wrap(~ outcome, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Monthly count") +
  theme_minimal(base_size = 11)
```

Checkpoint:

- Which directly targeted series move in a similar direction?
- Does the rear-seat series look unchanged enough to be a convincing negative control?
- Why is a fall in `VanKilled` a warning against treating it as an automatically unaffected control?

## Step 3: Estimate The Break For Each Outcome

Use the same seasonal segmented model for every outcome so differences do not arise from outcome-specific specification searching.

```{r}
fit_outcome <- function(outcome_name) {
  formula <- as.formula(paste0(
    "log(", outcome_name, ") ~ time + law + post_time + season_sin + season_cos"
  ))
  gls(
    formula,
    data = seatbelts,
    correlation = corAR1(form = ~ time),
    method = "ML"
  )
}

outcome_fits <- setNames(lapply(outcome_names, fit_outcome), outcome_names)

summarise_outcome <- function(outcome_name) {
  model <- outcome_fits[[outcome_name]]
  coefficients <- coef(model)
  immediate <- coefficients[["law"]]
  month_12 <- immediate + 12 * coefficients[["post_time"]]
  data.frame(
    outcome = unname(outcome_labels[[outcome_name]]),
    immediate_percent = 100 * (exp(immediate) - 1),
    month_12_percent = 100 * (exp(month_12) - 1)
  )
}

outcome_results <- do.call(rbind, lapply(outcome_names, summarise_outcome))
outcome_results$immediate_percent <- round(outcome_results$immediate_percent, 1)
outcome_results$month_12_percent <- round(outcome_results$month_12_percent, 1)
outcome_results

stopifnot(
  nrow(outcome_results) == length(outcome_names),
  all(is.finite(outcome_results$immediate_percent)),
  all(is.finite(outcome_results$month_12_percent))
)
```

Agreement among directly targeted outcomes strengthens the descriptive story. Movement in nominally untargeted outcomes complicates attribution because it may indicate shared road-safety changes, changes in who sat where, or an unsuitable comparison.

## Step 4: Audit Candidate Comparisons Before Controlling

A comparison outcome should have a credible causal role and a stable pre-law relationship with the treated outcome. Correlation is only a diagnostic; it does not establish either condition.

```{r}
pre_law <- subset(seatbelts, law == 0)

comparison_audit <- data.frame(
  candidate = c("Rear-seat casualties", "Van-driver deaths"),
  pre_law_log_correlation = c(
    cor(log(pre_law$front), log(pre_law$rear)),
    cor(log(pre_law$front), log(pre_law$VanKilled))
  ),
  causal_concern = c(
    "Passenger redistribution and different exposure",
    "Weak pre-law relationship and its own intervention-period fall"
  )
)

comparison_audit
```

Rear-seat casualties are the more informative candidate, but they remain imperfect. The controlled model below is therefore a diagnostic contrast, not an automatic upgrade over the uncontrolled ITS.

```{r}
controlled_data <- rbind(
  transform(seatbelts, series = "Rear-seat comparison", log_outcome = log(rear)),
  transform(seatbelts, series = "Front-seat targeted", log_outcome = log(front))
)
controlled_data$series <- relevel(factor(controlled_data$series), ref = "Rear-seat comparison")
controlled_data <- controlled_data[order(controlled_data$series, controlled_data$time), ]

controlled_fit <- gls(
  log_outcome ~ series * (time + law + post_time) + season_sin + season_cos,
  data = controlled_data,
  correlation = corAR1(form = ~ time | series),
  method = "ML"
)

controlled_terms <- summary(controlled_fit)$tTable[c(
  "seriesFront-seat targeted:law",
  "seriesFront-seat targeted:post_time"
), ]

controlled_terms

controlled_immediate_percent <- 100 * (
  exp(coef(controlled_fit)[["seriesFront-seat targeted:law"]]) - 1
)
controlled_immediate_percent

stopifnot(
  all(is.finite(controlled_terms)),
  comparison_audit$pre_law_log_correlation[[1]] >
    comparison_audit$pre_law_log_correlation[[2]]
)
```

The interaction asks whether the front-seat series changed differently from the rear-seat series. It does not prove that rear-seat casualties represent the untreated path front-seat casualties would have followed.

## Step 5: Run A Compact Prespecified Sensitivity Set

Distance driven is a plausible exposure measure and petrol price may affect travel. Both are observed contextual variables, but neither should be inserted mechanically: distance could itself respond to policy or other contemporaneous conditions, and petrol price is not the intervention.

```{r}
model_formulas <- list(
  "Level only" = log(front) ~ time + law + season_sin + season_cos,
  "Level and trend" = log(front) ~ time + law + post_time + season_sin + season_cos,
  "With distance and petrol price" = log(front) ~
    time + law + post_time + log(kms) + PetrolPrice + season_sin + season_cos
)

sensitivity_fits <- lapply(model_formulas, function(formula) {
  gls(
    formula,
    data = seatbelts,
    correlation = corAR1(form = ~ time),
    method = "ML"
  )
})

sensitivity_results <- do.call(rbind, lapply(names(sensitivity_fits), function(name) {
  model <- sensitivity_fits[[name]]
  data.frame(
    specification = name,
    immediate_percent = 100 * (exp(coef(model)[["law"]]) - 1),
    AIC = AIC(model)
  )
}))

row.names(sensitivity_results) <- NULL
sensitivity_results$immediate_percent <- round(sensitivity_results$immediate_percent, 1)
sensitivity_results$AIC <- round(sensitivity_results$AIC, 1)
sensitivity_results

stopifnot(
  nrow(sensitivity_results) == 3L,
  all(is.finite(sensitivity_results$immediate_percent)),
  all(is.finite(sensitivity_results$AIC))
)
```

Do not select the row with the largest change or smallest p-value. Explain what assumption changes between rows and whether the substantive direction survives.

## Optional Extension: Test Prespecified False Dates

False dates ask whether similarly large breaks appear when no seatbelt-law change occurred. Restrict false-date models to the pre-law period so the real intervention cannot contaminate them.

```{r}
fit_interruption_date <- function(data, interruption_date) {
  false_data <- data
  false_time <- min(false_data$time[false_data$date >= interruption_date])
  false_data$interrupted <- as.integer(false_data$time >= false_time)
  false_data$time_after_interruption <- pmax(0, false_data$time - false_time)

  model <- gls(
    log(front) ~
      time + interrupted + time_after_interruption + season_sin + season_cos,
    data = false_data,
    correlation = corAR1(form = ~ time),
    method = "ML"
  )

  data.frame(
    date = interruption_date,
    immediate_percent = 100 * (exp(coef(model)[["interrupted"]]) - 1)
  )
}

false_dates <- as.Date(c("1978-02-01", "1980-02-01", "1981-02-01"))
placebo_results <- do.call(
  rbind,
  lapply(false_dates, function(date) fit_interruption_date(pre_law, date))
)

actual_result <- data.frame(
  date = intervention_date,
  immediate_percent = sensitivity_results$immediate_percent[
    sensitivity_results$specification == "Level and trend"
  ]
)

rbind(placebo_results, actual_result)

stopifnot(
  nrow(placebo_results) == length(false_dates),
  all(placebo_results$date < intervention_date),
  all(is.finite(placebo_results$immediate_percent))
)
```

Placebo dates are diagnostics, not randomization-inference p-values. A quiet placebo set cannot rule out a concurrent event at the true date.

## Final Design Memo

Complete five short statements:

1. **Primary outcome:** why front-seat killed-or-seriously-injured casualties match the intervention.
2. **Comparison outcome:** what rear-seat casualties add and why they are imperfect.
3. **Sensitivity:** whether the direction changes across the three prespecified specifications.
4. **Concurrent change:** stricter drink-driving enforcement and evidential breath testing were discussed as alternative 1983 explanations and are not measured in `Seatbelts` [@ukparliament1986seatbeltdebate].
5. **Defensible claim:** the series show a reduction associated with the law, but this dataset alone does not isolate compulsory belt wearing from every simultaneous road-safety change.

## Next Step

Continue to [Interrupted Time Series Counterfactual Validation](https://defenceeconomist.github.io/qedlabs/labs/interrupted-time-series-counterfactual-validation-lab.html) to choose an untreated forecasting model without using post-intervention outcomes.
