---
title: "Lab: Black Politicians"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as a guided worksheet rather than as a finished report.

- Read the short framing note for each step.
- Run the starter code in your own R session.
- Write down what you learn before moving to the next section.
- Fill in the comparison table near the end so you can compare designs side by side.

::: {.callout-note}
The code is shown but not executed when the website is rendered. That keeps the page readable even if `causaldata` is not installed yet.
:::

## Training Goal

Practice the design-first workflow from the matching report on a second dataset:

1. define the estimand before choosing a method
2. map treatment, outcome, and covariates clearly
3. diagnose raw imbalance
4. compare exact matching, CEM, and entropy balancing
5. interpret the tradeoff between balance, retention, and weight concentration

## Dataset At A Glance

`causaldata::black_politicians` comes from Broockman (2013), a U.S. field experiment on legislator responsiveness that is also used in the *Matching* chapter of *The Effect*.

- Observations: `5,593`
- Variables: `14`
- Unit of analysis: one email-to-legislator observation
- Built-in randomized factor: `treat_out` (whether the email was out-of-district)

The important design feature is that this dataset contains both:

- an experimental component: `treat_out`, which was randomized in the original study
- an observational comparison: legislator characteristics such as `leg_black`, `leg_democrat`, and district composition

That makes it useful for training because you can separate two different questions:

1. what the original experiment randomized
2. what you are doing when you instead compare observed groups of legislators

In this lab, we deliberately focus on the second question. We are **not** estimating the effect of the randomized out-of-district email treatment. Instead, we treat `leg_black` as the focal training comparison and ask whether Black and non-Black legislators can be made more comparable on a compact set of pre-outcome covariates.

For this training version, we use:

- focal training comparison: `leg_black`
- binary outcome: `responded`
- core design covariates: `medianhhincom`, `blackpercent`, `leg_democrat`
- extension covariates for later rounds: `leg_senator`, `south`, `urbanpercent`, `totalpop`

These variables work well for a matching lab because they mix:

- legislator attributes such as party and chamber
- district composition measures such as racial composition, income, population, and urbanicity
- an observable behavioral outcome, `responded`

That structure supports a clear design-first sequence: inspect raw imbalance, test whether literal exact matching is feasible, then compare CEM and entropy balancing under a shared `ATT` target.

::: {.callout-warning}
This is best treated as a training example for matching logic, not as a literal policy treatment that could be assigned. `leg_black` is an observed attribute, so the main teaching value here is design transparency: what comparability would have to mean, where overlap is thin, and how the matched estimand changes as you tighten the design.
:::

## Core Design Choice For This Training Aid

For a first pass, keep the exercise simple and explicit.

| Role | Variable | Why it is used here |
|---|---|---|
| Treatment | `leg_black` | Focal comparison for the matching exercise |
| Outcome | `responded` | Whether the legislator replied |
| Estimand | `ATT` | Average difference for the Black-legislator group |
| Core covariates | `medianhhincom`, `blackpercent`, `leg_democrat` | Compact set that captures district resources, district racial composition, and party |
| Extension covariates | `leg_senator`, `south`, `urbanpercent`, `totalpop` | Useful once the basic workflow is clear |
| Post-design modeling variable | `treat_out` | Part of the original study design; add it after you understand the basic matched comparison |

Why not start with every variable at once? Because this page is meant to train judgement, not to hide the design behind a long formula.

## Suggested Session Flow

If you are teaching this live, a simple `35-45` minute structure works well:

1. `5 minutes`: frame the question and walk through the variable map
2. `10 minutes`: run raw descriptives and balance diagnostics
3. `5 minutes`: try exact matching and discuss why it struggles
4. `10 minutes`: run CEM and inspect the retention-versus-balance tradeoff
5. `10 minutes`: run entropy balancing and inspect effective sample size
6. `5 minutes`: compare designs and decide which one you would defend

## Step 1: Load The Data

Start by loading the packages, bringing in the data, and creating a compact analysis object.

```{r}
required_packages <- c(
  "causaldata",
  "MatchIt",
  "WeightIt",
  "cobalt",
  "dplyr",
  "ggplot2"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

dat <- causaldata::black_politicians |>
  mutate(
    treat = as.integer(leg_black),
    outcome = as.integer(responded)
  )

design_covariates <- c("medianhhincom", "blackpercent", "leg_democrat")

dat |>
  select(treat, outcome, treat_out, all_of(design_covariates)) |>
  glimpse()
```

Checkpoint:

- Is the treatment binary?
- Is the outcome binary?
- Are the matching covariates measured before the outcome?

## Step 2: Describe The Raw Comparison

Do not start with a matching algorithm. First check whether the untreated group looks remotely comparable to the treated group on the covariates you care about.

```{r}
raw_summary <- dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    response_rate = mean(outcome, na.rm = TRUE),
    mean_income = mean(medianhhincom, na.rm = TRUE),
    mean_blackpercent = mean(blackpercent, na.rm = TRUE),
    prop_democrat = mean(leg_democrat == 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "Black legislators", "Non-Black legislators")) |>
  select(group, n, response_rate, mean_income, mean_blackpercent, prop_democrat)

raw_summary
```

What to look for:

- large differences in `medianhhincom`
- large differences in `blackpercent`
- major party imbalance on `leg_democrat`

If the raw groups are far apart, that does not mean matching will fail, but it does mean the burden on the design is real.

## Step 3: Baseline Balance Diagnostics

Use `MatchIt` with `method = NULL` to get a formal pre-adjustment balance benchmark.

```{r}
m_raw <- matchit(
  treat ~ medianhhincom + blackpercent + leg_democrat,
  data = dat,
  method = NULL,
  estimand = "ATT"
)

summary(m_raw, un = TRUE)

cobalt::bal.tab(m_raw, un = TRUE, m.threshold = 0.1)

cobalt::love.plot(
  m_raw,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)
```

Checkpoint:

- Which covariate is least balanced?
- Are any absolute standardized mean differences already below `0.1`?
- If balance is poor, is the problem modest or structural?

## Step 4: Exact Matching Feasibility Check

Exact matching is easiest to explain, but it becomes restrictive quickly. Use a reduced discrete specification to show both its appeal and its limits.

```{r}
m_exact <- matchit(
  treat ~ leg_democrat + leg_senator + south,
  data = dat,
  method = "exact",
  estimand = "ATT"
)

summary(m_exact, un = TRUE)

exact_dat <- match.data(m_exact)

exact_dat |>
  count(treat) |>
  mutate(group = if_else(treat == 1L, "Treated", "Control"))
```

What to discuss:

- Exact matching is transparent because every retained comparison is literal on the chosen variables.
- It will usually ignore important continuous differences unless you bin them first.
- If many treated units are dropped, the estimand begins to drift toward "the treated units for whom exact matches exist."

## Step 5: Build A CEM Version

CEM relaxes exact matching by coarsening continuous covariates into bins. That usually improves feasibility while keeping the matching logic visible.

```{r}
create_even_breaks <- function(x, n) {
  min_x <- min(x, na.rm = TRUE)
  max_x <- max(x, na.rm = TRUE)
  min_x + ((0:n) / n) * (max_x - min_x)
}

cem_cutpoints <- list(
  medianhhincom = quantile(dat$medianhhincom, probs = (0:6) / 6, na.rm = TRUE),
  blackpercent = create_even_breaks(dat$blackpercent, 6)
)

m_cem <- matchit(
  treat ~ medianhhincom + blackpercent + leg_democrat + leg_senator + south,
  data = dat,
  method = "cem",
  estimand = "ATT",
  cutpoints = cem_cutpoints
)

summary(m_cem, un = TRUE)

cobalt::bal.tab(m_cem, un = TRUE, m.threshold = 0.1)

cobalt::love.plot(
  m_cem,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)
```

Questions to answer after running it:

1. How much better is balance than in the raw sample?
2. How many treated units are retained?
3. Which binning choices matter most for the result?

## Step 6: Try Entropy Balancing

Entropy balancing keeps the treated group intact and reweights the controls. That often gives excellent balance, but the price can be concentrated weights and a smaller effective control sample.

```{r}
w_ebal <- weightit(
  treat ~ medianhhincom + I(medianhhincom^2) +
    blackpercent + I(blackpercent^2) +
    leg_democrat + leg_senator + south,
  data = dat,
  method = "ebal",
  estimand = "ATT"
)

summary(w_ebal)

cobalt::bal.tab(w_ebal, un = TRUE, m.threshold = 0.1)

cobalt::love.plot(
  w_ebal,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)

ess <- function(w) {
  (sum(w)^2) / sum(w^2)
}

dat_ebal <- dat |>
  mutate(w = w_ebal$weights)

dat_ebal |>
  group_by(treat) |>
  summarise(
    raw_n = n(),
    ess = ess(w),
    max_weight = max(w),
    mean_weight = mean(w),
    .groups = "drop"
  )
```

Interpretation prompt:

- If balance looks excellent but control-side ESS collapses, would you still prefer this design?

## Step 7: Estimate And Compare

Keep the outcome model simple for the training version. The point is to compare designs, not to hide differences behind a complicated specification.

```{r}
raw_fit <- lm(outcome ~ treat, data = dat)

exact_fit <- lm(outcome ~ treat, data = exact_dat, weights = weights)

cem_dat <- match.data(m_cem)
cem_fit <- lm(outcome ~ treat, data = cem_dat, weights = weights)

ebal_fit <- lm(outcome ~ treat, data = dat_ebal, weights = w)

coef(summary(raw_fit))
coef(summary(exact_fit))
coef(summary(cem_fit))
coef(summary(ebal_fit))
```

You can stop here for an introductory session. If you want to connect the lab back to the original study design, add `treat_out` and relevant controls after the design stage rather than mixing everything together at the start.

## Comparison Table To Fill In

Use this table as the main training record.

| Design | Worst abs. SMD | Treated retained | Control ESS | Estimated effect | Your judgement |
|---|---|---|---|---|---|
| Raw sample |  |  |  |  |  |
| Exact matching |  |  |  |  |  |
| CEM |  |  |  |  |  |
| Entropy balancing |  |  |  |  |  |

## Debrief Questions

1. Which design gave the best balance on the covariates you actually care about?
2. Which design preserved the clearest connection to the original `ATT`?
3. Did better balance come from pruning observations, reweighting heavily, or both?
4. If you had to explain your preferred design to a non-technical policy audience, which one would be easiest to defend?

## Optional Extension

Once the core exercise is complete, extend the outcome model to reflect the original study more closely.
This same interaction model is now included in the evaluated-code artifact so you can compare your local run with the QA checks directly.

```{r}
final_fit <- lm(
  outcome ~ treat * treat_out +
    nonblacknonwhite +
    black_medianhh +
    white_medianhh +
    statessquireindex +
    totalpop +
    urbanpercent,
  data = cem_dat,
  weights = weights
)

summary(final_fit)
```

This extension is useful because it separates two tasks:

- design the comparison first
- fit the richer substantive model second
