---
title: "Lab: Difference-in-Differences Foundations With Kentucky Workers' Compensation"
aliases:
  - /labs/difference-in-differences-foundations-python-lab.html
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

This is the first difference-in-differences lab. The guided core takes about 45–60 minutes and starts with the four means that define the canonical two-group, two-period estimator.

<!-- lab-downloads:start -->
::: {.resource-links}
[Quarto document](downloads/difference-in-differences-foundations-lab.qmd){download="difference-in-differences-foundations-lab.qmd"}
[R Jupyter notebook](notebooks/difference-in-differences-foundations-lab.ipynb){download="difference-in-differences-foundations-lab.ipynb"}
[Complete lab ZIP](downloads/difference-in-differences-foundations-lab.zip){download="difference-in-differences-foundations-lab.zip"}
[Data and setup](data.qmd#difference-in-differences-foundations-lab)
:::
<!-- lab-downloads:end -->

- Declare the treated group, comparison group, policy date, outcome, and estimand before fitting a model.
- Calculate the contrast by hand before using an interaction regression.
- Separate a numerical estimate from the assumptions that give it a causal interpretation.
- Treat covariate adjustment and standard errors as design decisions, not automatic upgrades.

The workflow draws on Gertler et al.'s counterfactual framing, Huntington-Klein's design-first treatment, and Andrew Heiss's worked example [@gertler2016ch7did; @huntingtonklein2021did; @heiss2026didexample]. See the [DiD overview](../notes/did/difference-in-differences.qmd) and [Heiss source notes](../notes/did/andrew-heiss-difference-in-differences-notes.qmd) for the underlying reading.

[Tested environment and reproduction record](difference-in-differences-reproducibility.qmd)

## Training Goal

By the end of the lab, you should be able to:

1. identify the four cells in a canonical DiD design;
2. calculate the DiD contrast manually and recover it with regression;
3. interpret a log-outcome estimate on a percentage scale;
4. state the parallel-trends and composition assumptions precisely; and
5. explain what robust standard errors and covariate adjustment do—and do not—solve.

## Step 1: Declare The Design

Kentucky raised its cap on covered weekly earnings in 1980. The change affected high earners but not low earners, so high earners form the policy-exposed group and low earners form the comparison group. The outcome is the log duration of workers' compensation benefits following an injury [@meyer1995workerscomp].

| Design component | Definition |
|---|---|
| Treated group | High-earning Kentucky claimants |
| Comparison group | Low-earning Kentucky claimants |
| Before/after indicator | Claim occurred before or after the 1980 change |
| Outcome | Log weeks receiving benefits |
| Target contrast | Change for high earners minus the contemporaneous change for low earners |

The records are repeated cross-sections of claims, not a panel following the same injured workers over time. Under parallel trends, no anticipation, stable group composition, and no spillovers, the contrast can be interpreted as the average policy effect for the exposed group.

## Step 2: Load And Audit The Data

The `injury` data are distributed with the [`wooldridge` R package](https://search.r-project.org/CRAN/refmans/wooldridge/html/injury.html). We use a pinned snapshot bundled in this repository, not a runtime download.

```{r}
data_helpers <- c("data/load-data.R", "../data/load-data.R", "docs/labs/data/load-data.R")
data_helpers <- data_helpers[file.exists(data_helpers)]
if (!length(data_helpers)) stop("Extract the complete lab ZIP, including its data folder, before running.")
source(data_helpers[[1]])
```

```{r}
required_packages <- c(
  "digest", "dplyr", "ggplot2", "fixest")
missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]
if (length(missing_packages) > 0) {
  stop("Install the documented R environment first; missing: ", paste(missing_packages, collapse=", "), call.=FALSE)
}
invisible(lapply(required_packages, library, character.only = TRUE))

injury <- qed_data("injury")


injury_ky <- injury |>
  filter(ky == 1) |>
  transmute(
    duration = durat,
    log_duration = ldurat,
    after_change = afchnge,
    high_earner = highearn,
    male,
    married,
    age,
    hospitalized = hosp,
    industry = factor(indust),
    injury_type = factor(injtype),
    log_pre_wage = lprewage
  )

cell_counts <- with(injury_ky, table(after_change, high_earner))

data.frame(
  observations = nrow(injury_ky),
  before_claims = sum(injury_ky$after_change == 0),
  after_claims = sum(injury_ky$after_change == 1),
  low_earner_claims = sum(injury_ky$high_earner == 0),
  high_earner_claims = sum(injury_ky$high_earner == 1)
)

cell_counts

stopifnot(
  nrow(injury_ky) == 5626L,
  all(cell_counts > 0),
  all(is.finite(injury_ky$log_duration)),
  all(injury_ky$duration > 0)
)
```

Checkpoint: what would fail if the policy also materially changed which low earners filed claims after 1980?

## Step 3: Plot The Four Cells

```{r, fig.height=5, fig.width=9}
cell_means <- injury_ky |>
  group_by(after_change, high_earner) |>
  summarise(
    mean_log_duration = mean(log_duration),
    standard_error = sd(log_duration) / sqrt(n()),
    claims = n(),
    .groups = "drop"
  ) |>
  mutate(
    period = factor(
      after_change,
      levels = c(0, 1),
      labels = c("Before 1980", "After 1980")
    ),
    group = factor(
      high_earner,
      levels = c(0, 1),
      labels = c("Low earners", "High earners")
    ),
    lower_95 = mean_log_duration - 1.96 * standard_error,
    upper_95 = mean_log_duration + 1.96 * standard_error
  )

ggplot(
  cell_means,
  aes(period, mean_log_duration, colour = group, group = group)
) +
  geom_line(linewidth = 0.8) +
  geom_pointrange(
    aes(ymin = lower_95, ymax = upper_95),
    linewidth = 0.7
  ) +
  labs(
    x = NULL,
    y = "Mean log benefit duration",
    colour = NULL,
    title = "The canonical DiD plot contains four group-period means"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The graph describes the observed means. With only one before period, it cannot show whether the two groups followed parallel trends before the policy.

## Step 4: Calculate Difference-in-Differences By Hand

Let $\bar{Y}_{g,t}$ denote a group-period mean. The canonical contrast is

$$
\widehat{ATT}_{DiD} =
(\bar{Y}_{T,post} - \bar{Y}_{T,pre}) -
(\bar{Y}_{C,post} - \bar{Y}_{C,pre}).
$$

```{r}
cell_value <- function(after_value, high_value) {
  cell_means |>
    filter(
      after_change == after_value,
      high_earner == high_value
    ) |>
    pull(mean_log_duration)
}

treated_before <- cell_value(0, 1)
treated_after <- cell_value(1, 1)
control_before <- cell_value(0, 0)
control_after <- cell_value(1, 0)

treated_change <- treated_after - treated_before
control_change <- control_after - control_before
did_manual <- treated_change - control_change

did_table <- data.frame(
  group = c("Low earners", "High earners", "Difference"),
  before = c(control_before, treated_before, treated_before - control_before),
  after = c(control_after, treated_after, treated_after - control_after),
  change = c(control_change, treated_change, did_manual)
)

did_table |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

Checkpoint: identify the unobserved counterfactual mean that the low-earner change is being used to construct.

## Step 5: Recover The Same Contrast With Regression

```{r}
basic_fit <- feols(
  log_duration ~ high_earner * after_change,
  data = injury_ky,
  vcov = "hetero"
)

did_regression <- coef(basic_fit)[["high_earner:after_change"]]
did_standard_error <- se(basic_fit)[["high_earner:after_change"]]

basic_result <- data.frame(
  estimate_log_points = did_regression,
  heteroskedasticity_robust_se = did_standard_error,
  lower_95 = did_regression - 1.96 * did_standard_error,
  upper_95 = did_regression + 1.96 * did_standard_error
)

basic_result

stopifnot(
  abs(did_manual - did_regression) < 1e-10,
  is.finite(did_standard_error),
  did_standard_error > 0
)
```

The interaction coefficient is algebraically identical to the manual DiD contrast. Regression has not changed the identifying comparison; it has made inference and extensions easier.

## Step 6: Interpret The Log Outcome

For a contrast in mean log outcomes, $100\widehat{\delta}$ approximates the percentage difference in the ratio of geometric-mean changes. The exact transformation of that ratio is $100[\exp(\widehat{\delta})-1]$.

```{r}
percentage_approximation <- 100 * did_regression
percentage_exact <- 100 * (exp(did_regression) - 1)

data.frame(
  interpretation = c("Approximation", "Exact transformation"),
  percent_change = c(percentage_approximation, percentage_exact)
)

stopifnot(
  percentage_exact > percentage_approximation,
  percentage_exact > 0
)
```

State the 0.191 log-point result and its approximately 21% geometric-mean ratio interpretation. It is not automatically a 21% causal change in arithmetic mean weeks or the average individual percentage effect; those interpretations need additional assumptions.

## Step 7: Compare A Prespecified Adjusted Model

Heiss's example adds claimant, injury, and industry characteristics. This can improve precision and address observed composition differences only if the variables are measured consistently and are not consequences of the policy [@heiss2026didexample]. It cannot repair differential unobserved trends. Because some covariates are missing, compare models on a common complete-case sample before attributing a coefficient change to adjustment.

```{r}
adjustment_variables <- c(
  "male",
  "married",
  "age",
  "hospitalized",
  "industry",
  "injury_type",
  "log_pre_wage"
)

adjusted_sample <- injury_ky |>
  filter(if_all(all_of(adjustment_variables), ~ !is.na(.x)))

basic_common_sample_fit <- feols(
  log_duration ~ high_earner * after_change,
  data = adjusted_sample,
  vcov = "hetero"
)

adjusted_fit <- feols(
  log_duration ~ high_earner * after_change +
    male + married + age + hospitalized +
    industry + injury_type + log_pre_wage,
  data = adjusted_sample,
  vcov = "hetero"
)

model_comparison <- data.frame(
  model = c(
    "Unadjusted, full sample",
    "Unadjusted, common sample",
    "Adjusted, common sample"
  ),
  observations = c(
    nobs(basic_fit),
    nobs(basic_common_sample_fit),
    nobs(adjusted_fit)
  ),
  estimate = c(
    coef(basic_fit)[["high_earner:after_change"]],
    coef(basic_common_sample_fit)[["high_earner:after_change"]],
    coef(adjusted_fit)[["high_earner:after_change"]]
  ),
  robust_se = c(
    se(basic_fit)[["high_earner:after_change"]],
    se(basic_common_sample_fit)[["high_earner:after_change"]],
    se(adjusted_fit)[["high_earner:after_change"]]
  )
)

model_comparison$exact_percent <- 100 * (exp(model_comparison$estimate) - 1)
model_comparison

stopifnot(
  nrow(model_comparison) == 3L,
  nobs(basic_common_sample_fit) == nobs(adjusted_fit),
  nobs(adjusted_fit) < nobs(basic_fit),
  all(is.finite(as.matrix(model_comparison[c("estimate", "robust_se", "exact_percent")])))
)
```

Checkpoint: for each added variable, ask whether it is a pre-policy characteristic, a measure that could respond to the policy, or a proxy for changing claim composition.

## Step 8: Audit Identification And Inference

Complete this table before writing a causal conclusion.

| Issue | What the data show | What remains assumed |
|---|---|---|
| Parallel trends | One pre-policy mean per group | The untreated group gap would otherwise have remained stable |
| No anticipation | Policy timing is coded before/after | Claims and behaviour did not adjust before the recorded change |
| Composition | Claim characteristics can be compared | The policy did not differentially change who appears in each group |
| Spillovers | Low earners were not directly covered by the higher cap | Their claims were not indirectly affected |
| Inference | Thousands of claims permit heteroskedasticity-robust calculations | There are only two policy groups and one policy change |

The robust standard error allows unequal residual variance across claims. It does not create thousands of independent policy assignments, and clustering on only two earnings groups would not provide reliable cluster inference. The design's identifying variation remains group by period.

## Practical Implications

- The four-cell calculation is the design; the regression interaction is its representation.
- A two-period design cannot use pre-policy leads to diagnose parallel trends.
- Covariate adjustment may improve comparability or precision, but only after the covariates' causal timing is defended.
- Individual-level sample size should not be confused with the number of independent policy shocks.
- A defensible conclusion reports the estimate, the comparison group, and the unsupported assumptions together.

Next, use the [staggered-adoption diagnostics lab](difference-in-differences-staggered-diagnostics-lab.qmd) to see why the meaning of TWFE changes when treatment begins at different times.

## Worked answers

- The regression interaction equals the manual DiD because the unadjusted model is saturated in the four group-period cells.
- A flat-looking two-point graph contains no evidence about pre-treatment trends.
- Compare adjusted and unadjusted estimates on a common sample to separate covariate adjustment from missing-data selection.
- Exponentiating the log contrast describes a ratio of geometric-mean changes, not automatically the effect on arithmetic mean weeks.
