---
title: "Lab: Modern Multi-Period Difference-in-Differences"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

This is the third difference-in-differences lab. The guided core takes about 45–60 minutes; the `HonestDiD` sensitivity analysis is an optional extension.

[Download the R Jupyter notebook](notebooks/difference-in-differences-modern-estimators-lab.ipynb){.cta-button download="difference-in-differences-modern-estimators-lab.ipynb"}

- Estimate group-time effects before choosing an aggregation.
- Make the comparison group and parallel-trends assumption explicit.
- Use simultaneous confidence bands for a dynamic effect path.
- Compare estimators by target and support, not only by numerical proximity.

The main workflow implements Callaway and Sant'Anna with the [`did` package](https://bcallaway11.github.io/did/) and compares it with Sun–Abraham in `fixest` [@callaway2021multipletime; @sunabraham2021dynamic]. The extension implements Rambachan and Roth with [`HonestDiD`](https://cran.r-project.org/web/packages/HonestDiD/HonestDiD.pdf) [@rambachanroth2023parallel]. See the [Callaway–Sant'Anna notes](../notes/did/callaway-santanna-multiple-periods-notes.qmd) and [modern DiD synthesis](../notes/did/roth-et-al-did-synthesis-notes.qmd).

[Python companion](difference-in-differences-modern-estimators-python-lab.qmd) · [Tested environment and reproduction record](difference-in-differences-reproducibility.qmd)

## Training Goal

By the end of the core lab, you should be able to:

1. define and estimate $ATT(g,t)$ for staggered adoption;
2. distinguish never-treated from not-yet-treated comparisons;
3. aggregate effects by cohort, event time, and calendar time;
4. distinguish unconditional from conditional parallel trends; and
5. explain why modern estimators can disagree without either containing a coding error.

## Step 1: Load And Audit `mpdta`

[`did::mpdta`](https://bcallaway11.github.io/did/reference/mpdta.html) is a balanced county-year panel based on minimum-wage applications. The outcome is log teen employment (`lemp`), `first.treat` is the first treatment year, and zero identifies never-treated counties.

```{r}
required_packages <- c(
  "digest",
  "did",
  "dplyr",
  "ggplot2",
  "fixest"
)
missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}
invisible(lapply(required_packages, library, character.only = TRUE))

# Match the Python lab's immutable upstream data and verify its checksum.
data_file <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/did/453a1be859411edc1c8933891940cb3a570d7095/data/mpdta.rda", data_file, mode = "wb", quiet = TRUE)
stopifnot(digest::digest(file = data_file, algo = "sha256") == "0be3bc5eaaedaa50e047deaf45ad0e193e7210df1644103c63a6c1baa75965d4")
load(data_file)
unlink(data_file)


mpdta <- mpdta |> mutate(exposed = as.integer(first.treat > 0 & year >= first.treat))

# treat is an ever-treated indicator; audit actual exposure instead.
county_audit <- mpdta |>
  arrange(countyreal, year) |>
  group_by(countyreal) |>
  summarise(
    observations = n(),
    first_year = min(year),
    last_year = max(year),
    cohort_values = n_distinct(first.treat),
    treatment_reversal = any(diff(exposed) < 0),
    .groups = "drop"
  )

cohort_table <- mpdta |>
  distinct(countyreal, first.treat) |>
  count(first.treat, name = "counties") |>
  mutate(
    cohort = if_else(
      first.treat == 0,
      "Never treated",
      as.character(first.treat)
    )
  ) |>
  select(cohort, first.treat, counties) |>
  arrange(first.treat)

data.frame(
  observations = nrow(mpdta),
  counties = n_distinct(mpdta$countyreal),
  years = n_distinct(mpdta$year)
)
cohort_table

stopifnot(
  !anyDuplicated(mpdta[c("countyreal", "year")]),
  length(unique(county_audit$observations)) == 1L,
  all(county_audit$cohort_values == 1L),
  !any(county_audit$treatment_reversal),
  any(mpdta$first.treat == 0),
  n_distinct(mpdta$first.treat[mpdta$first.treat > 0]) > 1L
)
```

## Step 2: Declare Group-Time Comparisons

For cohort $g$ in year $t$, the target is

$$
ATT(g,t) = E[Y_t(1) - Y_t(0) \mid G=g].
$$

The counterfactual can be built from:

- **never-treated counties**, which remain untreated throughout the panel; or
- **not-yet-treated counties**, which are untreated at $t$ but may adopt later.

```{r}
comparison_support <- mpdta |>
  distinct(countyreal, year, first.treat) |>
  group_by(year) |>
  summarise(
    never_treated = sum(first.treat == 0),
    not_yet_treated = sum(first.treat > year),
    eligible_never_control = never_treated,
    eligible_not_yet_control = never_treated + not_yet_treated,
    .groups = "drop"
  )

comparison_support

stopifnot(
  all(comparison_support$eligible_not_yet_control >=
    comparison_support$eligible_never_control),
  all(comparison_support$eligible_never_control > 0)
)
```

The not-yet-treated option can add support in early years but requires those future adopters to be valid controls before treatment.

County clustering is used to reproduce the package teaching example. Minimum-wage assignment may induce dependence across counties within a state; policy inference needs assignment-level identifiers and clustering justified by that design.

## Step 3: Plot Cohort Outcome Paths

```{r, fig.height=5.2, fig.width=10}
cohort_trends <- mpdta |>
  mutate(
    cohort = if_else(
      first.treat == 0,
      "Never treated",
      paste("First treated", first.treat)
    )
  ) |>
  group_by(cohort, first.treat, year) |>
  summarise(mean_log_employment = mean(lemp), .groups = "drop")

ggplot(
  cohort_trends,
  aes(year, mean_log_employment, colour = cohort, group = cohort)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.7) +
  labs(
    x = NULL,
    y = "Mean log teen employment",
    colour = "Cohort",
    title = "Treatment cohorts contribute different pre- and post-treatment windows"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Checkpoint: which cohort contributes the longest post-treatment history, and why can that cohort receive disproportionate weight in a simple average?

## Step 4: Estimate Unconditional Group-Time Effects

The [`att_gt()` documentation](https://bcallaway11.github.io/did/reference/att_gt.html) makes the outcome, time, unit, cohort, comparison group, anticipation period, and base period explicit. We use a universal base period so pre-treatment pseudo-effects share the same interpretation and can feed later sensitivity analysis.

```{r}
set.seed(20260809)

att_never <- att_gt(
  yname = "lemp",
  tname = "year",
  idname = "countyreal",
  gname = "first.treat",
  xformla = ~ 1,
  data = mpdta,
  control_group = "nevertreated",
  anticipation = 0,
  base_period = "universal",
  est_method = "dr",
  bstrap = TRUE,
  biters = 499,
  cband = TRUE,
  clustervars = "countyreal",
  print_details = FALSE
)

summary(att_never)

stopifnot(
  inherits(att_never, "MP"),
  length(att_never$att) > 0,
  any(is.finite(att_never$att)),
  att_never$DIDparams$control_group == "nevertreated"
)
```

The default doubly robust score combines outcome regression and propensity-score components. With `xformla = ~ 1`, the design still invokes unconditional rather than covariate-conditional parallel trends.

## Step 5: Control The Aggregation

There is no single compulsory way to turn $ATT(g,t)$ into one result. Each aggregation answers a different question [@callaway2021multipletime].

```{r}
aggregate_simple <- aggte(att_never, type = "simple", na.rm = TRUE)
aggregate_group <- aggte(att_never, type = "group", na.rm = TRUE)
aggregate_dynamic <- aggte(att_never, type = "dynamic", na.rm = TRUE)
aggregate_calendar <- aggte(att_never, type = "calendar", na.rm = TRUE)

aggregation_summary <- data.frame(
  aggregation = c("Simple", "Group", "Dynamic", "Calendar"),
  overall_att = c(
    aggregate_simple$overall.att,
    aggregate_group$overall.att,
    aggregate_dynamic$overall.att,
    aggregate_calendar$overall.att
  ),
  standard_error = c(
    aggregate_simple$overall.se,
    aggregate_group$overall.se,
    aggregate_dynamic$overall.se,
    aggregate_calendar$overall.se
  )
)

aggregation_summary

stopifnot(
  nrow(aggregation_summary) == 4L,
  all(is.finite(aggregation_summary$overall_att)),
  all(aggregation_summary$standard_error > 0),
  length(aggregate_dynamic$egt) == length(aggregate_dynamic$att.egt)
)
```

Interpret the rows separately:

- `simple` averages available post-treatment group-time effects;
- `group` gives cohort-specific averages before combining cohorts;
- `dynamic` organizes effects by exposure length;
- `calendar` organizes effects by calendar year.

## Step 6: Plot Dynamic Effects With Simultaneous Bands

```{r, fig.height=5.2, fig.width=10}
ggdid(aggregate_dynamic) +
  labs(
    x = "Periods relative to treatment",
    y = "ATT",
    title = "Callaway–Sant'Anna dynamic effects",
    subtitle = "Bands are simultaneous because cband = TRUE"
  ) +
  theme_minimal(base_size = 12)
```

Simultaneous bands address uncertainty across the displayed path. They do not turn pre-treatment estimates into proof of parallel trends or solve anticipation and spillover concerns.

## Step 7: Change The Comparison Group

```{r}
set.seed(20260809)

att_not_yet <- att_gt(
  yname = "lemp",
  tname = "year",
  idname = "countyreal",
  gname = "first.treat",
  xformla = ~ 1,
  data = mpdta,
  control_group = "notyettreated",
  anticipation = 0,
  base_period = "universal",
  est_method = "dr",
  bstrap = TRUE,
  biters = 499,
  cband = TRUE,
  clustervars = "countyreal",
  print_details = FALSE
)

dynamic_not_yet <- aggte(att_not_yet, type = "dynamic", na.rm = TRUE)

control_group_comparison <- data.frame(
  comparison_group = c("Never treated", "Not yet treated"),
  dynamic_overall_att = c(
    aggregate_dynamic$overall.att,
    dynamic_not_yet$overall.att
  ),
  standard_error = c(
    aggregate_dynamic$overall.se,
    dynamic_not_yet$overall.se
  )
)

control_group_comparison

stopifnot(
  att_not_yet$DIDparams$control_group == "notyettreated",
  all(is.finite(control_group_comparison$dynamic_overall_att)),
  all(control_group_comparison$standard_error > 0)
)
```

Do not select a control group because it produces the preferred coefficient. Defend whether future-treated counties are plausible untreated counterfactuals in each period.

## Step 8: Condition On Population

```{r}
set.seed(20260809)

att_conditional <- att_gt(
  yname = "lemp",
  tname = "year",
  idname = "countyreal",
  gname = "first.treat",
  xformla = ~ lpop,
  data = mpdta,
  control_group = "nevertreated",
  anticipation = 0,
  base_period = "universal",
  est_method = "dr",
  bstrap = TRUE,
  biters = 499,
  cband = TRUE,
  clustervars = "countyreal",
  print_details = FALSE
)

dynamic_conditional <- aggte(att_conditional, type = "dynamic", na.rm = TRUE)

parallel_trends_comparison <- data.frame(
  assumption = c(
    "Unconditional parallel trends",
    "Parallel trends conditional on log population"
  ),
  dynamic_overall_att = c(
    aggregate_dynamic$overall.att,
    dynamic_conditional$overall.att
  ),
  standard_error = c(
    aggregate_dynamic$overall.se,
    dynamic_conditional$overall.se
  )
)

parallel_trends_comparison

stopifnot(
  all(is.finite(parallel_trends_comparison$dynamic_overall_att)),
  all(parallel_trends_comparison$standard_error > 0)
)
```

Adding `lpop` changes the identifying assumption: untreated trends must now be parallel after conditioning on log population. It is defensible only if population is measured before the relevant outcome change and the conditional overlap is adequate.

## Step 9: Compare With Sun–Abraham

```{r}
sun_abraham_fit <- feols(
  lemp ~ sunab(first.treat, year, ref.p = -1) | countyreal + year,
  data = mpdta,
  vcov = ~ countyreal
)

sun_abraham_terms <- data.frame(
  event_time = as.integer(sub(".*::", "", names(coef(sun_abraham_fit)))),
  estimate = unname(coef(sun_abraham_fit)),
  standard_error = unname(se(sun_abraham_fit))
)

callaway_santanna_terms <- data.frame(
  event_time = aggregate_dynamic$egt,
  estimate = aggregate_dynamic$att.egt,
  standard_error = aggregate_dynamic$se.egt
)

modern_event_comparison <- bind_rows(
  transform(callaway_santanna_terms, estimator = "Callaway–Sant'Anna"),
  transform(sun_abraham_terms, estimator = "Sun–Abraham")
) |>
  mutate(
    lower_95 = estimate - 1.96 * standard_error,
    upper_95 = estimate + 1.96 * standard_error
  ) |>
  filter(
    is.finite(estimate),
    is.finite(standard_error),
    event_time >= -4,
    event_time <= 3
  )

modern_event_comparison

stopifnot(
  all(c("Callaway–Sant'Anna", "Sun–Abraham") %in%
    modern_event_comparison$estimator),
  any(modern_event_comparison$event_time < 0),
  any(modern_event_comparison$event_time >= 0),
  all(is.finite(modern_event_comparison$standard_error))
)
```

```{r, fig.height=5.2, fig.width=10}
ggplot(
  modern_event_comparison,
  aes(event_time, estimate, colour = estimator)
) +
  geom_hline(yintercept = 0, colour = "grey55") +
  geom_vline(xintercept = -1, linetype = "dashed", colour = "grey55") +
  geom_line(linewidth = 0.75) +
  geom_point(size = 2) +
  labs(
    x = "Periods relative to treatment",
    y = "Estimated effect on log teen employment",
    colour = NULL,
    title = "Modern estimators need not produce identical weighted targets"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Differences can arise from cohort weights, available comparisons, nuisance estimation, and support. Numerical agreement is reassuring but is not the criterion that defines a valid estimator.

## Optional Extension: Sensitivity To Parallel-Trends Violations

`HonestDiD` requires an ordered event-study coefficient vector and its full covariance matrix. To keep the exercise transparent, restrict the data to the 2006 cohort and never-treated counties. This creates a non-staggered design with two estimated pre-treatment coefficients, an omitted 2005 reference period, and two post-treatment coefficients.

```{r}
if (!requireNamespace("HonestDiD", quietly = TRUE)) {
  install.packages("HonestDiD", repos = "https://cloud.r-project.org")
}

sensitivity_sample <- mpdta |>
  filter(first.treat %in% c(0, 2006)) |>
  mutate(treated_2006 = as.integer(first.treat == 2006))

cohort_event_fit <- feols(
  lemp ~ i(year, treated_2006, ref = 2005) | countyreal + year,
  data = sensitivity_sample,
  vcov = ~ countyreal
)

event_coefficients <- coef(cohort_event_fit)
event_covariance <- vcov(cohort_event_fit)
term_year <- as.integer(sub(
  "year::([0-9]+).*",
  "\\1",
  names(event_coefficients)
))
term_order <- order(term_year)

event_coefficients <- event_coefficients[term_order]
event_covariance <- event_covariance[term_order, term_order, drop = FALSE]
term_year <- term_year[term_order]

data.frame(
  year = term_year,
  period = c("Pre", "Pre", "Post", "Post"),
  estimate = unname(event_coefficients),
  standard_error = sqrt(diag(event_covariance))
)

stopifnot(
  identical(term_year, c(2003L, 2004L, 2006L, 2007L)),
  length(event_coefficients) == 4L,
  all(dim(event_covariance) == c(4L, 4L)),
  all(is.finite(event_covariance))
)
```

The relative-magnitudes restriction indexes how large post-treatment deviations from parallel trends may be relative to the largest pre-treatment change in the untreated group gap. The procedure accounts for sampling uncertainty in the pre-treatment coefficients; it does not treat the observed maximum as known.

```{r}
sensitivity_results <- HonestDiD::createSensitivityResults_relativeMagnitudes(
  betahat = event_coefficients,
  sigma = event_covariance,
  numPrePeriods = 2,
  numPostPeriods = 2,
  Mbarvec = seq(0.5, 2, by = 0.5),
  gridPoints = 100,
  seed = 20260809
)

original_interval <- HonestDiD::constructOriginalCS(
  betahat = event_coefficients,
  sigma = event_covariance,
  numPrePeriods = 2,
  numPostPeriods = 2
)

zero_included <- with(
  sensitivity_results,
  lb <= 0 & ub >= 0
)
first_zero_including_mbar <- if (any(zero_included)) {
  min(sensitivity_results$Mbar[zero_included])
} else {
  NA_real_
}

sensitivity_results
data.frame(first_Mbar_including_zero = first_zero_including_mbar)

stopifnot(
  nrow(sensitivity_results) == 4L,
  all(sensitivity_results$lb <= sensitivity_results$ub),
  all(is.finite(sensitivity_results$Mbar)),
  nrow(original_interval) == 1L
)
```

```{r, fig.height=5.2, fig.width=9}
HonestDiD::createSensitivityPlot_relativeMagnitudes(
  sensitivity_results,
  original_interval
)
```

The default target is the first post-treatment effect (2006), not the average of both post-treatment periods. The sensitivity interval for this cohort already includes zero at the first reported restriction value. The exercise is still informative: sensitivity analysis should report how conclusions change under stated violations, not manufacture a decisive result.

Most importantly, this is a **2006-cohort-versus-never-treated** estimand. It is not a sensitivity analysis for the full staggered-adoption aggregate estimated earlier.

## Practical Implications

- Estimate $ATT(g,t)$ before deciding how to aggregate heterogeneous effects.
- Comparison-group choice changes both support and the parallel-trends assumption.
- Covariate adjustment replaces unconditional with conditional parallel trends; it does not weaken the need for a credible design.
- Dynamic estimators can differ because their targets and weights differ.
- Sensitivity analysis should name the estimand, restriction, and breakdown point together.
- No modern package repairs anticipation, spillovers, outcome miscoding, or a substantively indefensible comparison group.

## Worked answers

- A simple aggregate gives early cohorts more post-treatment cells; cohort, calendar, and event averages answer different questions.
- Not-yet-treated counties add support only while untreated; they add a substantive comparison assumption.
- A universal baseline compares each lead to the last untreated period, whereas a varying baseline uses adjacent pre-period changes.
- The sensitivity exercise concerns the 2006 cohort and first post-period only. Report the tested grid rather than claiming an exact breakdown threshold.
