---
title: "Difference-in-Differences for Causal Inference"
subtitle: "Parallel trends, comparison groups, and heterogeneous treatment timing"
author: "Luke Heley"
date: 19 September 2026
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  fig-width: 7
  fig-height: 4
---

## Executive summary

Difference-in-differences (DiD) compares an exposed group's change with a credible
unexposed group's change. It allows baseline differences in outcome levels, but
requires an argument about the exposed group's unobserved untreated trend. A
regression interaction implements the contrast; it cannot establish that argument.

This report starts with Kentucky workers' compensation, where the four cell means
make the comparison transparent. It then uses a bundled minimum-wage teaching
panel to show why staggered adoption requires explicit cohort-time effects and
aggregation. The goal is a defensible comparison, not agreement among package
outputs [@gertler2016ch7did; @roth2023trending].

## Design and target effect

### Specify who changes treatment and when

Define the policy, eligibility, exposure date, outcome scale, target population,
and comparison group before looking at estimates. Distinguish policy adoption
from announcement, implementation, and behavioral response. Audit anticipation,
spillovers, repeated treatment, reversals, and policy co-adoption.

In the Kentucky example, a 1980 increase in the benefit cap affected high earners
differently from low earners. The outcome is log weeks receiving benefits. The
observations are claims before and after the reform, not a panel of the same
claimants. Changes in who claims, injury severity, or industry composition can
therefore alter the comparison [@meyer1995workerscomp].

### Identify the counterfactual change

Let $G=1$ denote the exposed group and $t=1$ the post-period. The target is
$ATT=E[Y_1(1)-Y_1(0)\mid G=1]$. With no anticipation, parallel untreated trends
requires

$$
E[Y_1(0)-Y_0(0)\mid G=1]
=E[Y_1(0)-Y_0(0)\mid G=0].
$$

Consistency, no relevant interference, and comparable population measurement are
also needed. The assumption is about untreated potential outcomes, not equality
of observed post-policy trends. Baseline levels may differ. Parallel trends in
levels need not imply parallel trends in logs, so the outcome scale is part of
the identifying claim.

Relative to matching, DiD can accommodate time-invariant differences associated
with outcome levels, but not arbitrary group-specific shocks. Relative to ITS,
it borrows a contemporaneous comparison trend rather than projecting only the
treated series. A poor comparison group is not repaired by adding fixed effects.

## Kentucky: a transparent four-cell analysis

The [data catalogue](data.qmd#injury) documents the local snapshot. Execution assumes
the R packages are installed and never downloads data or installs software.

```{r}
source("data/load-data.R")
library(ggplot2)
library(did)
ky <- subset(qed_data("injury"), ky == 1)
stopifnot(nrow(ky) == 5626, all(ky$durat > 0),
          !anyNA(ky[c("ldurat", "highearn", "afchnge")]))
means <- with(ky, tapply(ldurat, list(high=highearn, after=afchnge), mean))
counts <- with(ky, table(high=highearn, after=afchnge))
knitr::kable(means, digits=4, caption="Mean log benefit duration")
knitr::kable(counts, caption="Claims in each group-period cell")
```

Every cell must represent a substantively comparable population. Record exclusions
and changes in composition rather than letting complete-case regression define
the sample silently. Repeated cross-sections require stable population definitions;
panel data instead introduce concerns about attrition and individual histories.

### Calculate and recover the contrast

$$
\widehat{ATT}=(\bar Y_{1,1}-\bar Y_{1,0})
             -(\bar Y_{0,1}-\bar Y_{0,0}).
$$

```{r}
comparison_change <- means["0", "1"] - means["0", "0"]
counterfactual <- means["1", "0"] + comparison_change
manual <- means["1", "1"] - counterfactual
model <- lm(ldurat ~ highearn * afchnge, data=ky)
interaction <- unname(coef(model)["highearn:afchnge"])
stopifnot(abs(manual-interaction) < 1e-10,
          abs(manual-0.1906012007) < 1e-8)
knitr::kable(data.frame(manual, interaction,
                        geometric_percent=100*expm1(manual)), digits=4)
```

The interaction in $Y=\alpha+\gamma G+\lambda Post+\tau(G\times Post)+u$
reproduces the four-cell contrast. Algebraic equality checks coding; it does not
validate parallel trends. The estimate is about 0.191 log points. Its exponential
transformation is a contrast of geometric means, not automatically the percentage
effect on arithmetic mean benefit duration.

```{r}
#| fig-cap: "The exposed group's counterfactual adds the low-earner change to its own baseline."
#| fig-alt: "High- and low-earner mean log durations before and after reform, with the high-earner counterfactual shown by a dashed line."
p <- aggregate(ldurat ~ afchnge + highearn, data=ky, FUN=mean)
cf <- data.frame(afchnge=c(0,1), ldurat=c(means["1","0"], counterfactual))
ggplot(p, aes(afchnge, ldurat, color=factor(highearn))) +
  geom_line(linewidth=0.8) + geom_point(size=2) +
  geom_line(data=cf, aes(afchnge, ldurat), inherit.aes=FALSE,
            linetype=2, color="#177b72") +
  scale_color_manual(values=c("#245ca4", "#177b72"),
                     labels=c("Low earners", "High earners")) +
  scale_x_continuous(breaks=c(0,1), labels=c("Before", "After")) +
  labs(x=NULL, y="Mean log benefit duration", color=NULL) +
  theme_minimal(base_size=11) + theme(legend.position="bottom")
```

### Separate estimation from inference

Thousands of claims do not imply thousands of independently assigned policy
changes. An individual-level regression standard error cannot manufacture policy
replication. In richer applications, cluster at a level consistent with assignment
and dependence, examine the number of independent clusters, and use a defensible
small-sample strategy when clusters are few [@bertrand2004trustdid].

This four-cell sample contains no multi-period pre-trend evidence. An honest
interpretation is a conditional contrast accompanied by institutional arguments
and composition checks, not a fully diagnosed causal conclusion. Baseline
covariates may support conditional parallel trends if they are unaffected by
treatment and have adequate overlap. Post-treatment adjustment can remove a
mechanism or introduce selection bias [@santannazhao2020drdid].

## Staggered adoption changes the comparison

When cohorts adopt in different periods, a conventional two-way fixed-effects
coefficient can compare newly treated units with already-treated units. If effects
change with exposure duration or differ across cohorts, those comparisons can
contaminate the intended policy average. Conventional event-study leads and lags
can also mix heterogeneous effects [@goodmanbacon2021timing; @sunabraham2021dynamic].

Define $ATT(g,t)=E[Y_t(g)-Y_t(0)\mid G=g]$ for cohort $g$ and calendar time $t$.
Specify never-treated or not-yet-treated controls, an anticipation window, and the
periods with actual comparison support. Treatment reversals or repeated doses need
a method appropriate to those histories; an absorbing-treatment estimator is not
automatically applicable.

### Estimate group-time effects

```{r}
mpdta <- qed_data("mpdta")
stopifnot(nrow(mpdta) == 2500,
          length(unique(mpdta$countyreal)) == 500,
          !anyDuplicated(mpdta[c("countyreal", "year")]))
knitr::kable(with(mpdta, table(cohort=first.treat, year=year)))
set.seed(20260809)
att <- att_gt(yname="lemp", tname="year", idname="countyreal",
  gname="first.treat", data=mpdta, xformla=~1,
  control_group="nevertreated", anticipation=0,
  base_period="universal", est_method="dr",
  bstrap=TRUE, biters=499, cband=TRUE, clustervars="countyreal")
overall <- aggte(att, type="simple")
dynamic <- aggte(att, type="dynamic")
stopifnot(abs(overall$overall.att-(-0.0399512752)) < 1e-8)
knitr::kable(data.frame(ATT=overall$overall.att,
                        SE=overall$overall.se), digits=4)
```

The panel contains 500 counties over five years. Zero in `first.treat` means
never treated. The example uses log teen employment, never-treated controls,
zero anticipation, and an unconditional specification. The universal base period
normalizes contrasts to the final pre-treatment period. A doubly robust estimator
name does not make the identifying assumptions true [@callaway2021multipletime].

The seed and 499 bootstrap draws make this a reproducible classroom exercise.
Investigate simulation stability for a substantive analysis. County clustering
here is a teaching implementation, not a defense of county-level inference for
a state-level policy without the appropriate assignment identifiers.

### Aggregation is part of the estimand

```{r}
#| fig-cap: "Dynamic aggregation with simultaneous bands from the stated teaching specification."
#| fig-alt: "Event-time treatment effects on log employment with simultaneous confidence bands."
ggdid(dynamic) + labs(x="Years relative to treatment", y="ATT on log employment")
```

The simple aggregate is approximately -0.040 log points. Earlier cohorts can
contribute more post-treatment cells, so it is not automatically an equally
weighted average across cohorts or exposure durations. Dynamic aggregation groups
effects by $e=t-g$. At longer horizons, fewer cohorts may contribute. A changing
event-time profile can therefore reflect both dynamics and changing composition.
State the weights, supported horizons, and any balanced-exposure restriction.

## Diagnostics, sensitivity, and extensions

Plot raw outcome paths and cohort sizes before adjusted event studies. Audit
missingness, differential composition, anticipation, spillovers, and co-occurring
policies. Compare eligible control definitions only when both are substantively
defensible; a changed estimate can reflect a changed target or changed support.

Pre-treatment coefficients can reveal some violations. Failure to reject does
not establish parallel trends, especially with few pre-periods or low power.
Choosing a model because it passes a pre-test also changes the inferential
problem [@roth2022pretest]. Simultaneous bands address a collection of estimates;
pointwise intervals answer a different coverage question.

The diagnostics lab implements the Goodman-Bacon decomposition and a Sun-Abraham
event study. The modern-estimators lab adds comparison-group choices and
`HonestDiD` sensitivity to departures from parallel trends. Sensitivity restrictions
must be explained in substantive units; they do not certify the design
[@rambachanroth2023parallel]. Imputation estimators make the untreated outcome
model explicit and offer another approach when their assumptions and treatment
histories fit [@borusyak2024eventstudy].

Do not rank methods by the smallest p-value. Explain which comparisons each uses,
which population it represents, and which heterogeneity it accommodates. Agreement
is useful as a coding check but does not rule out a shared identifying failure.

## Reporting and limits

Report the exposure rule, dates, outcome scale, population, comparison group,
sample exclusions, anticipation assumptions, estimator, aggregation weights, and
dependence model. Include raw paths, effect estimates and intervals, support at
each horizon, diagnostic evidence, and sensitivity. Name the strongest plausible
differential shock that the design cannot exclude.

For Kentucky, foreground repeated cross-sections and absent multi-period pre-trend
evidence. For the minimum-wage panel, foreground the teaching sample and inference
limitations. Neither example should be described as a new full replication of
the original policy study.

## Teaching route

- [Teaching deck](../slides/did.qmd) and [presenter notes](../slides/did_script.qmd).
- [Practical R guide](../notes/did/how-to-do-difference-in-differences.qmd) for the shorter walkthrough.
- [Foundations lab](difference-in-differences-foundations-lab.qmd) for the four-cell calculation and regression.
- [Staggered diagnostics lab](difference-in-differences-staggered-diagnostics-lab.qmd) for comparison weights and event studies.
- [Modern estimators lab](difference-in-differences-modern-estimators-lab.qmd) for group-time effects, aggregation, and sensitivity.
- [Reproduction record](difference-in-differences-reproducibility.qmd) for the environment and verified outputs.

All three exercises have Quarto, R Jupyter, and offline ZIP downloads in the
[Labs directory](index.qmd#difference-in-differences-labs).

## References
