---
title: "Lab: HISP IE in Practice"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as a guided replication worksheet for the HISP case-study material from *Impact Evaluation in Practice*.

- Run the sections in the same order as the original Stata do-file.
- Treat the early methods as design contrasts, not as equally credible estimates.
- Keep notes on how the estimate moves as the design gets stronger.
- Use the in-page code blocks as the primary worksheet sequence in this repo.

::: {.callout-note}
Unlike the other labs in this repository, this page is organized to follow the logic of `docs/data/hisp_ie_in_practice/HISP Book replication.do` as closely as possible. The aim is to translate that Stata workflow into R, not to rewrite it into a different teaching sequence.
:::

## Source Baseline

This lab uses the tracked build bundle in `docs/data/hisp_ie_in_practice`:

- primary workflow baseline: [HISP Book replication.do](../data/hisp_ie_in_practice/HISP%20Book%20replication.do)
- Stata input dataset: [evaluation.dta](../data/hisp_ie_in_practice/evaluation.dta)
- CSV analysis dataset used for rendering: [evaluation.csv](../data/hisp_ie_in_practice/evaluation.csv)

The renderable bundle lives under `docs/` so the site can build without depending on ignored raw-data paths in the repository root.

## Step 1: Convert The Stata Files And Load The HISP Data

```{r}
required_packages <- c(
  "haven",
  "readr",
  "dplyr",
  "ggplot2",
  "MatchIt",
  "sandwich",
  "boot",
  "broom",
  "nlme"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

data_dir <- "../data/hisp_ie_in_practice"
dta_files <- list.files(data_dir, pattern = "[.]dta$", full.names = TRUE)

csv_export_manifest <- dplyr::bind_rows(lapply(dta_files, function(dta_path) {
  csv_path <- sub("[.]dta$", ".csv", dta_path)
  csv_exists <- file.exists(csv_path)
  dat <- haven::read_dta(dta_path)

  if (!csv_exists) {
    readr::write_csv(dat, csv_path)
  }

  data.frame(
    dta_file = basename(dta_path),
    csv_file = basename(csv_path),
    csv_status = ifelse(csv_exists, "tracked_existing", "exported_from_dta"),
    rows = nrow(dat),
    columns = ncol(dat)
  )
}))

dat <- readr::read_csv(file.path(data_dir, "evaluation.csv"), show_col_types = FALSE)

controls <- c(
  "age_hh",
  "age_sp",
  "educ_hh",
  "educ_sp",
  "female_hh",
  "indigenous",
  "hhsize",
  "dirtfloor",
  "bathroom",
  "land",
  "hospital_distance"
)

dat |>
  dplyr::select(
    household_identifier,
    locality_identifier,
    treatment_locality,
    promotion_locality,
    eligible,
    enrolled,
    round,
    health_expenditures
  ) |>
  head(10)
```

Checkpoint:

- Does the tracked `evaluation.csv` keep the same `19,827` rows and `22` columns as the Stata source?
- If any CSV file was missing, did the export manifest recreate it from the bundled `.dta` file?

## Step 2: Method 1, Before-After Among Enrolled Households

The Stata do-file starts with the simplest comparison: enrolled households in treatment localities, before versus after.

```{r}
method_1_dat <- dat |>
  dplyr::filter(treatment_locality == 1, enrolled == 1)

t.test(
  health_expenditures ~ factor(round, levels = c(0, 1)),
  data = method_1_dat,
  var.equal = TRUE
)

fit_m1_raw <- lm(health_expenditures ~ round, data = method_1_dat)
sandwich::vcovCL(fit_m1_raw, cluster = method_1_dat$locality_identifier, type = "HC1")

fit_m1_controls <- lm(
  health_expenditures ~ round + age_hh + age_sp + educ_hh + educ_sp +
    female_hh + indigenous + hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = method_1_dat
)
```

What to note:

- This estimate is easy to compute but vulnerable to time trends.
- The control-adjusted version follows the do-file, but it does not solve the design problem.

## Step 3: Method 2, Enrolled Versus Not Enrolled At Follow-Up

This is the second naive comparison in the book replication script: follow-up outcomes in treatment localities, comparing enrolled and nonenrolled households.

```{r}
method_2_dat <- dat |>
  dplyr::filter(treatment_locality == 1, round == 1)

t.test(
  health_expenditures ~ factor(enrolled, levels = c(0, 1)),
  data = method_2_dat,
  var.equal = TRUE
)

fit_m2_raw <- lm(health_expenditures ~ enrolled, data = method_2_dat)
fit_m2_controls <- lm(
  health_expenditures ~ enrolled + age_hh + age_sp + educ_hh + educ_sp +
    female_hh + indigenous + hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = method_2_dat
)
```

## Step 4: Method 3, Randomized Assignment Among Eligible Households

The book then moves to the randomized-assignment case, using eligible households and comparing treatment versus comparison villages.

```{r}
method_3_dat <- dat |>
  dplyr::filter(eligible == 1)

t.test(
  health_expenditures ~ factor(treatment_locality, levels = c(0, 1)),
  data = dplyr::filter(method_3_dat, round == 0),
  var.equal = TRUE
)

balance_controls <- dplyr::bind_rows(lapply(controls, function(var_name) {
  t.test(
    method_3_dat[[var_name]][method_3_dat$round == 0] ~
      factor(method_3_dat$treatment_locality[method_3_dat$round == 0], levels = c(0, 1)),
    var.equal = TRUE
  ) |>
    broom::tidy() |>
    dplyr::mutate(variable = var_name)
}))

fit_m3_raw <- lm(
  health_expenditures ~ treatment_locality,
  data = dplyr::filter(method_3_dat, round == 1)
)

fit_m3_controls <- lm(
  health_expenditures ~ treatment_locality + age_hh + age_sp + educ_hh + educ_sp +
    female_hh + indigenous + hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = dplyr::filter(method_3_dat, round == 1)
)
```

## Step 5: Method 4, Instrumental Variables Under Random Promotion

The IV branch uses `promotion_locality` as the instrument for `enrolled_rp` at follow-up.

```{r}
method_4_dat <- dat |>
  dplyr::filter(round == 1)

t.test(
  enrolled_rp ~ factor(promotion_locality, levels = c(0, 1)),
  data = method_4_dat,
  var.equal = TRUE
)

first_stage <- lm(
  enrolled_rp ~ promotion_locality + age_hh + age_sp + educ_hh + educ_sp +
    female_hh + indigenous + hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = method_4_dat
)

# Draft placeholder:
# replace this with an in-repo helper or inline 2SLS workflow once the
# full HISP execution bundle is migrated.
# method_4_iv_regressions
```

## Step 6: Method 5, Regression Discontinuity

The do-file normalizes the poverty index around the `58` cutoff and estimates a piecewise-linear discontinuity model in treatment localities.

```{r}
method_5_dat <- dat |>
  dplyr::filter(treatment_locality == 1) |>
  dplyr::mutate(
    poverty_index_left = ifelse(poverty_index <= 58, poverty_index - 58, 0),
    poverty_index_right = ifelse(poverty_index > 58, poverty_index - 58, 0)
  )

fit_rdd_followup <- lm(
  health_expenditures ~ poverty_index_left + poverty_index_right + eligible,
  data = dplyr::filter(method_5_dat, round == 1)
)

fit_rdd_controls <- lm(
  health_expenditures ~ eligible + poverty_index_left + poverty_index_right +
    age_hh + age_sp + educ_hh + educ_sp + female_hh + indigenous +
    hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = dplyr::filter(method_5_dat, round == 1)
)

ggplot2::ggplot(
  dplyr::filter(method_5_dat, round == 1, health_expenditures < 60),
  ggplot2::aes(x = poverty_index, y = health_expenditures)
) +
  ggplot2::geom_point(alpha = 0.12) +
  ggplot2::geom_vline(xintercept = 58, linetype = "dashed")
```

## Step 7: Method 6, Difference-in-Differences

The DiD branch stays in treatment localities and adds the `enrolled * round` interaction.

```{r}
method_6_dat <- dat |>
  dplyr::filter(treatment_locality == 1) |>
  dplyr::mutate(enrolled_round = enrolled * round)

t.test(
  health_expenditures ~ factor(round, levels = c(0, 1)),
  data = dplyr::filter(method_6_dat, enrolled == 0),
  var.equal = TRUE
)

t.test(
  health_expenditures ~ factor(round, levels = c(0, 1)),
  data = dplyr::filter(method_6_dat, enrolled == 1),
  var.equal = TRUE
)

fit_did_raw <- lm(
  health_expenditures ~ enrolled_round + round + enrolled,
  data = method_6_dat
)

fit_did_controls <- lm(
  health_expenditures ~ enrolled_round + round + enrolled +
    age_hh + age_sp + educ_hh + educ_sp + female_hh + indigenous +
    hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = method_6_dat
)
```

## Step 8: Method 7, Propensity-Score Matching

The matching section in the do-file reshapes the panel to wide form, estimates a probit propensity score, randomizes row order, and runs nearest-neighbor matching first on a restricted covariate set and then on the full control set.

```{r}
wide_dat <- reshape(
  as.data.frame(dat),
  direction = "wide",
  idvar = "household_identifier",
  timevar = "round",
  sep = ""
)

coalesce_pair <- function(left, right) ifelse(is.na(left), right, left)

for (var_name in c(
  "enrolled", "age_hh", "age_sp", "educ_hh", "educ_sp", "female_hh",
  "indigenous", "hhsize", "dirtfloor", "bathroom", "land", "hospital_distance"
)) {
  wide_dat[[var_name]] <- coalesce_pair(wide_dat[[paste0(var_name, "0")]], wide_dat[[paste0(var_name, "1")]])
}

matching_dat <- wide_dat[
  complete.cases(wide_dat[, c(
    "enrolled",
    "health_expenditures0",
    "health_expenditures1",
    "age_hh",
    "educ_hh",
    "age_sp",
    "educ_sp",
    "female_hh",
    "indigenous",
    "hhsize",
    "dirtfloor",
    "bathroom",
    "land",
    "hospital_distance"
  )]),
]

probit_fit <- glm(
  enrolled ~ age_hh + educ_hh,
  data = matching_dat,
  family = binomial(link = "probit")
)

matching_dat$pscore <- predict(probit_fit, type = "response")

set.seed(100)
m_restricted <- MatchIt::matchit(
  enrolled ~ age_hh + educ_hh,
  data = matching_dat,
  method = "nearest",
  distance = "glm",
  link = "probit",
  ratio = 1,
  replace = FALSE,
  m.order = "random",
  estimand = "ATT",
  normalize = FALSE
)

matched_restricted <- MatchIt::match.data(m_restricted)
lm(health_expenditures1 ~ enrolled, data = matched_restricted, weights = weights)

set.seed(100)
m_full <- MatchIt::matchit(
  enrolled ~ age_hh + age_sp + educ_hh + educ_sp + female_hh + indigenous +
    hhsize + dirtfloor + bathroom + land + hospital_distance,
  data = matching_dat,
  method = "nearest",
  distance = "glm",
  link = "probit",
  ratio = 1,
  replace = FALSE,
  m.order = "random",
  estimand = "ATT",
  normalize = FALSE
)
```

For the full-set branch, swap `age_hh + educ_hh` for the full control list used earlier.

## Step 9: Matched Difference-in-Differences

The do-file closes the matching chapter by computing a matched DiD on the full matched sample.

```{r}
matched_full <- MatchIt::match.data(m_full)

matched_pairs <- matched_full |>
  dplyr::arrange(subclass, dplyr::desc(enrolled)) |>
  dplyr::group_by(subclass) |>
  dplyr::summarise(
    matched_dd = (health_expenditures1[enrolled == 1] - health_expenditures0[enrolled == 1]) -
      (health_expenditures1[enrolled == 0] - health_expenditures0[enrolled == 0]),
    .groups = "drop"
  )

matched_full |>
  dplyr::mutate(diff = health_expenditures1 - health_expenditures0) |>
  lm(diff ~ enrolled, data = _, weights = weights)
```

## Step 10: Power Calculations

The final branch follows the book's power-calculation logic using follow-up treated means, standard deviations, and ICC estimates.

```{r}
two_sample_n_per_arm <- function(sd_outcome, minimum_detectable_effect, power, alpha = 0.05) {
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(power)
  ceiling(((z_alpha + z_beta)^2 * (sd_outcome^2 + sd_outcome^2)) / minimum_detectable_effect^2)
}

power_dat <- dat |>
  dplyr::filter(eligible == 1, round == 1, treatment_locality == 1)

m1 <- mean(power_dat$health_expenditures, na.rm = TRUE)
sd1 <- sd(power_dat$health_expenditures, na.rm = TRUE)

data.frame(
  minimum_detectable_effect = c(1, 2, 3),
  target_mean = m1 - c(1, 2, 3),
  n_per_arm_90 = vapply(c(1, 2, 3), two_sample_n_per_arm, numeric(1), sd_outcome = sd1, power = 0.9),
  n_per_arm_80 = vapply(c(1, 2, 3), two_sample_n_per_arm, numeric(1), sd_outcome = sd1, power = 0.8)
)
```

## Comparison Question

After running the full sequence, compare the main effect terms across methods:

- Which branches are obviously biased because the design is too weak?
- Which branches move closest to the more credible design-based estimates?
- How much do the estimates shift once the workflow reaches RDD, DiD, IV, and matching?

The evaluated-code page collects those outputs into one place, and the QA log documents where the R workflow intentionally mirrors the Stata source.
