---
title: "Lab: Interrupted Time Series Data and Mechanics"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this as the first interrupted time-series lab.

[Download the R Jupyter notebook](notebooks/interrupted-time-series-mechanics-lab.ipynb){.cta-button download="interrupted-time-series-mechanics-lab.ipynb"}

- Work through the data system before fitting a model.
- Treat the repair-service records as a deterministic teaching simulation, not evidence about a real service.
- Keep the announcement and operational dates fixed.
- Hand back the validation table, stock-flow reconciliation, model comparison, horizon effects, and design judgment.

::: {.callout-note}
The code is shown but not executed when the site is rendered. The downloadable notebook runs offline and recreates every value from the fixed seed below.
:::

## Training Goal

By the end of the lab, you should be able to:

1. validate a regular monthly time series;
2. distinguish stocks, flows, balancing residuals, counts, and exposure-adjusted rates;
3. document and test an anomalous source value;
4. encode announcement, rollout, transition, and post-rollout time correctly;
5. compare segmented OLS with AR(1) generalized least squares; and
6. estimate immediate and horizon-specific effects with valid covariance calculations.

## The Synthetic Service

A public housing-repairs service announces an operational improvement programme in January 2021 and begins rollout in April 2021. The first three rollout months are transitional. At rollout, responsibility for 620 open cases also transfers outside the reporting boundary.

The extract contains:

- the stock of open cases at month end;
- flows of new requests and closures;
- closure categories;
- the number of eligible properties;
- one deliberately corrupted stock value; and
- an incomplete extract used only to demonstrate a missing-period audit.

## Step 1: Load Packages And Generate The Data

```{r}
required_packages <- c("ggplot2", "nlme")

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

simulate_repairs <- function(seed = 48127L) {
  set.seed(seed)
  n_months <- 132L
  announcement_time <- 85L
  implementation_time <- 88L
  time <- seq_len(n_months)
  date <- seq(as.Date("2014-01-01"), by = "month", length.out = n_months)
  announcement <- as.integer(time == announcement_time)
  implemented <- as.integer(time >= implementation_time)
  time_after <- pmax(0L, time - implementation_time)
  transition <- as.integer(time %in% implementation_time:(implementation_time + 2L))
  season_sin <- sin(2 * pi * time / 12)
  season_cos <- cos(2 * pi * time / 12)
  demand_pressure <- 100 + 0.10 * time + 5 * season_sin +
    as.numeric(arima.sim(list(ar = 0.45), n = n_months, sd = 1.3))
  service_noise <- as.numeric(arima.sim(list(ar = 0.60), n = n_months, sd = 10))
  programme_effect <- 70 * implemented + 1.1 * time_after - 25 * transition

  completed <- pmax(1L, round(
    735 + 0.30 * time + 20 * season_cos + 0.9 * (demand_pressure - 100) +
      programme_effect + service_noise
  ))
  withdrawn <- pmax(1L, round(72 + 5 * season_sin + rnorm(n_months, 0, 3)))
  transferred <- pmax(1L, round(28 + rnorm(n_months, 0, 2)))
  closures <- completed + withdrawn + transferred
  new_requests <- pmax(1L, round(
    875 + 0.55 * time + 17 * season_sin + 1.2 * (demand_pressure - 100) +
      rnorm(n_months, 0, 8)
  ))
  scope_change <- ifelse(time == implementation_time, -620L, 0L)

  open_cases <- integer(n_months)
  open_cases[[1]] <- 4400L
  for (index in 2:n_months) {
    open_cases[[index]] <- open_cases[[index - 1L]] +
      new_requests[[index]] - closures[[index]] + scope_change[[index]]
  }

  eligible_properties <- round(24500 + 9 * time - 1100 * implemented)
  published_open_cases <- open_cases
  published_open_cases[[50]] <- round(open_cases[[50]] * 0.38)

  data.frame(
    time,
    date,
    announcement,
    implemented,
    time_after,
    transition,
    season_sin,
    season_cos,
    demand_pressure,
    completed,
    withdrawn,
    transferred,
    closures,
    new_requests,
    scope_change,
    open_cases,
    published_open_cases,
    eligible_properties
  )
}

source_data <- simulate_repairs()
incomplete_extract <- source_data[-40L, ]

stopifnot(
  nrow(source_data) == 132L,
  identical(source_data$date, seq(
    as.Date("2014-01-01"),
    by = "month",
    length.out = 132L
  )),
  !anyDuplicated(source_data$date),
  all(source_data$time_after[source_data$implemented == 1L][1] == 0L)
)
```

## Step 2: Audit Dates, Categories, And Schema

```{r}
expected_dates <- seq(
  min(incomplete_extract$date),
  max(incomplete_extract$date),
  by = "month"
)

missing_dates <- setdiff(expected_dates, incomplete_extract$date)

validation_table <- data.frame(
  Check = c(
    "Unique dates",
    "Complete source calendar",
    "Incomplete extract detects one gap",
    "Closure categories reconcile",
    "No missing core measures"
  ),
  Passed = c(
    !anyDuplicated(source_data$date),
    identical(source_data$date, seq(min(source_data$date), max(source_data$date), by = "month")),
    length(missing_dates) == 1L,
    all(source_data$closures == source_data$completed + source_data$withdrawn + source_data$transferred),
    !anyNA(source_data[c("published_open_cases", "new_requests", "closures")])
  )
)

validation_table
missing_dates

stopifnot(all(validation_table$Passed))
```

The incomplete extract is never used for estimation. It exists to show that checking the number of rows is weaker than checking the exact expected calendar.

## Step 3: Reconcile The Stock And Flows

For every month after the first:

$$
\text{open}_t = \text{open}_{t-1} + \text{requests}_t - \text{closures}_t + \text{scope}_t.
$$

```{r}
source_data$lag_published_open <- c(NA, head(source_data$published_open_cases, -1L))
source_data$balancing_inflow_without_scope <-
  source_data$published_open_cases - source_data$lag_published_open + source_data$closures
source_data$balancing_inflow_with_scope <-
  source_data$balancing_inflow_without_scope - source_data$scope_change
source_data$reconciliation_gap <-
  source_data$balancing_inflow_with_scope - source_data$new_requests

largest_gaps <- source_data[order(abs(source_data$reconciliation_gap), decreasing = TRUE), c(
  "date",
  "published_open_cases",
  "new_requests",
  "scope_change",
  "balancing_inflow_without_scope",
  "balancing_inflow_with_scope",
  "reconciliation_gap"
)]

head(largest_gaps, 5)
```

Checkpoint:

- Why does omitting `scope_change` make the rollout month look like an unusual inflow?
- Why does the corrupted stock create gaps in two adjacent months?
- When observed inflows are unavailable, which unobserved processes enter the balancing residual?

## Step 4: Document And Repair The Source Anomaly

The published stock at February 2018 collapses for one month and rebounds without matching flows. The primary treatment uses centred interpolation; omission and retention are reserved for sensitivity analysis.

```{r}
anomaly_date <- as.Date("2018-02-01")
anomaly_index <- which(source_data$date == anomaly_date)

interpolated_value <- round(mean(source_data$published_open_cases[c(
  anomaly_index - 1L,
  anomaly_index + 1L
)]))

imputation_log <- data.frame(
  date = anomaly_date,
  variable = "published_open_cases",
  original_value = source_data$published_open_cases[anomaly_index],
  replacement_value = interpolated_value,
  rule = "Centred interpolation from adjacent months",
  reason = "One-month collapse is incompatible with recorded flows"
)

analysis_data <- source_data
analysis_data$analysis_open_cases <- analysis_data$published_open_cases
analysis_data$analysis_open_cases[anomaly_index] <- interpolated_value

imputation_log

stopifnot(
  nrow(imputation_log) == 1L,
  imputation_log$replacement_value > imputation_log$original_value,
  analysis_data$analysis_open_cases[anomaly_index] == interpolated_value
)
```

## Step 5: Construct Counts, Exposure, And Rates

Closures are the event count. The cases open at the end of the previous month are the population exposed to closure during the current month.

```{r}
analysis_data$lag_open_cases <- c(NA, head(analysis_data$analysis_open_cases, -1L))
analysis_data$closure_rate_per_1000 <-
  1000 * analysis_data$closures / analysis_data$lag_open_cases
analysis_data$completed_share <- analysis_data$completed / analysis_data$closures
analysis_data$closures_per_1000_properties <-
  1000 * analysis_data$closures / analysis_data$eligible_properties

measure_summary <- data.frame(
  Measure = c(
    "Monthly closures",
    "Closures per 1,000 open cases",
    "Closures per 1,000 eligible properties",
    "Completed share of closures"
  ),
  Mean = c(
    mean(analysis_data$closures),
    mean(analysis_data$closure_rate_per_1000, na.rm = TRUE),
    mean(analysis_data$closures_per_1000_properties),
    mean(analysis_data$completed_share)
  )
)

measure_summary

stopifnot(
  all(is.finite(analysis_data$closure_rate_per_1000[-1L])),
  all(analysis_data$completed_share > 0 & analysis_data$completed_share < 1)
)
```

Write one sentence for the question answered by each measure. Do not use "programme effect" until you have defined the counterfactual.

## Step 6: Plot The Series And Event Timeline

```{r, fig.height=6, fig.width=10}
announcement_date <- analysis_data$date[analysis_data$announcement == 1L]
implementation_date <- analysis_data$date[min(which(analysis_data$implemented == 1L))]

ggplot(analysis_data, aes(date, closure_rate_per_1000)) +
  geom_line(colour = "#24527a", linewidth = 0.7, na.rm = TRUE) +
  geom_point(colour = "#24527a", size = 1, na.rm = TRUE) +
  geom_vline(xintercept = announcement_date, linetype = "dotted", colour = "#c59b3d") +
  geom_vline(xintercept = implementation_date, linetype = "dashed", colour = "#c05a2a") +
  annotate(
    "rect",
    xmin = implementation_date,
    xmax = analysis_data$date[min(which(analysis_data$implemented == 1L)) + 2L],
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.08,
    fill = "#c05a2a"
  ) +
  labs(x = NULL, y = "Closures per 1,000 open cases") +
  theme_minimal(base_size = 12)
```

Checkpoint:

- Is the annual cycle plausible?
- Is a single linear pre-rollout trend adequate as a transparent benchmark?
- What different mechanisms could operate at announcement and rollout?

## Step 7: Fit The Transparent Segmented Model

```{r}
model_data <- subset(analysis_data, is.finite(closure_rate_per_1000))

its_formula <- closure_rate_per_1000 ~
  time + announcement + implemented + time_after + season_sin + season_cos

ols_fit <- lm(its_formula, data = model_data)
ols_terms <- coef(summary(ols_fit))[c("announcement", "implemented", "time_after"), ]
ols_terms

stopifnot(
  coef(ols_fit)[["implemented"]] > 0,
  coef(ols_fit)[["time_after"]] > 0,
  all(is.finite(ols_terms))
)
```

The rollout term estimates an immediate rate change. The post-rollout trend term changes the slope. Neither coefficient by itself is "the effect" at every later month.

## Step 8: Expose The Post-Time Coding Trap

```{r}
model_data$wrong_time_after <- model_data$time * model_data$implemented

wrong_fit <- lm(
  closure_rate_per_1000 ~
    time + announcement + implemented + wrong_time_after + season_sin + season_cos,
  data = model_data
)

parameterization_comparison <- data.frame(
  Specification = c("Elapsed post-time", "Calendar time × implemented"),
  Implementation = c(coef(ols_fit)[["implemented"]], coef(wrong_fit)[["implemented"]]),
  Trend_change = c(coef(ols_fit)[["time_after"]], coef(wrong_fit)[["wrong_time_after"]]),
  check.names = FALSE
)

maximum_fitted_difference <- max(abs(fitted(ols_fit) - fitted(wrong_fit)))
parameterization_comparison
maximum_fitted_difference

stopifnot(
  maximum_fitted_difference < 1e-8,
  abs(coef(ols_fit)[["implemented"]] - coef(wrong_fit)[["implemented"]]) > 1
)
```

Identical fitted values do not make the second implementation coefficient meaningful [@xiao2021parameterization].

## Step 9: Diagnose Dependence And Fit AR(1) Errors

```{r, fig.height=4.8, fig.width=9}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
acf(residuals(ols_fit), main = "OLS residual ACF", xlab = "Lag (months)")
pacf(residuals(ols_fit), main = "OLS residual PACF", xlab = "Lag (months)")
par(old_par)

gls_fit <- gls(
  its_formula,
  data = model_data,
  correlation = corAR1(form = ~ time),
  method = "REML"
)

gls_terms <- summary(gls_fit)$tTable[c("announcement", "implemented", "time_after"), ]

model_comparison <- data.frame(
  Model = rep(c("OLS", "AR(1) GLS"), each = 2),
  Term = rep(c("Immediate rollout", "Monthly trend change"), 2),
  Estimate = c(
    coef(ols_fit)[c("implemented", "time_after")],
    coef(gls_fit)[c("implemented", "time_after")]
  ),
  Standard_error = c(
    coef(summary(ols_fit))[c("implemented", "time_after"), "Std. Error"],
    gls_terms[c("implemented", "time_after"), "Std.Error"]
  )
)

model_comparison

stopifnot(
  all(is.finite(model_comparison$Estimate)),
  all(model_comparison$Standard_error > 0)
)
```

State what the AR(1) model changes. Then name one concurrent event it cannot address.

## Step 10: Estimate Effects At Policy-Relevant Horizons

```{r}
effect_at_horizon <- function(model, horizon) {
  estimates <- coef(model)
  covariance <- vcov(model)
  weights <- c(implemented = 1, time_after = horizon)
  relevant_covariance <- covariance[names(weights), names(weights), drop = FALSE]
  estimate <- sum(weights * estimates[names(weights)])
  standard_error <- sqrt(as.numeric(t(weights) %*% relevant_covariance %*% weights))

  data.frame(
    Horizon = horizon,
    Estimate = estimate,
    Standard_error = standard_error,
    Lower_95 = estimate - qnorm(0.975) * standard_error,
    Upper_95 = estimate + qnorm(0.975) * standard_error
  )
}

horizon_effects <- do.call(rbind, lapply(c(0, 6, 12), function(horizon) {
  effect_at_horizon(gls_fit, horizon)
}))

horizon_effects$Horizon <- c("Immediate", "6 months", "12 months")
horizon_effects

stopifnot(
  all(is.finite(as.matrix(horizon_effects[-1L]))),
  all(horizon_effects$Estimate > 0)
)
```

Explain why the confidence interval cannot be obtained by adding the separate standard errors of the rollout and trend-change coefficients.

## Step 11: Plot Fitted And No-Programme Trajectories

```{r, fig.height=5.5, fig.width=10}
counterfactual_data <- transform(
  model_data,
  announcement = 0L,
  implemented = 0L,
  time_after = 0L
)

model_data$fitted_programme <- as.numeric(predict(gls_fit, newdata = model_data))
model_data$no_programme <- as.numeric(predict(gls_fit, newdata = counterfactual_data))

trajectory_data <- rbind(
  data.frame(date = model_data$date, series = "Fitted programme path", value = model_data$fitted_programme),
  data.frame(date = model_data$date, series = "Estimated no-programme path", value = model_data$no_programme)
)

ggplot(model_data, aes(date, closure_rate_per_1000)) +
  geom_point(size = 1, alpha = 0.55, colour = "#4b615b") +
  geom_line(
    data = trajectory_data,
    aes(y = value, colour = series, linetype = series),
    linewidth = 0.9
  ) +
  geom_vline(xintercept = implementation_date, linetype = "dashed", colour = "#c05a2a") +
  scale_colour_manual(values = c(
    "Fitted programme path" = "#24527a",
    "Estimated no-programme path" = "#bf6b21"
  )) +
  labs(x = NULL, y = "Closures per 1,000 open cases", colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The graph makes the counterfactual inspectable. It does not rule out the same-month boundary transfer or another operational change.

## Final Design Judgment

Write seven sentences:

1. Define the stock, flows, and primary exposure-adjusted outcome.
2. Explain what the balancing residual contains when scope is omitted.
3. State how the source anomaly was treated and which sensitivity is still needed.
4. State the announcement, rollout, and transition periods.
5. Report the immediate and six-month rate effects.
6. Explain what the AR(1) error model changed.
7. Name the strongest causal conclusion the model alone can support.

## Next Step

Continue to the [ITS Design and Diagnostics Lab](interrupted-time-series-design-diagnostics-lab.qmd) to test shared shocks, phased rollout, boundary changes, measurement breaks, and placebo dates.
