---
title: "Lab: Interrupted Time Series Design and Diagnostics"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this as the second interrupted time-series lab, after the data-and-mechanics exercise.

[Download the R Jupyter notebook](notebooks/interrupted-time-series-design-diagnostics-lab.ipynb){.cta-button download="interrupted-time-series-design-diagnostics-lab.ipynb"}

- Treat each scenario as a different identification problem.
- Keep all event dates and sensitivity checks fixed before fitting.
- Compare uncontrolled and controlled estimates before interpreting a break.
- Finish with a one-page design memo rather than a preferred p-value.

::: {.callout-note}
The notebook runs offline with deterministic synthetic repair-service data. Every scenario changes one named feature of the same underlying service system.
:::

## Training Goal

Learn which design and diagnostic tools address which threats:

1. shared shocks and comparison series;
2. phased implementation and transition periods;
3. reporting-boundary and treated-only measurement changes;
4. omitted seasonality and serial dependence;
5. announcement-versus-rollout timing;
6. anomaly, baseline, scope, and transition sensitivities; and
7. false pre-programme interruption dates.

## The Five Scenarios

| Scenario | Added feature | Teaching purpose |
|---|---|---|
| Valid rollout | Only the programme area changes | Reference result |
| Shared shock | Both areas rise at rollout | What a credible control can repair |
| Phased rollout | The effect builds over six months | Why delivery shape matters |
| Boundary change | The programme-area rate shifts with reporting scope | Why constant-scope data matter |
| Measurement break | Programme-area recording rises at rollout | What a control cannot identify |

## Step 1: Pre-Specify The Design

Before running code, record:

1. **Announcement:** January 2021.
2. **Operational rollout:** April 2021.
3. **Transition:** April through June 2021.
4. **Immediate estimand:** programme-area change at operational rollout relative to its no-programme path.
5. **Six-month estimand:** programme-versus-no-programme difference six months after rollout.
6. **Primary error model:** annual Fourier terms and AR(1) errors.
7. **Primary comparison:** the differential change against an unexposed area.
8. **Sensitivity set:** alternative date, shorter baseline, anomaly omission/interpolation, known scope adjustment, transition exclusion, and false dates.

Do not revise this list after viewing results.

## Step 2: Generate The Scenario Panel

```{r}
required_packages <- c("ggplot2", "nlme")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}
invisible(lapply(required_packages, library, character.only = TRUE))

set.seed(73519)
n_months <- 132L
announcement_time <- 85L
implementation_time <- 88L

time_data <- data.frame(
  time = seq_len(n_months),
  date = seq(as.Date("2014-01-01"), by = "month", length.out = n_months)
)
time_data$announcement <- as.integer(time_data$time == announcement_time)
time_data$implemented <- as.integer(time_data$time >= implementation_time)
time_data$time_after <- pmax(0L, time_data$time - implementation_time)
time_data$transition <- as.integer(time_data$time %in% implementation_time:(implementation_time + 2L))
time_data$season_sin <- sin(2 * pi * time_data$time / 12)
time_data$season_cos <- cos(2 * pi * time_data$time / 12)

simulate_ar1 <- function(seed, rho = 0.55, sd = 4) {
  set.seed(seed)
  as.numeric(arima.sim(list(ar = rho), n = n_months, sd = sd))
}

comparison_base <- 178 + 0.12 * time_data$time +
  9 * time_data$season_cos + 3 * time_data$season_sin + simulate_ar1(73520)
treated_base <- 185 + 0.15 * time_data$time +
  9 * time_data$season_cos + 3 * time_data$season_sin + simulate_ar1(73521)

programme_effect <- 24 * time_data$implemented +
  0.32 * time_data$time_after - 9 * time_data$transition
shared_shock <- 18 * time_data$implemented
phased_effect <- pmin(1, pmax(0, time_data$time - implementation_time + 1) / 6) * 26 +
  0.22 * time_data$time_after
boundary_shift <- 12 * time_data$implemented
measurement_shift <- 20 * time_data$implemented

make_scenario <- function(name, treated_outcome, comparison_outcome) {
  result <- rbind(
    transform(time_data, scenario = name, series = "Comparison area", outcome = comparison_outcome),
    transform(time_data, scenario = name, series = "Programme area", outcome = treated_outcome)
  )
  result$series <- relevel(factor(result$series), ref = "Comparison area")
  result[order(result$series, result$time), ]
}

scenario_data <- list(
  "Valid rollout" = make_scenario("Valid rollout", treated_base + programme_effect, comparison_base),
  "Shared shock" = make_scenario("Shared shock", treated_base + programme_effect + shared_shock, comparison_base + shared_shock),
  "Phased rollout" = make_scenario("Phased rollout", treated_base + phased_effect, comparison_base),
  "Boundary change" = make_scenario("Boundary change", treated_base + programme_effect + boundary_shift, comparison_base),
  "Measurement break" = make_scenario("Measurement break", treated_base + programme_effect + measurement_shift, comparison_base)
)

diagnostic_data <- do.call(rbind, scenario_data)
row.names(diagnostic_data) <- NULL

stopifnot(
  nrow(diagnostic_data) == 5L * 2L * n_months,
  all(diagnostic_data$time_after[diagnostic_data$time == implementation_time] == 0L),
  all(is.finite(diagnostic_data$outcome))
)
```

## Step 3: Plot Every Scenario Before Modelling

```{r, fig.height=9, fig.width=11}
ggplot(diagnostic_data, aes(date, outcome, colour = series)) +
  geom_line(linewidth = 0.6) +
  geom_vline(
    xintercept = time_data$date[implementation_time],
    linetype = "dashed",
    colour = "#c05a2a"
  ) +
  facet_wrap(~ scenario, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("Comparison area" = "#bf6b21", "Programme area" = "#24527a")) +
  labs(x = NULL, y = "Closures per 1,000 open cases", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Checkpoint:

- Which shared change becomes visible only after adding the comparison area?
- Why are the boundary and measurement scenarios visually compatible with a real effect?
- Which graph suggests that a sharp step is the wrong impact shape?

## Step 4: Fit Uncontrolled And Controlled ITS

```{r}
fit_uncontrolled <- function(data) {
  treated <- subset(data, series == "Programme area")
  gls(
    outcome ~ time + announcement + implemented + time_after + season_sin + season_cos,
    data = treated,
    correlation = corAR1(form = ~ time),
    method = "ML"
  )
}

fit_controlled <- function(data) {
  data <- data[order(data$series, data$time), ]
  gls(
    outcome ~ series * (time + implemented + time_after) + announcement + season_sin + season_cos,
    data = data,
    correlation = corAR1(form = ~ time | series),
    method = "ML"
  )
}

uncontrolled_fits <- lapply(scenario_data, fit_uncontrolled)
controlled_fits <- lapply(scenario_data, fit_controlled)

extract_results <- function(name) {
  uncontrolled <- uncontrolled_fits[[name]]
  controlled <- controlled_fits[[name]]
  data.frame(
    Scenario = name,
    Model = c("Uncontrolled", "Controlled differential"),
    Immediate = c(
      coef(uncontrolled)[["implemented"]],
      coef(controlled)[["seriesProgramme area:implemented"]]
    ),
    Trend_change = c(
      coef(uncontrolled)[["time_after"]],
      coef(controlled)[["seriesProgramme area:time_after"]]
    ),
    check.names = FALSE
  )
}

scenario_results <- do.call(rbind, lapply(names(scenario_data), extract_results))
row.names(scenario_results) <- NULL
scenario_results

shared_uncontrolled <- subset(scenario_results, Scenario == "Shared shock" & Model == "Uncontrolled")
shared_controlled <- subset(scenario_results, Scenario == "Shared shock" & Model == "Controlled differential")

stopifnot(
  shared_uncontrolled$Immediate > shared_controlled$Immediate,
  shared_controlled$Immediate > 5,
  all(is.finite(as.matrix(scenario_results[c("Immediate", "Trend_change")])))
)
```

The control removes the shared shock when it is genuinely common and the comparison is unexposed. It cannot distinguish a programme-area measurement shift from a programme-area service change.

## Step 5: Represent Phased Rollout Directly

```{r}
phased_treated <- subset(scenario_data[["Phased rollout"]], series == "Programme area")
phased_treated$ramp <- pmin(1, pmax(0, phased_treated$time - implementation_time + 1) / 6)

sharp_fit <- gls(
  outcome ~ time + implemented + time_after + season_sin + season_cos,
  data = phased_treated,
  correlation = corAR1(form = ~ time),
  method = "ML"
)

ramp_fit <- gls(
  outcome ~ time + ramp + time_after + season_sin + season_cos,
  data = phased_treated,
  correlation = corAR1(form = ~ time),
  method = "ML"
)

phase_comparison <- data.frame(
  Model = c("Sharp step", "Six-month ramp"),
  AIC = c(AIC(sharp_fit), AIC(ramp_fit)),
  Normalized_residual_SD = c(
    sd(residuals(sharp_fit, type = "normalized")),
    sd(residuals(ramp_fit, type = "normalized"))
  )
)

phase_comparison
stopifnot(phase_comparison$AIC[phase_comparison$Model == "Six-month ramp"] < phase_comparison$AIC[phase_comparison$Model == "Sharp step"])
```

The ramp was specified from the known six-month delivery profile. Do not infer the ramp duration by searching the outcome for the best fit.

## Step 6: Diagnose Seasonality And Serial Dependence

```{r, fig.height=5, fig.width=10}
valid_treated <- subset(scenario_data[["Valid rollout"]], series == "Programme area")

seasonal_fit <- gls(
  outcome ~ time + announcement + implemented + time_after + season_sin + season_cos,
  data = valid_treated,
  correlation = corAR1(form = ~ time),
  method = "ML"
)

no_season_fit <- gls(
  outcome ~ time + announcement + implemented + time_after,
  data = valid_treated,
  correlation = corAR1(form = ~ time),
  method = "ML"
)

ols_valid <- lm(
  outcome ~ time + announcement + implemented + time_after + season_sin + season_cos,
  data = valid_treated
)

lag_12_acf <- function(values) {
  as.numeric(acf(values, plot = FALSE, lag.max = 12)$acf[13])
}

diagnostic_summary <- data.frame(
  Specification = c("No seasonal terms", "Seasonal AR(1)", "Seasonal OLS"),
  Lag_12_ACF = c(
    lag_12_acf(residuals(no_season_fit, type = "normalized")),
    lag_12_acf(residuals(seasonal_fit, type = "normalized")),
    lag_12_acf(residuals(ols_valid))
  ),
  Immediate_SE = c(
    summary(no_season_fit)$tTable["implemented", "Std.Error"],
    summary(seasonal_fit)$tTable["implemented", "Std.Error"],
    coef(summary(ols_valid))["implemented", "Std. Error"]
  )
)

diagnostic_summary

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
acf(residuals(no_season_fit, type = "normalized"), lag.max = 24, main = "No seasonal terms")
acf(residuals(seasonal_fit, type = "normalized"), lag.max = 24, main = "Seasonal AR(1)")
par(old_par)

stopifnot(
  abs(diagnostic_summary$Lag_12_ACF[1]) > abs(diagnostic_summary$Lag_12_ACF[2]),
  all(diagnostic_summary$Immediate_SE > 0)
)
```

Omitted seasonality is a mean-model problem. Serial dependence is an error-model problem. Neither is an attribution strategy.

## Step 7: Run The Pre-Specified Sensitivity Set

```{r}
fit_segmented <- function(data) {
  lm(outcome ~ time + announcement + implemented + time_after + season_sin + season_cos, data = data)
}

announcement_data <- transform(
  valid_treated,
  alternative_post = as.integer(time >= announcement_time),
  alternative_after = pmax(0L, time - announcement_time)
)
announcement_fit <- lm(
  outcome ~ time + alternative_post + alternative_after + season_sin + season_cos,
  data = announcement_data
)

short_baseline_fit <- fit_segmented(subset(valid_treated, time >= 25L))

transition_fit <- fit_segmented(subset(
  valid_treated,
  !(time %in% implementation_time:(implementation_time + 2L))
))

anomaly_data <- valid_treated
anomaly_data$outcome[anomaly_data$time == 50L] <- anomaly_data$outcome[anomaly_data$time == 50L] + 70
anomaly_index <- which(anomaly_data$time == 50L)
interpolated_data <- anomaly_data
interpolated_data$outcome[anomaly_index] <- mean(anomaly_data$outcome[c(anomaly_index - 1L, anomaly_index + 1L)])
omitted_data <- anomaly_data[-anomaly_index, ]

retained_anomaly_fit <- fit_segmented(anomaly_data)
interpolated_anomaly_fit <- fit_segmented(interpolated_data)
omitted_anomaly_fit <- fit_segmented(omitted_data)

boundary_data <- subset(scenario_data[["Boundary change"]], series == "Programme area")
boundary_published_fit <- fit_segmented(boundary_data)
boundary_data$constant_scope_outcome <- boundary_data$outcome - boundary_shift
boundary_constant_scope_fit <- lm(
  constant_scope_outcome ~ time + announcement + implemented + time_after + season_sin + season_cos,
  data = boundary_data
)

sensitivity_results <- data.frame(
  Specification = c(
    "Primary rollout date",
    "Announcement date",
    "Shorter baseline",
    "Exclude transition",
    "Retain anomaly",
    "Interpolate anomaly",
    "Omit anomaly",
    "Published boundary",
    "Constant scope"
  ),
  Immediate = c(
    coef(ols_valid)[["implemented"]],
    coef(announcement_fit)[["alternative_post"]],
    coef(short_baseline_fit)[["implemented"]],
    coef(transition_fit)[["implemented"]],
    coef(retained_anomaly_fit)[["implemented"]],
    coef(interpolated_anomaly_fit)[["implemented"]],
    coef(omitted_anomaly_fit)[["implemented"]],
    coef(boundary_published_fit)[["implemented"]],
    coef(boundary_constant_scope_fit)[["implemented"]]
  )
)

sensitivity_results

stopifnot(
  all(is.finite(sensitivity_results$Immediate)),
  abs(
    sensitivity_results$Immediate[sensitivity_results$Specification == "Published boundary"] -
      sensitivity_results$Immediate[sensitivity_results$Specification == "Constant scope"]
  ) > 5
)
```

Interpret each change against its named concern. Do not select the row with the largest estimate or smallest standard error.

## Step 8: Test False Pre-Programme Dates

```{r}
placebo_dates <- c(49L, 61L, 73L)
pre_programme <- subset(valid_treated, time < announcement_time)

placebo_effect <- function(interruption) {
  placebo <- transform(
    pre_programme,
    placebo_post = as.integer(time >= interruption),
    placebo_after = pmax(0L, time - interruption)
  )
  fit <- lm(
    outcome ~ time + placebo_post + placebo_after + season_sin + season_cos,
    data = placebo
  )
  unname(coef(fit)[["placebo_post"]])
}

placebo_results <- data.frame(
  False_interruption_time = placebo_dates,
  Immediate = vapply(placebo_dates, placebo_effect, numeric(1))
)

observed_immediate <- unname(coef(ols_valid)[["implemented"]])
placebo_results$At_least_as_large_as_observed <-
  abs(placebo_results$Immediate) >= abs(observed_immediate)

placebo_results

stopifnot(
  max(placebo_dates) < announcement_time,
  all(is.finite(placebo_results$Immediate))
)
```

These false dates overlap and rollout was not randomized. Use the results as a stability diagnostic, not a randomisation-inference p-value.

## Step 9: Classify What Each Design Can Support

Fill the last column before comparing your answers.

| Scenario | What the controlled estimate sees | Defensible interpretation |
|---|---|---|
| Valid rollout | Programme-area differential | Consistent with the simulated service change |
| Shared shock | Differential after removing the common rise | Attribution strengthens if the comparison is unexposed |
| Phased rollout | An average sharp break unless the ramp is modelled | Delivery shape must be represented |
| Boundary change | Programme-area effect plus rate-definition change | Requires constant-scope reconstruction |
| Measurement break | Programme plus recording discontinuity | Effect remains unidentified from these series |

## Final Design Memo

Write one page with these headings:

### Design And Estimands

State the areas, outcome, exposure, announcement, rollout, transition, immediate estimand, and six-month estimand.

### Primary Result

Compare uncontrolled and controlled results for the valid and shared-shock scenarios.

### Diagnostics And Sensitivity

Summarize seasonal structure, OLS-versus-AR(1) uncertainty, timing, baseline, anomaly, transition, scope, and placebo evidence.

### Identification Judgment

Explain why the control helps with the shared shock but cannot resolve a programme-area reporting or boundary discontinuity without additional data.

### Defensible Claim

State the strongest conclusion you would defend and the most important remaining limitation.

## Next Step

Continue to the [ITS Counterfactual Validation Lab](interrupted-time-series-counterfactual-validation-lab.qmd) to select a short-horizon untreated model using only pre-rollout forecast performance.
