---
title: "Interrupted Time Series for Causal Inference"
subtitle: "Intervention timing, counterfactual trends, and serial dependence"
author: "Luke Heley"
date: 19 September 2026
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
    keep-tex: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 7
  fig-height: 4
---

## Executive summary

Interrupted time series (ITS) evaluates a change in an outcome's trajectory at a
known intervention date. Its counterfactual comes from a model of how the outcome
would have evolved without intervention. Repeated observations make trend and
seasonality visible, but they do not create an untreated population. A concurrent
shock can produce precisely the same break as the policy being evaluated.

This report develops the design before the regression. It uses monthly British
road-casualty data to connect intervention coding, a segmented log model, serial
dependence, and effects at decision-relevant horizons. The worked example is
reproducible with repository data and is deliberately cautious about attributing
the observed change solely to seatbelt legislation [@lopezbernal2017tutorial;
@harvey1986seatbelts].

## Design and estimand

### Define the intervention and outcome

Write down the announcement, implementation, enforcement, and expected behavioral
response dates. Distinguish an abrupt policy from a rollout. Specify the population,
measurement interval, outcome scale, and the horizon over which an effect matters.
A change in counts, a change in rates, and a change in log counts are different
estimands. The denominator must be chosen on substantive grounds.

The front-seat belt requirement took effect on 31 January 1983. With monthly data,
February is the first treated observation. The outcome here is front-seat
passengers killed or seriously injured, not deaths alone and not a rate per mile.
There are 169 pre-policy and 23 post-policy months. The long baseline helps study
seasonality, while the short post-period limits information about a new trend
[@rseatbeltsdata; @ukparliament1986seatbeltdebate].

### State the missing comparison

For time $t$ after intervention, the causal contrast is
$E[Y_t(1)-Y_t(0)]$. Only $Y_t(1)$ is observed. Identification requires that the
model for the untreated path remains appropriate after implementation. Relevant
threats include concurrent interventions, changing exposure, altered recording,
anticipation, and an unstable pre-policy process.

Unlike DiD, a single-series ITS does not borrow a contemporaneous untreated trend.
Unlike SCM, it does not construct a weighted donor comparison. This makes ITS
useful when no credible untreated population exists, but also places more weight
on the time-series counterfactual. More frequent observations improve temporal
resolution; they do not remove confounding at the intervention date.

## A reproducible Seatbelts analysis

The [data catalogue](data.qmd#seatbelts) records the bundled snapshot. Packages must
already be installed; this report performs no installation or data download.

```{r}
source("data/load-data.R")
library(ggplot2)
library(nlme)
d <- as.data.frame(qed_data("Seatbelts"))
d$date <- seq(as.Date("1969-01-01"), by="month", length.out=nrow(d))
d$index <- seq_len(nrow(d))
first <- which(d$law == 1)[1]
d$time <- d$index - first
d$post_time <- pmax(0, d$time)
d$season_sin <- sin(2*pi*d$index/12)
d$season_cos <- cos(2*pi*d$index/12)
d$log_front <- log(d$front)
stopifnot(nrow(d) == 192, all(d$front > 0),
          d$date[first] == as.Date("1983-02-01"),
          sum(d$law == 0) == 169, sum(d$law == 1) == 23)
knitr::kable(d[(first-2):(first+2),
               c("date", "time", "law", "post_time")])
```

```{r}
#| fig-cap: "Monthly front-seat casualties, with the first treated month marked."
#| fig-alt: "Monthly casualty counts from 1969 to 1984, with February 1983 marked by a vertical line."
ggplot(d, aes(date, front)) +
  geom_line(color="#245ca4", linewidth=0.5) +
  geom_vline(xintercept=as.Date("1983-02-01"), linetype=2) +
  labs(x=NULL, y="Front-seat passengers killed or seriously injured") +
  theme_minimal(base_size=11)
```

Inspect missing periods, outliers, seasonality, and recording breaks before fitting
the intervention model. Changes in driving exposure can change casualties without
changing risk. Contemporaneous kilometres might be a confounder, a mediator, or
part of the effect of interest; including them is not automatically an improvement.

### Encode the intervention

Let $t=0$ identify the first treated month, $D_t$ indicate treatment, and
$P_t=\max(0,t)$. A segmented model is

$$
\log Y_t=\beta_0+\beta_1t+\beta_2D_t+\beta_3P_t
 +\gamma_1\sin(2\pi m_t/12)+\gamma_2\cos(2\pi m_t/12)+\epsilon_t.
$$

The immediate log-level contrast is $\beta_2$; the change in monthly log slope is
$\beta_3$; the post-policy slope is $\beta_1+\beta_3$. Starting post-time at one
instead of zero changes the interpretation of the level coefficient. Check the
actual rows around implementation [@wagner2002segmented; @xiao2021parameterization].

```{r}
spec <- log_front ~ time + law + post_time + season_sin + season_cos
ols <- lm(spec, data=d)
fit <- gls(spec, data=d, correlation=corAR1(form=~index), method="REML")
knitr::kable(summary(fit)$tTable, digits=4)
```

The annual harmonic models a smooth seasonal cycle. AR(1) models residual
dependence that decays geometrically with elapsed months. They solve different
problems. GLS changes estimation and uncertainty, not the causal identification
assumption. A misspecified trend can also appear as autocorrelated noise.

```{r}
#| fig-cap: "Residual dependence before and after the working AR(1) model."
#| fig-alt: "Two residual autocorrelation plots compare OLS with normalized GLS residuals."
par(mfrow=c(1, 2))
acf(residuals(ols), main="OLS residuals")
acf(residuals(fit, type="normalized"), main="Normalized GLS residuals")
par(mfrow=c(1, 1))
```

### Report contrasts at named horizons

At horizon $h$, the contrast is $\delta_h=\beta_2+h\beta_3$. Its variance is
$V_{22}+2hV_{23}+h^2V_{33}$. Omitting the covariance term can materially misstate
uncertainty. The horizons below are within the observed post-policy period.

```{r}
contrast <- function(model, h) {
  b <- coef(model)
  L <- setNames(rep(0, length(b)), names(b))
  L["law"] <- 1
  if ("post_time" %in% names(b)) L["post_time"] <- h
  delta <- sum(L*b)
  se <- sqrt(drop(t(L) %*% vcov(model) %*% L))
  data.frame(month=h, log_effect=delta, percent=100*expm1(delta),
             lower=100*expm1(delta-1.96*se),
             upper=100*expm1(delta+1.96*se))
}
effects <- do.call(rbind, lapply(c(0, 6, 12), function(h) contrast(fit, h)))
knitr::kable(effects, digits=3)
```

These are approximate model-based 95% intervals. The transformation describes a
percentage contrast on the exponentiated expected-log scale, not automatically
a percentage effect on arithmetic mean casualties. An injuries-prevented total
requires an explicit retransformation and aggregation argument.

### Show the counterfactual

```{r}
#| fig-cap: "The jointly fitted intervention path and the same fit with intervention terms switched off."
#| fig-alt: "Observed log casualties and fitted paths with and without the intervention terms."
untreated <- d
untreated$law <- 0
untreated$post_time <- 0
d$fitted <- as.numeric(predict(fit))
d$counterfactual <- as.numeric(predict(fit, newdata=untreated))
ggplot(d, aes(date, log_front)) +
  geom_line(color="grey65", linewidth=0.4) +
  geom_line(aes(y=fitted, color="Segmented fit"), linewidth=0.7) +
  geom_line(aes(y=counterfactual, color="No intervention terms"),
            linewidth=0.7, linetype=2) +
  scale_color_manual(values=c("No intervention terms"="#177b72",
                             "Segmented fit"="#245ca4")) +
  labs(x=NULL, y="Log front-seat casualties", color=NULL) +
  theme_minimal(base_size=11) + theme(legend.position="bottom")
```

This is a model fitted to the whole series, with intervention terms removed for
prediction. It is not a forecast estimated only before the policy. For forecasting,
reserve untreated validation windows, compare models at common rolling origins,
and avoid using treated outcomes to select the untreated model. Good historical
forecast accuracy still cannot rule out a new concurrent cause after implementation.

## Sensitivity and extensions

Compare effect shapes and seasonal specifications because they express competing
scientific assumptions. Use ML, not REML likelihoods, when comparing different
fixed-effect formulas. Keep the outcome, sample, and horizon fixed where possible.

```{r}
d$month <- factor(format(d$date, "%m"))
models <- list(
  "Level only"=gls(log_front ~ time + law + season_sin + season_cos,
    data=d, correlation=corAR1(form=~index), method="ML"),
  "Level and slope"=update(fit, method="ML"),
  "Month indicators"=gls(log_front ~ time + law + post_time + month,
    data=d, correlation=corAR1(form=~index), method="ML"))
sensitivity <- do.call(rbind, lapply(names(models), function(name) {
  cbind(model=name, contrast(models[[name]], 12))
}))
knitr::kable(sensitivity[c("model", "percent", "lower", "upper")], digits=2)
```

Also examine defensible pre-period lengths, implementation lags, exposure-adjusted
outcomes, and influential months. Report changes in the estimate rather than
selecting only the most reassuring specification. A placebo interruption tests
whether similar apparent breaks arise without the policy; searching many dates
and reporting only favorable results defeats that purpose.

A controlled ITS adds a contemporaneous series to help account for common shocks.
The comparison needs a credible untreated relationship with the target outcome
and must not itself be affected by the intervention. Rear-seat passengers and van
drivers are not automatically valid controls for a seatbelt policy. Spillovers,
different exposure trends, and concurrent road-safety measures require substantive
assessment [@lopezbernal2018controls]. Count models, offsets, richer seasonal
structure, and dynamic intervention models can be useful when their estimands
and assumptions fit the question; extra flexibility does not replace a control.

## Interpretation and reporting

The analysis estimates a trajectory change conditional on the specified time
process. Concurrent drink-driving enforcement and other road-safety changes limit
attribution to the belt requirement alone. Do not convert a significant model
contrast into a uniquely identified policy effect, or an imprecise slope change
into proof of no effect.

Report the intervention timeline, outcome and exposure, 169/23 observation split,
effect shape, seasonal and error specifications, horizon contrasts and intervals,
residual checks, sensitivity, and strongest remaining alternative explanation.
Separate sampling/model uncertainty from uncertainty about the counterfactual.
When the latter is decisive, recommend a more credible comparison or a bounded
descriptive conclusion rather than another estimator.

## Teaching route

- [Teaching deck](../slides/its.qmd) and [presenter notes](../slides/its_script.qmd).
- [Practical R guide](../notes/other-methods/how-to-do-interrupted-time-series.qmd) for a shorter procedural walkthrough.
- [Foundations lab](interrupted-time-series-mechanics-lab.qmd) for coding and contrasts.
- [Design and robustness lab](interrupted-time-series-design-diagnostics-lab.qmd) for outcomes, controls, and attribution.
- [Counterfactual validation lab](interrupted-time-series-counterfactual-validation-lab.qmd) for untreated rolling-origin forecasts.

Each lab supplies Quarto, an R Jupyter notebook, and an offline bundle through the
[Labs directory](index.qmd#interrupted-time-series-labs).

## References
