---
title: "Lab: Interrupted Time Series Foundations With Seatbelts"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

This is the first interrupted time-series lab. The guided core is designed for a 45–60 minute workshop; the AR(1) section is an optional extension.

[Download the R Jupyter notebook](notebooks/interrupted-time-series-mechanics-lab.ipynb){.cta-button download="interrupted-time-series-mechanics-lab.ipynb"}

- Work from the intervention date and outcome definition before looking at coefficients.
- Compare a pre/post summary with a model that preserves the trajectory.
- Keep level and trend changes separate.
- Interpret log-scale estimates as percentage changes at named horizons.

::: {.callout-note}
The lab uses `Seatbelts`, which is included with R. It contains monthly road-casualty series for Great Britain from January 1969 through December 1984, so no external data download is required.
:::

## Training Goal

By the end of the core lab, you should be able to:

1. turn a monthly `ts` object into an analysis data frame;
2. code an interruption and post-intervention time correctly;
3. fit and interpret a seasonal segmented regression;
4. estimate effects at policy-relevant horizons; and
5. describe why the estimated break is not automatically causal.

## Step 1: Load And Describe The Data

The primary outcome is `front`: front-seat passengers killed or seriously injured each month. This is a **count**, not a fatality-only measure and not a rate. Compulsory front-seat belt wearing began on 31 January 1983, so the supplied monthly `law` indicator first equals one in February 1983 [@harvey1986seatbelts; @ukparliament2012seatbelts].

```{r}
required_packages <- c("ggplot2", "nlme")
missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}
invisible(lapply(required_packages, library, character.only = TRUE))

data("Seatbelts", package = "datasets")

seatbelts <- as.data.frame(Seatbelts)
seatbelts$time <- seq_len(nrow(seatbelts))
seatbelts$date <- seq(
  as.Date("1969-01-01"),
  by = "month",
  length.out = nrow(seatbelts)
)
seatbelts$post_time <- ifelse(
  seatbelts$law == 1,
  seatbelts$time - min(seatbelts$time[seatbelts$law == 1]),
  0
)
seatbelts$season_sin <- sin(2 * pi * seatbelts$time / 12)
seatbelts$season_cos <- cos(2 * pi * seatbelts$time / 12)

intervention_date <- min(seatbelts$date[seatbelts$law == 1])

data.frame(
  observations = nrow(seatbelts),
  first_month = min(seatbelts$date),
  last_month = max(seatbelts$date),
  first_law_month = intervention_date,
  pre_law_months = sum(seatbelts$law == 0),
  post_law_months = sum(seatbelts$law == 1)
)

stopifnot(
  nrow(seatbelts) == 192L,
  intervention_date == as.Date("1983-02-01"),
  seatbelts$post_time[seatbelts$date == intervention_date] == 0,
  all(diff(seatbelts$date) > 0),
  all(seatbelts$front > 0)
)
```

Checkpoint: why is February, rather than January, the first treated monthly observation?

## Step 2: Plot The Series Before Modelling

```{r, fig.height=4.8, fig.width=10}
ggplot(seatbelts, aes(date, front)) +
  geom_line(linewidth = 0.6, colour = "#24527a") +
  geom_vline(
    xintercept = intervention_date,
    linetype = "dashed",
    colour = "#a23b3b"
  ) +
  annotate(
    "text",
    x = intervention_date,
    y = max(seatbelts$front),
    label = "Law in effect",
    hjust = -0.08,
    vjust = 1.2,
    colour = "#a23b3b"
  ) +
  labs(
    x = NULL,
    y = "Front-seat passengers killed or seriously injured",
    title = "Great Britain road casualties before and after the seatbelt law"
  ) +
  theme_minimal(base_size = 12)
```

Before fitting a model, record what the graph suggests about:

- long-run trend;
- annual seasonality;
- the apparent change around February 1983; and
- the short, 23-month post-law period.

## Step 3: Show What A Pre/Post Comparison Discards

```{r}
prepost_summary <- aggregate(
  front ~ law,
  data = seatbelts,
  FUN = function(x) c(mean = mean(x), sd = sd(x), months = length(x))
)

prepost_table <- data.frame(
  period = c("Before law", "Law in effect"),
  mean_front_casualties = round(prepost_summary$front[, "mean"], 1),
  sd_front_casualties = round(prepost_summary$front[, "sd"], 1),
  months = prepost_summary$front[, "months"]
)

prepost_table
```

The averages mix different calendar months and ignore the pre-law trend. They also compare 169 pre-law observations with only 23 post-law observations. This is a useful description, but it is not the ITS counterfactual.

## Step 4: Fit A Transparent Segmented Regression

The model uses a log outcome so intervention coefficients can be converted into proportional changes. A sine/cosine pair represents annual seasonality without estimating eleven separate month coefficients.

```{r}
its_formula <- log(front) ~
  time + law + post_time + season_sin + season_cos

ols_fit <- lm(its_formula, data = seatbelts)

ols_terms <- coef(summary(ols_fit))[c("law", "post_time"), ]
ols_results <- data.frame(
  term = c("Immediate log change", "Monthly log-trend change"),
  estimate = ols_terms[, "Estimate"],
  standard_error = ols_terms[, "Std. Error"],
  p_value = ols_terms[, "Pr(>|t|)"],
  row.names = NULL
)

ols_results

immediate_percent <- 100 * (exp(coef(ols_fit)[["law"]]) - 1)
immediate_percent

stopifnot(
  all(is.finite(coef(ols_fit))),
  immediate_percent > -100,
  length(residuals(ols_fit)) == nrow(seatbelts)
)
```

Interpret the two intervention terms separately:

- `law` is the estimated immediate change in February 1983 relative to the projected no-law path;
- `post_time` is the change in monthly trend after that first law month.

Neither coefficient alone describes every post-law month.

## Step 5: Estimate Effects At Named Horizons

At horizon $h$, the log-scale contrast is

$$
\widehat{\delta}(h) = \widehat{\beta}_{law} + h\widehat{\beta}_{post}.
$$

Its standard error must include the covariance between the two coefficients.

```{r}
effect_at_horizon <- function(model, horizon) {
  weights <- c(law = 1, post_time = horizon)
  coefficients <- coef(model)[names(weights)]
  covariance <- vcov(model)[names(weights), names(weights)]
  log_effect <- sum(weights * coefficients)
  standard_error <- sqrt(drop(t(weights) %*% covariance %*% weights))

  data.frame(
    horizon_months = horizon,
    percent_change = 100 * (exp(log_effect) - 1),
    lower_95 = 100 * (exp(log_effect - 1.96 * standard_error) - 1),
    upper_95 = 100 * (exp(log_effect + 1.96 * standard_error) - 1)
  )
}

horizon_results <- do.call(
  rbind,
  lapply(c(0, 6, 12), function(h) effect_at_horizon(ols_fit, h))
)

round(horizon_results, 1)

stopifnot(
  nrow(horizon_results) == 3L,
  all(is.finite(as.matrix(horizon_results)))
)
```

Checkpoint: why should the 12-month result usually be less certain than the immediate result?

## Step 6: Plot The Estimated No-Law Path

```{r, fig.height=5.2, fig.width=10}
seatbelts$fitted_observed <- exp(predict(ols_fit, newdata = seatbelts))

no_law_data <- seatbelts
no_law_data$law <- 0
no_law_data$post_time <- 0
seatbelts$fitted_no_law <- exp(predict(ols_fit, newdata = no_law_data))

ggplot(seatbelts, aes(date)) +
  geom_line(aes(y = front, colour = "Observed"), linewidth = 0.55) +
  geom_line(aes(y = fitted_observed, colour = "Fitted with law"), linewidth = 0.75) +
  geom_line(
    aes(y = fitted_no_law, colour = "Projected no-law path"),
    linewidth = 0.75,
    linetype = "dashed"
  ) +
  geom_vline(xintercept = intervention_date, linetype = "dotted") +
  scale_colour_manual(values = c(
    "Observed" = "#303030",
    "Fitted with law" = "#24527a",
    "Projected no-law path" = "#bf6b21"
  )) +
  labs(
    x = NULL,
    y = "Front-seat passengers killed or seriously injured",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The no-law line is a model projection. The gap is estimated rather than observed, and a coincident event could generate the same pattern.

## Optional Extension: Account For Serial Dependence

The core OLS model makes the parameterization transparent. Monthly residuals may still be correlated, so compare it with generalized least squares using AR(1) errors.

```{r, fig.height=4.2, fig.width=9}
gls_fit <- gls(
  its_formula,
  data = seatbelts,
  correlation = corAR1(form = ~ time),
  method = "REML"
)

gls_terms <- summary(gls_fit)$tTable[c("law", "post_time"), ]

model_comparison <- data.frame(
  model = rep(c("OLS", "GLS with AR(1) errors"), each = 2),
  term = rep(c("Immediate change", "Trend change"), times = 2),
  estimate = c(ols_terms[, "Estimate"], gls_terms[, "Value"]),
  standard_error = c(ols_terms[, "Std. Error"], gls_terms[, "Std.Error"])
)

model_comparison

par(mfrow = c(1, 2))
acf(residuals(ols_fit), main = "OLS residuals")
acf(residuals(gls_fit, type = "normalized"), main = "Normalized GLS residuals")
par(mfrow = c(1, 1))

stopifnot(
  all(is.finite(gls_terms)),
  abs(coef(gls_fit$modelStruct$corStruct, unconstrained = FALSE)) < 1
)
```

AR(1) errors address residual dependence. They do not remove a concurrent policy, repair a changing outcome definition, or make a weak comparison series credible.

## Final Design Judgment

Write three sentences:

1. Define the outcome, intervention month, and post-law observation window.
2. Describe the estimated immediate and 12-month changes without relying on a p-value alone.
3. State the most important reason the estimated break may not equal the causal effect of compulsory seatbelt wearing.

## Next Step

Continue to [Seatbelt Law Design and Robustness](interrupted-time-series-design-diagnostics-lab.qmd) to compare outcomes, candidate controls, and a compact prespecified sensitivity set.
