---
title: "Lab: LaLonde Matching"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as the guided lab version of the main matching overview.

- Keep the [Matching and Weighting](../notes/matching/matching-and-weighting.qmd) note open in another tab.
- Run each step in order rather than skipping to the estimates.
- Record balance, retention, and weighting cost as you go.
- End with a design judgement, not just a preferred coefficient.

::: {.callout-note}
The code is shown but not executed when the site is rendered. That keeps the page readable while preserving the main workflow for live teaching and self-study.
:::

## Training Goal

Turn the `lalonde` comparison into a repeatable lab exercise:

1. define the `ATT` before any adjustment choice
2. inspect the raw treated-versus-control comparison
3. compare exact matching, `CEM`, and entropy balancing on the same covariate set
4. judge each method on balance, overlap, retention, and weight concentration
5. decide which design you would defend for programme participants

## Dataset At A Glance

This lab uses `MatchIt::lalonde`, the same dataset used in the main matching overview.

- Treated group: National Supported Work programme participants
- Comparison group: observational controls
- Outcome: `re78`
- Design covariates: `age`, `educ`, `race`, `married`, `nodegree`, `re74`, `re75`
- Target estimand: `ATT`

This is the cleanest place to practice the repo's design-first workflow because the data are already familiar from the notes and the tradeoffs are visible quickly.

## What To Hand Back

By the end of the lab, you should be able to report:

- the worst raw imbalance before adjustment
- the strongest exact-matching result you can defend
- the `CEM` retention-versus-balance tradeoff
- the entropy-balancing control `ESS` and maximum control weight
- the design you would defend, plus one limitation you would still state clearly

## Step 1: Load Packages And Prepare The Data

```{r}
required_packages <- c(
  "MatchIt",
  "WeightIt",
  "cobalt",
  "dplyr",
  "ggplot2"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

data("lalonde", package = "MatchIt")

dat <- lalonde |>
  mutate(
    treat = as.integer(treat),
    outcome = re78,
    married = factor(married, levels = c(0, 1), labels = c("not_married", "married")),
    nodegree = factor(nodegree, levels = c(0, 1), labels = c("has_degree", "no_degree"))
  )

design_covariates <- c(
  "age", "educ", "race", "married", "nodegree", "re74", "re75"
)

dat |>
  select(treat, outcome, all_of(design_covariates)) |>
  glimpse()
```

Checkpoint:

- Is the estimand clear before any method is chosen?
- Are all design covariates measured before the outcome?

## Step 2: Describe The Raw Comparison

```{r}
pre_match_summary <- dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    mean_outcome = mean(outcome, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE),
    mean_educ = mean(educ, na.rm = TRUE),
    prop_married = mean(married == "married", na.rm = TRUE),
    prop_no_degree = mean(nodegree == "no_degree", na.rm = TRUE),
    mean_re74 = mean(re74, na.rm = TRUE),
    mean_re75 = mean(re75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "Treated", "Control")) |>
  select(
    group, n, mean_outcome, mean_age, mean_educ,
    prop_married, prop_no_degree, mean_re74, mean_re75
  ) |>
  mutate(across(-c(group, n), ~ round(.x, 3)))

pre_match_summary
```

What to look for:

- lower prior earnings in the treated group
- lower marriage rates in the treated group
- whether the untreated group looks remotely usable without design work

## Step 3: Build The Baseline Balance Benchmark

```{r}
m_out0 <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = dat,
  method = NULL,
  estimand = "ATT"
)

summary(m_out0, un = TRUE)

unadjusted_balance <- bal.tab(
  m_out0,
  un = TRUE,
  binary = "std",
  disp.v.ratio = TRUE,
  m.threshold = 0.1
)

unadjusted_balance
```

```{r fig.height=5.5}
love.plot(
  m_out0,
  stats = "mean.diffs",
  abs = TRUE,
  binary = "std",
  thresholds = c(m = 0.1),
  var.order = "unadjusted"
)
```

Checkpoint:

- Which covariates are most imbalanced before adjustment?
- Does the raw sample fail on one variable or on most of the design?

## Step 4: Exact Matching On A Reduced Discrete Set

```{r}
m_exact <- matchit(
  treat ~ race + married + nodegree + educ,
  data = dat,
  method = "exact",
  estimand = "ATT"
)

matched_exact <- match.data(m_exact)

exact_retention <- tibble(
  metric = c(
    "Treated units in full sample",
    "Treated units retained",
    "Control units in full sample",
    "Control units retained",
    "Matched subclasses"
  ),
  value = c(
    sum(dat$treat == 1),
    sum(matched_exact$treat == 1),
    sum(dat$treat == 0),
    sum(matched_exact$treat == 0),
    dplyr::n_distinct(matched_exact$subclass)
  )
)

exact_retention
```

```{r}
summary(m_exact, un = TRUE)

exact_balance <- bal.tab(
  m_exact,
  data = dat,
  un = TRUE,
  binary = "std",
  disp.v.ratio = TRUE,
  m.threshold = 0.1,
  addl = ~ age + re74 + re75
)

exact_balance
```

```{r}
fit_exact <- lm(outcome ~ treat, data = matched_exact, weights = weights)

exact_effect <- {
  exact_coef <- coef(summary(fit_exact))["treat", ]
  tibble(
    method = "Exact matching",
    estimate = unname(exact_coef["Estimate"]),
    std_error = unname(exact_coef["Std. Error"]),
    p_value = unname(exact_coef["Pr(>|t|)"])
  )
} |>
  mutate(across(c(estimate, std_error, p_value), ~ round(.x, 3)))

exact_effect
```

What to discuss:

- Exact matching is transparent because the rules are visible.
- The price is that important continuous variables still sit outside the exact constraints.
- If `re74` remains above `0.1`, exact matching has not fully solved the design problem.

## Step 5: Compare Two CEM Specifications

```{r}
m_cem_default <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = dat,
  method = "cem",
  estimand = "ATT"
)

matched_cem_default <- match.data(m_cem_default)

cem_cutpoints <- list(
  age = "q5",
  educ = 4,
  re74 = "q4",
  re75 = "q4"
)

m_cem <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = dat,
  method = "cem",
  estimand = "ATT",
  cutpoints = cem_cutpoints
)

matched_cem <- match.data(m_cem)

cem_default_balance <- bal.tab(
  m_cem_default,
  un = TRUE,
  binary = "std",
  disp.v.ratio = TRUE,
  m.threshold = 0.1
)

cem_balance <- bal.tab(
  m_cem,
  un = TRUE,
  binary = "std",
  disp.v.ratio = TRUE,
  m.threshold = 0.1
)

cem_comparison <- tibble(
  specification = c("Default CEM", "Documented custom cutpoints"),
  treated_retained = c(
    sum(matched_cem_default$treat == 1),
    sum(matched_cem$treat == 1)
  ),
  control_retained = c(
    sum(matched_cem_default$treat == 0),
    sum(matched_cem$treat == 0)
  ),
  matched_subclasses = c(
    dplyr::n_distinct(matched_cem_default$subclass),
    dplyr::n_distinct(matched_cem$subclass)
  ),
  max_abs_adjusted_smd = c(
    max(abs(cem_default_balance$Balance$Diff.Adj), na.rm = TRUE),
    max(abs(cem_balance$Balance$Diff.Adj), na.rm = TRUE)
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

cem_comparison
```

```{r}
fit_cem <- lm(outcome ~ treat, data = matched_cem, weights = weights)

cem_effect <- {
  cem_coef <- coef(summary(fit_cem))["treat", ]
  tibble(
    method = "Coarsened exact matching",
    estimate = unname(cem_coef["Estimate"]),
    std_error = unname(cem_coef["Std. Error"]),
    p_value = unname(cem_coef["Pr(>|t|)"])
  )
} |>
  mutate(across(c(estimate, std_error, p_value), ~ round(.x, 3)))

cem_effect
```

Checkpoint:

- Which `CEM` version would you defend?
- How much treated retention are you willing to lose for stronger balance?

## Step 6: Compare Entropy-Balancing Specifications

```{r}
w_ebal_default <- weightit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = dat,
  method = "ebal",
  estimand = "ATT"
)

w_ebal_moments <- weightit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = dat,
  method = "ebal",
  estimand = "ATT",
  moments = c(age = 2, re74 = 2, re75 = 2)
)

ebal_default_summary <- summary(w_ebal_default)
ebal_moments_summary <- summary(w_ebal_moments)

ebal_default_tuning_balance <- bal.tab(
  w_ebal_default,
  un = TRUE,
  binary = "std",
  addl = ~ I(age^2) + I(re74^2) + I(re75^2)
)

ebal_moments_tuning_balance <- bal.tab(
  w_ebal_moments,
  un = TRUE,
  binary = "std",
  addl = ~ I(age^2) + I(re74^2) + I(re75^2)
)

second_moment_terms <- c("I(age^2)", "I(re74^2)", "I(re75^2)")

ebal_tuning_comparison <- tibble(
  specification = c(
    "Default means only",
    "Tuned squares for age and prior earnings"
  ),
  control_ess = c(
    ebal_default_summary$effective.sample.size["Weighted", "Control"],
    ebal_moments_summary$effective.sample.size["Weighted", "Control"]
  ),
  max_control_weight = c(
    max(w_ebal_default$weights[dat$treat == 0]),
    max(w_ebal_moments$weights[dat$treat == 0])
  ),
  max_abs_adj_smd_means = c(
    max(abs(ebal_default_tuning_balance$Balance[setdiff(rownames(ebal_default_tuning_balance$Balance), second_moment_terms), "Diff.Adj"]), na.rm = TRUE),
    max(abs(ebal_moments_tuning_balance$Balance[setdiff(rownames(ebal_moments_tuning_balance$Balance), second_moment_terms), "Diff.Adj"]), na.rm = TRUE)
  ),
  max_abs_adj_smd_selected_squares = c(
    max(abs(ebal_default_tuning_balance$Balance[second_moment_terms, "Diff.Adj"]), na.rm = TRUE),
    max(abs(ebal_moments_tuning_balance$Balance[second_moment_terms, "Diff.Adj"]), na.rm = TRUE)
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

ebal_tuning_comparison
```

```{r}
w_ebal <- w_ebal_default
ebal_summary <- ebal_default_summary

dat_ebal <- dat |>
  mutate(ebal_weight = w_ebal$weights)

ebal_weight_diagnostics <- tibble(
  metric = c(
    "Treated units",
    "Control units",
    "Treated effective sample size",
    "Control effective sample size",
    "Maximum control weight",
    "Control coefficient of variation"
  ),
  value = c(
    sum(dat$treat == 1),
    sum(dat$treat == 0),
    ebal_summary$effective.sample.size["Weighted", "Treated"],
    ebal_summary$effective.sample.size["Weighted", "Control"],
    max(dat_ebal$ebal_weight[dat_ebal$treat == 0]),
    unname(ebal_summary$coef.of.var["control"])
  )
) |>
  mutate(value = round(value, 3))

ebal_weight_diagnostics

ebal_balance <- bal.tab(
  w_ebal,
  un = TRUE,
  binary = "std",
  disp.v.ratio = TRUE,
  m.threshold = 0.1
)

ebal_balance
```

```{r}
fit_ebal <- lm_weightit(
  outcome ~ treat,
  data = dat,
  weightit = w_ebal
)

ebal_effect <- {
  ebal_coef <- coef(summary(fit_ebal))["treat", ]
  tibble(
    method = "Entropy balancing",
    estimate = unname(ebal_coef["Estimate"]),
    std_error = unname(ebal_coef["Std. Error"]),
    p_value = unname(ebal_coef[4])
  )
} |>
  mutate(across(c(estimate, std_error, p_value), ~ round(.x, 3)))

ebal_effect
```

What to discuss:

- Entropy balancing can preserve the treated population while shrinking the control information.
- The weighting design is only defensible if the balance gain is worth the `ESS` loss and weight concentration.

## Step 7: Compare The Methods Side By Side

```{r}
comparison_balance <- tibble(
  method = c("Raw sample", "Exact matching", "CEM", "Entropy balancing"),
  max_abs_smd = c(
    max(abs(unadjusted_balance$Balance$Diff.Un), na.rm = TRUE),
    max(abs(exact_balance$Balance$Diff.Adj), na.rm = TRUE),
    max(abs(cem_balance$Balance$Diff.Adj), na.rm = TRUE),
    max(abs(ebal_balance$Balance$Diff.Adj), na.rm = TRUE)
  ),
  covariates_above_0_1 = c(
    sum(abs(unadjusted_balance$Balance$Diff.Un) > 0.1, na.rm = TRUE),
    sum(abs(exact_balance$Balance$Diff.Adj) > 0.1, na.rm = TRUE),
    sum(abs(cem_balance$Balance$Diff.Adj) > 0.1, na.rm = TRUE),
    sum(abs(ebal_balance$Balance$Diff.Adj) > 0.1, na.rm = TRUE)
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

comparison_balance
```

```{r}
comparison_information <- tibble(
  method = c("Exact matching", "CEM", "Entropy balancing"),
  treated_units_used = c(
    sum(matched_exact$treat == 1),
    sum(matched_cem$treat == 1),
    sum(dat$treat == 1)
  ),
  control_units_used = c(
    sum(matched_exact$treat == 0),
    sum(matched_cem$treat == 0),
    sum(dat$treat == 0)
  ),
  control_information = c(
    sum(matched_exact$treat == 0),
    sum(matched_cem$treat == 0),
    ebal_summary$effective.sample.size["Weighted", "Control"]
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

comparison_information
```

```{r}
treatment_effect_summary <- bind_rows(
  exact_effect,
  cem_effect,
  ebal_effect
)

treatment_effect_summary
```

## Step 8: Write The Design Judgement

Use your outputs to answer all five:

1. Which method gave the strongest observed balance?
2. Which method stayed closest to the original treated population?
3. Which method paid the highest information cost?
4. Which design would you defend for an evaluator-facing `ATT`?
5. What limitation would you still state even after adjustment?

## Suggested Close

If you want to go deeper after this page:

- review the [Matching and Weighting](../notes/matching/matching-and-weighting.qmd) note for the method overview
- move next to the [Black Politicians Lab](black-politicians-lab.qmd) for the first dataset extension
