---
title: "Lab: NSW and CPS Benchmark"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as a guided benchmark worksheet.

- Read the benchmark framing before running any model.
- Run each design branch in order: raw, exact, CEM, entropy balancing.
- Track retention, balance, and benchmark gap together.
- Treat the experimental NSW estimate as the reference target.

::: {.callout-note}
Code is shown but not executed when this page is rendered. That keeps the page readable even if your local R library is incomplete.
:::

## Training Goal

This lab focuses on one practical question:

Can observational adjustment recover the known NSW experimental result when the control pool comes from CPS?

In *Causal Inference: The Mixtape*, this example matters because it sits inside a longer debate about whether matching and propensity-score methods can recover credible programme effects from observational data. The NSW experiment gives a randomized benchmark, while the CPS comparison pool creates the harder observational problem. That makes the exercise more than a coding drill: it is a direct test of whether a design built on observed covariates can get back to something close to the experimental answer.

The workflow mirrors the main report's design-first sequence:

1. define the estimand
2. isolate benchmark versus observational datasets
3. diagnose raw imbalance and overlap
4. compare exact matching, CEM, and entropy balancing
5. evaluate each estimate against the experimental benchmark

## Dataset At A Glance

We use two `causaldata` datasets:

- `nsw_mixtape`: treated and control units from the **National Supported Work Demonstration (NSW)** job-training experiment
- `cps_mixtape`: observational controls drawn from the **Current Population Survey (CPS)**

### Acronyms And Background

- **NSW = National Supported Work Demonstration.**
  This was a U.S. randomized employment and training demonstration used as a benchmark in later causal-inference comparisons.  
  Background and evaluation context:
  [MDRC summary report (1980)](https://www.mdrc.org/work/publications/summary-and-findings-national-supported-work-demonstration),
  [LaLonde (1986), *American Economic Review* metadata](https://econpapers.repec.org/RePEc%3Aaea%3Aaecrev%3Av%3A76%3Ay%3A1986%3Ai%3A4%3Ap%3A604-20).

- **CPS = Current Population Survey.**
  This is the monthly household labor-force survey jointly run by the U.S. Census Bureau and the U.S. Bureau of Labor Statistics.  
  Official background:
  [Census CPS program page](https://www.census.gov/programs-surveys/cps.html),
  [BLS CPS overview](https://www.bls.gov/opub/hom/cps/).

### Dataset Provenance For This Lab

- `nsw_mixtape` documentation and variable definitions:
  [CRAN reference](https://search.r-project.org/CRAN/refmans/causaldata/html/nsw_mixtape.html) (`445` rows, `11` variables).
- `cps_mixtape` documentation and variable definitions:
  [CRAN reference](https://search.r-project.org/CRAN/refmans/causaldata/html/cps_mixtape.html) (`15,992` rows, `11` variables).
- The benchmark-comparison framing (NSW experiment versus observational controls) is tied to:
  [Dehejia and Wahba (1998/1999) NBER working-paper version](https://www.nber.org/papers/w6586) and its published *JASA* article listed there.

### Why This Example Matters In The Mixtape

The Mixtape uses the NSW setting to illustrate a core matching problem: a randomized study can tell us what the programme did in the experimental sample, but analysts often have to work with non-experimental control groups in practice. LaLonde's comparison between the experimental NSW estimate and observational alternatives made this dataset a benchmark case because many standard non-experimental estimators performed poorly against the experimental result. Later work, especially Dehejia and Wahba, argued that design choices such as covariate selection and propensity-score-based adjustment could do better.

For study purposes, the important context is:

- NSW gives the benchmark because treatment and control were randomly assigned inside that sample.
- CPS gives the observational challenge because those controls were not randomized against the NSW treated group.
- The exercise is therefore not "estimate a treatment effect from scratch." It is "see how close different observational designs get to a reference effect that is already available from the experiment."
- This is why balance, overlap, retained sample size, and benchmark gap should all be interpreted together.

Core setup:

- experimental benchmark: `ATT` from NSW treated versus NSW controls
- observational design: NSW treated versus CPS controls
- outcome: `re78` (post-program earnings)
- covariates: `age`, `educ`, `black`, `hisp`, `marr`, `nodegree`, `re74`, `re75`

## Step 1: Load Packages And Build Analysis Datasets

```{r}
required_packages <- c(
  "causaldata",
  "MatchIt",
  "WeightIt",
  "cobalt",
  "dplyr",
  "ggplot2"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

nsw <- causaldata::nsw_mixtape |>
  mutate(
    treat = as.integer(treat),
    outcome = re78
  )

cps <- causaldata::cps_mixtape |>
  mutate(
    treat = as.integer(treat),
    outcome = re78
  )

# Experimental benchmark sample: NSW treated + NSW controls.
benchmark_dat <- nsw

# Observational sample: NSW treated + CPS controls.
obs_dat <- bind_rows(
  nsw |> filter(treat == 1L) |> mutate(source = "NSW treated"),
  cps |> filter(treat == 0L) |> mutate(source = "CPS controls")
)

design_covariates <- c("age", "educ", "black", "hisp", "marr", "nodegree", "re74", "re75")

obs_dat |>
  select(source, treat, outcome, all_of(design_covariates)) |>
  glimpse()
```

Checkpoint:

- Is the benchmark estimate coming only from NSW randomized groups?
- Is the observational sample restricted to NSW treated plus CPS controls?

## Step 2: Compute The Experimental Benchmark

```{r}
benchmark_summary <- benchmark_dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    mean_re78 = mean(outcome, na.rm = TRUE),
    mean_re74 = mean(re74, na.rm = TRUE),
    mean_re75 = mean(re75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "NSW treated", "NSW controls")) |>
  select(group, n, mean_re78, mean_re74, mean_re75)

benchmark_fit <- lm(outcome ~ treat, data = benchmark_dat)
benchmark_att <- coef(summary(benchmark_fit))["treat", ]

benchmark_summary
benchmark_att
```

This `treat` coefficient is the reference effect we try to recover with observational adjustment.

## Step 3: Raw Observational Estimate (NSW Treated Vs CPS Controls)

```{r}
raw_summary <- obs_dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    mean_re78 = mean(outcome, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE),
    mean_educ = mean(educ, na.rm = TRUE),
    prop_black = mean(black == 1, na.rm = TRUE),
    mean_re74 = mean(re74, na.rm = TRUE),
    mean_re75 = mean(re75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "NSW treated", "CPS controls")) |>
  select(group, n, mean_re78, mean_age, mean_educ, prop_black, mean_re74, mean_re75)

raw_fit <- lm(outcome ~ treat, data = obs_dat)
raw_obs_att <- coef(summary(raw_fit))["treat", ]

raw_summary
raw_obs_att
```

Compare this raw observational estimate with the benchmark from Step 2.

## Step 4: Baseline Balance And Overlap Diagnostics

```{r}
m_raw <- matchit(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  method = NULL,
  estimand = "ATT"
)

summary(m_raw, un = TRUE)

cobalt::bal.tab(
  m_raw,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

cobalt::love.plot(
  m_raw,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)
```

```{r}
# Quick propensity-score overlap check.
ps_fit <- glm(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  family = binomial()
)

obs_dat <- obs_dat |>
  mutate(ps = predict(ps_fit, type = "response"))

overlap_bounds <- obs_dat |>
  group_by(treat) |>
  summarise(
    min_ps = min(ps, na.rm = TRUE),
    max_ps = max(ps, na.rm = TRUE),
    .groups = "drop"
  )

common_support_low <- max(overlap_bounds$min_ps)
common_support_high <- min(overlap_bounds$max_ps)

overlap_summary <- obs_dat |>
  group_by(treat) |>
  summarise(
    n = n(),
    min_ps = min(ps, na.rm = TRUE),
    p10_ps = quantile(ps, 0.10, na.rm = TRUE),
    p50_ps = quantile(ps, 0.50, na.rm = TRUE),
    p90_ps = quantile(ps, 0.90, na.rm = TRUE),
    max_ps = max(ps, na.rm = TRUE),
    in_common_support = mean(ps >= common_support_low & ps <= common_support_high),
    .groups = "drop"
  ) |>
  mutate(group = if_else(treat == 1L, "NSW treated", "CPS controls")) |>
  select(group, n, min_ps, p10_ps, p50_ps, p90_ps, max_ps, in_common_support)

overlap_summary
```

Questions:

1. Which covariates are most imbalanced before adjustment?
2. How much treated-support mass lies in thin CPS-support regions?

## Step 5: Exact Matching Branch

```{r}
m_exact <- matchit(
  treat ~ black + hisp + marr + nodegree + educ,
  data = obs_dat,
  method = "exact",
  estimand = "ATT"
)

summary(m_exact, un = TRUE)

exact_dat <- match.data(m_exact)

exact_dat |>
  count(treat) |>
  mutate(group = if_else(treat == 1L, "Treated", "Control"))

cobalt::bal.tab(
  m_exact,
  data = obs_dat,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1,
  addl = ~ age + re74 + re75
)
```

Interpretation prompt:

- Exact matching can preserve transparency, but does it solve imbalance on prior earnings?

## Step 6: CEM Branch

```{r}
cem_cutpoints <- list(
  age = "q5",
  educ = 4,
  re74 = "q6",
  re75 = "q6"
)

m_cem <- matchit(
  treat ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
  data = obs_dat,
  method = "cem",
  estimand = "ATT",
  cutpoints = cem_cutpoints
)

summary(m_cem, un = TRUE)

cobalt::bal.tab(
  m_cem,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

cobalt::love.plot(
  m_cem,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)

cem_dat <- match.data(m_cem)

cem_dat |>
  count(treat) |>
  mutate(group = if_else(treat == 1L, "Treated", "Control"))
```

Track whether CEM improves balance while retaining enough treated units to keep the `ATT` target credible.

## Step 7: Entropy Balancing Branch

```{r}
w_ebal <- weightit(
  treat ~ age + I(age^2) +
    educ + I(educ^2) +
    black + hisp + marr + nodegree +
    re74 + I(re74^2) +
    re75 + I(re75^2),
  data = obs_dat,
  method = "ebal",
  estimand = "ATT"
)

summary(w_ebal)

cobalt::bal.tab(
  w_ebal,
  un = TRUE,
  binary = "std",
  m.threshold = 0.1
)

cobalt::love.plot(
  w_ebal,
  abs = TRUE,
  thresholds = c(m = 0.1),
  stars = "raw"
)

ess <- function(w) {
  (sum(w)^2) / sum(w^2)
}

obs_ebal <- obs_dat |>
  mutate(w = w_ebal$weights)

obs_ebal |>
  group_by(treat) |>
  summarise(
    raw_n = n(),
    ess = ess(w),
    max_weight = max(w),
    mean_weight = mean(w),
    .groups = "drop"
)
```

If balance is excellent but control ESS collapses, note that as a design cost.

## Step 8: Compare Against The Experimental Benchmark

```{r}
exact_fit <- lm(outcome ~ treat, data = exact_dat, weights = weights)
cem_fit <- lm(outcome ~ treat, data = cem_dat, weights = weights)
ebal_fit <- lm(outcome ~ treat, data = obs_ebal, weights = w)

get_treat_row <- function(fit, design_name) {
  out <- coef(summary(fit))
  data.frame(
    design = design_name,
    estimate = unname(out["treat", "Estimate"]),
    std_error = unname(out["treat", "Std. Error"]),
    p_value = unname(out["treat", "Pr(>|t|)"]),
    row.names = NULL,
    check.names = FALSE
  )
}

experimental_estimate <- unname(benchmark_att["Estimate"])

estimate_comparison <- bind_rows(
  get_treat_row(benchmark_fit, "Experimental benchmark (NSW RCT)"),
  get_treat_row(raw_fit, "Raw observational (NSW treated vs CPS)"),
  get_treat_row(exact_fit, "Exact matching"),
  get_treat_row(cem_fit, "CEM"),
  get_treat_row(ebal_fit, "Entropy balancing")
) |>
  mutate(
    benchmark_gap = estimate - experimental_estimate,
    abs_benchmark_gap = abs(benchmark_gap)
  )

estimate_comparison
```

This table is the core benchmark output for the lab.

## Comparison Table To Fill In

| Design | Worst abs. SMD | Treated retained | Control ESS | Estimate | Gap vs benchmark | Your judgement |
|---|---|---|---|---|---|---|
| Raw observational |  |  |  |  |  |  |
| Exact matching |  |  |  |  |  |  |
| CEM |  |  |  |  |  |  |
| Entropy balancing |  |  |  |  |  |  |

## Debrief Questions

1. Which method got closest to the NSW benchmark?
2. Which method's benchmark closeness depended most on aggressive reweighting?
3. Did better benchmark recovery come from tighter matching, richer balance constraints, or both?
4. If you had to defend one design to a policy audience, which would you choose and why?

## Optional Extension

Add one sensitivity branch and compare benchmark drift:

- trim extreme entropy weights (`trim()` in `WeightIt`)
- rerun CEM with slightly coarser and finer `re74`/`re75` bins
- report how benchmark gap and control ESS move under each choice
