---
title: "Lab: RDD design diagnostics and sensitivity"
notebook-language: r
aliases:
  - /labs/regression-discontinuity-diagnostics-python-lab.html
format:
  html: default
execute:
  echo: true
  eval: false
---

## How to use this lab

Allow 60–75 minutes. You should be comfortable with basic regression and R data frames. Restore the [tested environment](regression-discontinuity-reproducibility.qmd), then run every cell in order in a fresh kernel. Every run loads and verifies the bundled local data. No data are downloaded. The HTML page displays code; the downloadable notebook executes it.

<!-- lab-downloads:start -->
::: {.resource-links}
[Quarto document](downloads/regression-discontinuity-diagnostics-lab.qmd){download="regression-discontinuity-diagnostics-lab.qmd"}
[R Jupyter notebook](notebooks/regression-discontinuity-diagnostics-lab.ipynb){download="regression-discontinuity-diagnostics-lab.ipynb"}
[Complete lab ZIP](downloads/regression-discontinuity-diagnostics-lab.zip){download="regression-discontinuity-diagnostics-lab.zip"}
[Data and setup](data.qmd#regression-discontinuity-diagnostics-lab)
:::
<!-- lab-downloads:end -->

[RDD overview](../notes/other-methods/regression-discontinuity.qmd) · [Source reading map](../notes/rdd/regression-discontinuity-sources.qmd)

## Learning objectives

1. Distinguish the outcome sample from the sample used for density testing.
2. Evaluate observed-covariate discontinuities without assuming they are baseline measurements.
3. Audit bandwidth, polynomial, placebo and donut choices.
4. Report checks as challenges to a design, not a pass/fail certificate.

## Research question and design

Does threshold-based access to Uruguay's PANES transfer programme change reported support for government? These teaching data come from Manacorda, Miguel and Vigorito and the worked example in *The Effect* [@manacorda2011transfers; @huntingtonklein2025rdd]. They are a simplified teaching sample, not a replication of every paper specification.

| Component | Definition |
|---|---|
| Running variable | `Income_Centered`, an assignment-income measure already centered at zero |
| Treatment side | Below zero; verify against `Participation` |
| Outcome | `Support`: 0 = worse than previous government, 0.5 = same, 1 = better |
| Population | Survey observations already restricted to about ±0.02 of the threshold |
| Estimand | Effect of programme access at the threshold, under continuity and no competing discontinuity |

The outcome is a **support score**, not a binary probability. A change of 0.03 means 0.03 score units, or 3 points on a rescaled 0–100 index; it does not mean a 3-percentage-point change in the probability of supporting government. The programme includes benefits beyond a single cash payment, so interpret the treatment as the programme package.

Complete the [foundations lab](regression-discontinuity-foundations-lab.qmd) first.

## 1. Load both samples

```{r}
data_helpers <- c("data/load-data.R", "../data/load-data.R", "docs/labs/data/load-data.R")
data_helpers <- data_helpers[file.exists(data_helpers)]
if (!length(data_helpers)) stop("Extract the complete lab ZIP, including its data folder, before running.")
source(data_helpers[[1]])
```

```{r}
# Restore the isolated library documented on the setup page before running.
required <- c("rdrobust", "rddensity", "digest", "jsonlite")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly=TRUE)]
if (length(missing)) stop("Restore the RDD environment; missing: ", paste(missing, collapse=", "))
load_data <- function(name) as.data.frame(qed_data(name))
require_support <- function(x, h, cutoff=0, order=2) {
  x <- as.numeric(x) - cutoff
  if (!is.finite(h) || h <= 0 || !all(is.finite(x))) stop("Nonfinite score or invalid bandwidth")
  for (side in list(x[x < 0 & x > -h], x[x >= 0 & x < h])) {
    if (length(side) < 10 || length(unique(side)) < order+2)
      stop("Insufficient observations or distinct scores on a cutoff side")
  }
  invisible(TRUE)
}
rd_fit <- function(y, x, h=NULL, p=1, cutoff=0, treatment=NULL) {
  args <- list(y=as.numeric(y), x=as.numeric(x), c=cutoff, p=p, q=p+1,
               kernel="triangular", vce="hc0", bwselect="mserd", masspoints="adjust",
               stdvars=TRUE, level=95, bwrestrict=TRUE, scaleregul=1)
  if (!is.null(h)) {
    require_support(x, h, cutoff, p+1)
    args$h <- h; args$b <- h
  }
  if (!is.null(treatment)) args$fuzzy <- as.numeric(treatment)
  do.call(rdrobust::rdrobust, args)
}
rd_row <- function(fit, label) {
  data.frame(model=label, jump=fit$coef[1,1], bias_corrected=fit$coef[3,1],
             se_robust=fit$se[3,1], ci_low=fit$ci[3,1], ci_high=fit$ci[3,2],
             h_left=fit$bws[1,1], h_right=fit$bws[1,2],
             b_left=fit$bws[2,1], b_right=fit$bws[2,2],
             n_left=fit$N_h[1], n_right=fit$N_h[2], row.names=NULL)
}
binned_plot <- function(x, y, width, ylabel, xlabel="Centered assignment score") {
  bins <- cut(x, breaks=seq(-width, width, length.out=31), include.lowest=TRUE)
  means <- aggregate(cbind(x,y), list(bin=bins), mean)
  plot(means$x, means$y, pch=19, col="#174c63", xlab=xlabel, ylab=ylabel)
  abline(v=0, lty=2)
  invisible(means)
}
print(vapply(required, function(p) as.character(packageVersion(p)), character(1)))
```
```{r}
gt <- load_data("gov_transfers")
stopifnot(identical(names(gt), c("Income_Centered", "Education", "Age", "Participation", "Support")))
stopifnot(!anyNA(gt[c("Income_Centered", "Participation", "Support")]))
x <- gt$Income_Centered; y <- gt$Support
stopifnot(!any(x == 0), all(gt$Participation == as.integer(x < 0)))
stopifnot(setequal(unique(y), c(0, 0.5, 1)))
print(table(participation=gt$Participation, eligible=x < 0))
print(colSums(is.na(gt)))
print(c(n=nrow(gt), x_min=min(x), x_max=max(x)))
```

## 2. Score distribution and density discontinuity

Use the dedicated score dataset, not the 1,948-person outcome survey. Retain its observations within ±0.02 for this exercise. Density bandwidths are selected separately from outcome bandwidths.

```{r}
scores <- load_data("gov_transfers_density")
xd <- scores$Income_Centered
stopifnot(all(is.finite(xd)))
xd <- xd[abs(xd) < 0.02]
require_support(xd, 0.02, order=3)
hist(xd, breaks=seq(-0.02,0.02,length.out=41), col="#174c63",
     main="", xlab="Centered assignment score", ylab="Records in density sample")
abline(v=0,lty=2)
density_fit <- rddensity::rddensity(xd, c=0, p=2, q=3, fitselect="unrestricted",
  kernel="triangular", vce="jackknife", massPoints=TRUE, regularize=TRUE, bwselect="comb", bino=FALSE)
print(summary(density_fit))
density_results <- data.frame(metric=c("source_n","analysis_n","test_stat","p_value","h_left","h_right"),
  value=c(nrow(scores),length(xd),density_fit$test$t_jk,density_fit$test$p_jk,
          density_fit$h$left,density_fit$h$right))
print(density_results)
```

**Checkpoint:** the full source has 52,549 score records. Explain why surveying outcomes near a threshold could itself alter the observed score distribution. A large density-test p-value is not proof that scores were never manipulated.

## 3. Observed-covariate checks and missingness

Examine `Age` and `Education` with the same fixed 0.01 window. Use complete cases separately for each variable and report the number excluded. Their measurement timing is not established in the teaching data, so call these **observed-covariate checks**, not verified pre-treatment balance tests. Household composition and education could themselves be related to programme exposure.

```{r}
covariates <- do.call(rbind,lapply(c("Age","Education"), function(variable) {
  sample <- gt[complete.cases(gt[c("Income_Centered",variable)]),]
  row <- rd_row(rd_fit(sample[[variable]],sample$Income_Centered,h=0.01),variable)
  row$available_n <- nrow(sample); row$missing_n <- nrow(gt)-nrow(sample)
  row
}))
print(covariates)
stopifnot(covariates$missing_n[covariates$model == "Education"] == 51)
```

Explain any discontinuity rather than adding the variable mechanically as a control. Multiple noisy checks can produce isolated findings by chance; a common-sample comparison would also be needed to distinguish adjustment from sample changes.

## 4. Bandwidth and polynomial sensitivity

Use triangular kernels, `b=h`, and paired orders `(p,q)=(1,2)` and `(2,3)`. Every estimate uses the same outcome definition. Draw intervals around the bias-corrected programme effect, not around the conventional jump.

```{r}
rows <- list()
for (h in c(0.005,0.01,0.015,0.02)) for (p in c(1,2)) {
  row <- rd_row(rd_fit(y,x,h=h,p=p),sprintf("h%.3f_p%d",h,p))
  row$bandwidth <- h; row$order <- p
  rows[[length(rows)+1]] <- row
}
sensitivity <- do.call(rbind,rows)
print(sensitivity)
plot(NA,xlim=c(0.004,0.021),ylim=range(-sensitivity$ci_high,-sensitivity$ci_low),
     xlab="Bandwidth",ylab="Programme effect on support score (RBC 95% CI)")
for (p in c(1,2)) {
  s <- sensitivity[sensitivity$order == p,]
  lines(s$bandwidth,-s$bias_corrected,type="b",pch=15+p,col=p+1)
  arrows(s$bandwidth,-s$ci_high,s$bandwidth,-s$ci_low,angle=90,code=3,length=.04,col=p+1)
}
abline(h=0,lty=2);legend("topright",legend=c("Local linear","Local quadratic"),col=2:3,lty=1)
```

The source is already truncated to approximately ±0.02. A result at the widest available window is not an estimate using the full programme population. Changing plot bins would be a different exercise from changing these estimation windows.

## 5. Placebo thresholds and donut exclusions

Use artificial cutoffs at −0.01 and +0.01, each with a 0.004 estimation **and bias** bandwidth. Neither window crosses the genuine zero cutoff. Then, at the real cutoff, remove score distances below 0.001 and 0.002 while keeping `h=b=0.01`. Donut results rely on extrapolation across the excluded gap; their intervals are model-based sensitivity summaries, not a repair for sorting or proof of validity.

```{r}
placebos <- do.call(rbind,lapply(c(-0.01,0.01),function(cutoff) {
  h <- 0.004;stopifnot(abs(cutoff)>h)
  same_regime <- if (cutoff < 0) x < 0 else x > 0
  row <- rd_row(rd_fit(y[same_regime],x[same_regime],h=h,cutoff=cutoff),paste0("cutoff_",cutoff))
  row$cutoff <- cutoff;row
}))
print(placebos)
donuts <- do.call(rbind,lapply(c(0,0.001,0.002),function(radius) {
  keep <- abs(x) >= radius
  row <- rd_row(rd_fit(y[keep],x[keep],h=0.01),sprintf("donut_%.3f",radius))
  row$radius <- radius;row$removed_n <- sum(!keep);row
}))
print(donuts)
```

## 6. Design assessment and worked checkpoints

The density exercise keeps **20,463 of 52,549** records. Its jackknife statistic is **−0.923780049** (`p=0.355600845`); this does not establish absence of manipulation. The observed-age bias-corrected jump is **6.5123 years**, with interval **[0.2769, 12.7477]**. Investigate this finding and measurement timing; do not declare baseline balance verified. Education uses **1,897** complete cases, of which **521 below** and **388 above** zero fall inside the 0.01 window.

At bandwidths 0.005, 0.01, 0.015 and 0.02, the conventional local linear **programme effects** are **−0.02716, +0.03348, +0.07974 and +0.09585**. The sensitivity plot uses bias-corrected estimates instead, so its centers differ from this list. The narrow quadratic fit is especially unstable. Donut radii 0.001 and 0.002 remove **54 and 167** outcome-survey records. Record these changes alongside the estimates in your diagnostic table.

See the [reproduction record](regression-discontinuity-reproducibility.qmd) for verified outputs and the complete comparison tables.

Write a short assessment covering: the administrative allocation rule; what density and covariate checks do and do not establish; stability across windows and fits; and which uncertainty matters for the policy conclusion. Do not choose a result by significance, or describe every nonsignificant diagnostic as a successful validation.

Next, study [fuzzy RD and local IV](regression-discontinuity-fuzzy-lab.qmd).
