---
title: "Regression Discontinuity for Causal Inference"
subtitle: "Assignment thresholds, local comparisons, and sharp and fuzzy designs"
author: "Luke Heley"
date: 19 September 2026
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  fig-width: 7
  fig-height: 4
---

## Executive summary

Regression discontinuity (RD) uses a known threshold in treatment assignment to
construct a local comparison. Credibility comes from the assignment rule and the
continuity of relevant potential outcomes near the threshold, not from finding
a discontinuity in an outcome graph. The effect concerns units at the cutoff;
generalizing to everyone eligible requires additional evidence.

This report uses Uruguay's PANES transfer programme for a sharp design and a
veteran-status example for fuzzy RD. Both use bundled data. The worked analysis
keeps treatment direction, conventional estimates, bias correction, uncertainty,
and score support visible. It does not treat a sequence of insignificant diagnostic
tests as proof of validity [@gertler2016ch6rdd; @cunningham2021rdd].

## Assignment, assumptions, and estimands

### Start with the institutional rule

Identify the score, cutoff, eligibility rule, actual receipt, outcome, measurement
dates, and population. Ask who could influence the score and whether other policies
change at the same threshold. Verify the exact assignment score rather than a
later revision. Record whether treatment occurs above or below the cutoff.

In sharp RD, receipt changes deterministically at the threshold. In fuzzy RD,
crossing changes the probability of receipt. Imperfect compliance is not merely
an inconvenient sharp design: it changes the estimand and requires an instrumental
variables argument [@huntingtonklein2025rdd].

### State a local causal claim

Write the centered score as $X$ and the cutoff as zero. Under continuity of the
potential-outcome means, with treatment on the right, sharp RD identifies

$$
\tau_{SRD}=\lim_{x\downarrow 0}E[Y\mid X=x]
           -\lim_{x\uparrow 0}E[Y\mid X=x].
$$

With treatment on the left, reverse the sign to report treated minus untreated.
Continuity rules out a separate outcome discontinuity in the absence of treatment;
it does not require the two sides to be identical far from the threshold. Precise
sorting, selective observation, or a coincident eligibility rule can undermine it.

Compared with matching, the design uses a known assignment mechanism rather than
selection on observed covariates. Compared with DiD, it compares near-threshold
units rather than untreated changes over time. Its strength is a potentially
credible local comparison; its cost is limited external validity and sensitivity
to the information available near the cutoff.

## PANES: audit the sharp design

The outcome sample contains 1,948 observations around the threshold. Programme
participation occurs on the negative side of the centered score. `Support` takes
values 0, 0.5, and 1: it is a government-support score, not a binary probability.
The treatment is a programme package, not an isolated cash transfer
[@manacorda2011transfers].

```{r}
data_helpers <- c("data/load-data.R", "../data/load-data.R", "docs/labs/data/load-data.R")
data_helpers <- data_helpers[file.exists(data_helpers)]
if (!length(data_helpers)) stop("Keep the bundled data folder beside the notebook.")
source(data_helpers[[1]])
library(rdrobust)
library(rddensity)
library(ggplot2)
gt <- as.data.frame(qed_data("gov_transfers"))
gt$side <- as.integer(gt$Income_Centered >= 0)
stopifnot(nrow(gt) == 1948,
  !anyNA(gt[c("Income_Centered", "Participation", "Support")]),
  !any(gt$Income_Centered == 0),
  all(gt$Participation == as.integer(gt$Income_Centered < 0)))
knitr::kable(with(gt, table(participation=Participation, right=side)))
knitr::kable(data.frame(variable=names(gt), missing=colSums(is.na(gt))))
```

The supplied survey is restricted to roughly 0.02 on either side. That is sample
support, not an optimal bandwidth. Missing education values should not silently
drop otherwise usable outcome observations. The [data catalogue](data.qmd#gov-transfers)
records provenance and provides local downloads.

```{r}
#| fig-cap: "Government support around the eligibility threshold; treatment is on the left."
#| fig-alt: "Binned mean government-support scores on each side of zero, with separate colors and a marked cutoff."
gt$bin <- cut(gt$Income_Centered, seq(-0.02,0.02,by=0.001), include.lowest=TRUE)
bins <- aggregate(cbind(Income_Centered, Support) ~ bin + side, gt, mean)
ggplot(bins, aes(Income_Centered, Support, color=factor(side))) +
  geom_point(size=2) + geom_vline(xintercept=0, linetype=2) +
  scale_color_manual(values=c("#177b72", "#245ca4"), guide="none") +
  labs(x="Centered assignment score", y="Mean government-support score") +
  theme_minimal(base_size=11)
```

Binning is descriptive, not the effect estimator. Do not allow a bin to cross the
cutoff. A visual bin width and an estimation bandwidth serve different purposes.
Avoid a global high-order polynomial chosen because it draws an appealing curve.

## Local estimation and robust inference

### Make the conventional jump transparent

Inside a window $|X|<h$, fit separate lines by interacting the score with a
right-side indicator. Triangular weights $1-|X|/h$ give greater weight to observations
near zero. The intercept difference is right minus left.

```{r}
h <- 0.01
local <- subset(gt, abs(Income_Centered) < h)
local$weight <- 1-abs(local$Income_Centered)/h
manual <- lm(Support ~ side*Income_Centered, local, weights=weight)
jump <- unname(coef(manual)["side"])
stopifnot(sum(local$side == 0) == 537, sum(local$side == 1) == 400)
knitr::kable(data.frame(right_minus_left=jump, programme_effect=-jump), digits=4)
```

The conventional programme contrast is about +0.0335 score units. Calling it a
3.35-percentage-point change in a binary support probability would be incorrect.
The window and weights define a local approximation, not a global programme ATT.

### Account for boundary approximation bias

Local regression has approximation bias at the boundary. Ordinary weighted least
squares intervals do not address that bias. Use an RD procedure that reports the
conventional estimate, bias-corrected estimate, and robust interval distinctly.

```{r}
fit_rd <- function(y, x, h=NULL, treatment=NULL) {
  args <- list(y=y, x=x, c=0, p=1, q=2, kernel="triangular",
    vce="hc0", masspoints="adjust", stdvars=TRUE,
    bwselect="mserd", bwrestrict=TRUE, scaleregul=1, level=95)
  if (!is.null(h)) { args$h <- h; args$b <- h }
  if (!is.null(treatment)) args$fuzzy <- treatment
  do.call(rdrobust, args)
}
fixed <- fit_rd(gt$Support, gt$Income_Centered, 0.01)
automatic <- fit_rd(gt$Support, gt$Income_Centered)
programme <- function(fit, label) {
  data.frame(model=label, conventional=-fit$coef[1,1],
    corrected=-fit$coef[3,1], lower=-fit$ci[3,2], upper=-fit$ci[3,1])
}
stopifnot(abs(fixed$coef[1,1]-jump) < 1e-8,
          abs(-fixed$coef[3,1]-(-0.0416)) < 0.0001)
knitr::kable(rbind(programme(fixed,"Fixed window"),
                  programme(automatic,"MSE selected")), digits=4)
knitr::kable(automatic$bws, digits=4, caption="Automatic estimation and bias bandwidths")
```

Here $p=1$ is the local linear order and $q=2$ the order used for bias correction.
The estimation bandwidth $h$ and bias bandwidth $b$ have different roles.
Automatic MSE selection does not establish design validity. Reversing treatment
direction requires negating estimates and swapping the negated interval endpoints.

The fixed-window corrected programme effect is approximately -0.0416 with a robust
interval of [-0.1880, 0.1048]. The conventional and corrected signs differ. Report
this sensitivity rather than attaching a robust interval to an unlabeled estimate
or selecting the more attractive sign. The current data and specifications support
a local, imprecise conclusion.

## Threats to continuity and sensitivity

Use the separate assignment-score dataset for density analysis; the outcome
survey's selection can alter the score distribution. An insignificant density
test is not proof that manipulation is absent.

```{r}
scores <- qed_data("gov_transfers_density")
x <- scores$Income_Centered
x <- x[abs(x) < 0.02]
stopifnot(nrow(scores) == 52549, length(x) == 20463)
density <- rddensity(x, c=0, p=2, q=3, fitselect="unrestricted",
  kernel="triangular", vce="jackknife", massPoints=TRUE,
  regularize=TRUE, bwselect="comb", bino=FALSE)
summary(density)
```

Inspect covariates known to predate treatment and whether observation probability
changes at the threshold. The timing of `Age` and `Education` in this teaching
file is not established as pre-treatment. Describe their comparisons as
observed-covariate checks, not confirmed baseline balance. Education has 51 missing
values; report variable-specific samples.

```{r}
sensitivity <- do.call(rbind, lapply(c(0.005,0.01,0.015,0.02), function(h) {
  programme(fit_rd(gt$Support, gt$Income_Centered, h), paste("h =", h))
}))
knitr::kable(sensitivity, digits=4)
```

A bandwidth grid exposes dependence on distant scores and the bias/variance
tradeoff. Compare a local quadratic only where support permits it. Placebo cutoffs
must remain within a single treatment regime, including the bias window. A donut
exclusion removes the observations closest to the threshold and requires
extrapolation back to zero; it is not a generic cure for sorting. The diagnostics
lab implements these checks with explicit support safeguards.

## Fuzzy RD: a local instrumental-variables extension

The veteran-status example uses birth-quarter variation in receipt and a binary
homeownership outcome [@fetter2013mortgages]. Under continuity, relevance, exclusion,
and an appropriately oriented monotonicity assumption, the ratio of outcome and
receipt jumps identifies an effect for local compliers:

$$
\tau_{FRD}=\frac{\Delta E[Y\mid X=0]}{\Delta E[D\mid X=0]}.
$$

Use the same observations, weights, and score controls in both jumps. Veteran
status falls from left to right; do not reverse only one jump. An effect of veteran
status is not automatically an isolated effect of a mortgage subsidy, because
exclusion concerns every other channel affected by the threshold.

```{r}
raw <- as.data.frame(qed_data("mortgages"))
columns <- c("qob_minus_kw", "vet_wwko", "home_ownership")
v <- raw[complete.cases(raw[columns]) & !is.na(raw$qob_minus_kw) &
           abs(raw$qob_minus_kw) < 12, columns]
stopifnot(nrow(raw) == 214144, nrow(v) == 56901)
ratio <- function(h) {
  a <- subset(v, abs(qob_minus_kw) < h)
  a$side <- as.integer(a$qob_minus_kw >= 0)
  a$weight <- 1-abs(a$qob_minus_kw)/h
  first <- lm(vet_wwko ~ side*qob_minus_kw, a, weights=weight)
  reduced <- lm(home_ownership ~ side*qob_minus_kw, a, weights=weight)
  fs <- unname(coef(first)["side"])
  rf <- unname(coef(reduced)["side"])
  if (!is.finite(fs) || abs(fs) < 0.01) stop("Unusable first stage")
  data.frame(h=h, first_stage=fs, reduced_form=rf, ratio=rf/fs)
}
knitr::kable(do.call(rbind, lapply(c(6,9,12), ratio)), digits=4)
fuzzy <- fit_rd(v$home_ownership, v$qob_minus_kw, 12, v$vet_wwko)
stopifnot(abs(fuzzy$coef[1,1]-ratio(12)$ratio) < 1e-8)
summary(fuzzy)
```

At 12 quarters the conventional ratio is about 0.1863, or 18.63 percentage points
of homeownership probability. The corrected estimate is about 0.3093 with a robust
interval near [0.1057, 0.5130]. These are different inferential objects. The 0.01
first-stage safeguard is numerical, not a weak-instrument test; report first-stage
uncertainty and use weak-identification methods when needed.

Despite many people, there are only 12 distinct quarters on each side in the
widest window. More records at the same score do not create arbitrarily close
comparisons. Treat continuous-score asymptotics cautiously, discuss extrapolation,
and do not apply a continuous density test to these discrete quarters. Clustering
by quarter does not automatically solve functional-form bias.

## Reporting and teaching route

Report the assignment rule, treatment direction, population, score support,
missingness, kernel, polynomial orders, both bandwidths, effective sample sizes,
conventional and corrected effects, and robust intervals. Include diagnostic
evidence and sensitivity, then state whose local effect is identified and why
generalization beyond the cutoff may fail.

- [Teaching deck](../slides/rdd.qmd) and [presenter notes](../slides/rdd_script.qmd).
- [Practical R guide](../notes/rdd/how-to-do-regression-discontinuity.qmd) for the shorter procedural route.
- [Sharp foundations lab](regression-discontinuity-foundations-lab.qmd) for assignment and local estimation.
- [Diagnostics lab](regression-discontinuity-diagnostics-lab.qmd) for density, covariates, and sensitivity.
- [Fuzzy RD lab](regression-discontinuity-fuzzy-lab.qmd) for local ratios, IV, and coarse support.
- [Reproduction record](regression-discontinuity-reproducibility.qmd) for environment and numerical checks.

The [Labs directory](index.qmd#regression-discontinuity-labs) supplies the Quarto,
R Jupyter, and complete offline downloads for each exercise.

## References
