---
title: "Lab: Augmented SCM With Kansas"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as the advanced SCM lab.

- Keep the [Synthetic Control](../notes/scm/synthetic-control.qmd) overview open for the baseline design conditions.
- Use the [Augmented Synthetic Control Method notes](../notes/scm/augmented-synthetic-control-method-notes.qmd) when you want the paper-level rationale for controlled extrapolation.
- Read this lab as a design comparison, not as a software upgrade tutorial.
- End by deciding whether the Kansas case supports classical SCM, augmented SCM, or neither.

::: {.callout-note}
The code is shown but not executed when the site is rendered. That keeps the page readable while preserving a runnable workflow for teaching and self-study.
:::

## Training Goal

Use the Kansas tax-cut case to understand why augmentation exists:

1. build a clean comparative-case panel
2. fit a classical SCM benchmark
3. fit a ridge-augmented SCM on the same outcome
4. compare the two designs on pre-treatment fit, interpretability, and extrapolation risk
5. state whether augmentation solves a real design problem or only hides one

## Dataset At A Glance

This lab uses `augsynth::kansas`, a quarterly state panel used in the ASCM paper's empirical illustration.

- Substantive case: the Kansas tax cuts associated with the Brownback period
- Teaching role: advanced extension case rather than first exposure to SCM
- Main outcome here: log GDP per capita
- Main lesson: poor classical fit is a warning signal, and augmentation is a controlled response to that warning rather than an automatic improvement

## What To Hand Back

By the end of the lab, you should be able to report:

- how the Kansas panel was collapsed into an annual comparative-case dataset
- whether classical SCM alone fits Kansas well enough before treatment
- what changes when ridge augmentation is introduced
- why negative or non-simplex weighting is a cost rather than a free benefit
- whether the design should be defended with classical SCM, augmented SCM, or abandoned for a different strategy

## Step 1: Load Packages And Inspect The Kansas Panel

```{r}
required_packages <- c(
  "augsynth",
  "dplyr",
  "ggplot2",
  "tibble"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

data("kansas", package = "augsynth")

kansas |>
  glimpse()
```

Checkpoint:

- The raw data are quarterly and much richer than the smoking case.
- That makes this a better lab for weak-fit problems than for first-pass SCM mechanics.

## Step 2: Collapse To A Clean Annual Panel

This is a teaching choice. The underlying data are quarterly, but an annual panel makes the design logic easier to inspect in a first advanced lab.

```{r}
kansas_annual <- kansas |>
  filter(year >= 1990, year <= 2016) |>
  group_by(state, year) |>
  summarise(
    treated = max(treated, na.rm = TRUE),
    lngdpcapita = mean(lngdpcapita, na.rm = TRUE),
    revenuepop = mean(revenuepop, na.rm = TRUE),
    emplvlcapita = mean(emplvlcapita, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(treated = as.integer(treated))

kansas_annual |>
  glimpse()
```

What to discuss:

- aggregation is part of design, not only convenience
- when you collapse data, say clearly what information is being smoothed away
- the tradeoff here is readability versus finer-grained timing

## Step 3: Inspect The Raw Kansas Trend

```{r fig.height=5.5}
kansas_annual |>
  mutate(group = if_else(state == "Kansas", "Kansas", "Donor pool")) |>
  group_by(group, year) |>
  summarise(lngdpcapita = mean(lngdpcapita, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = lngdpcapita, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = 2012, linetype = 2, color = "gray40") +
  labs(
    x = NULL,
    y = "Log GDP per capita",
    color = NULL
  ) +
  theme_minimal(base_size = 12)
```

Checkpoint:

- This is only a benchmark plot.
- The key question is whether any convex combination of donors can reconstruct Kansas well enough before `2012`.

## Step 4: Fit A Classical SCM Benchmark

Use `augsynth` itself to fit the classical benchmark by turning the outcome model off.

```{r}
kansas_scm <- augsynth(
  lngdpcapita ~ treated | revenuepop + emplvlcapita,
  unit = state,
  time = year,
  t_int = 2012,
  data = kansas_annual,
  progfunc = "None",
  scm = TRUE
)

summary(kansas_scm)
```

Why start here:

- classical SCM remains the baseline
- if the baseline already fits well, the case for augmentation is weaker
- if the baseline fits poorly, that poor fit is the design problem augmentation is trying to address

## Step 5: Fit Ridge-Augmented SCM

```{r}
kansas_ridge <- augsynth(
  lngdpcapita ~ treated | revenuepop + emplvlcapita,
  unit = state,
  time = year,
  t_int = 2012,
  data = kansas_annual,
  progfunc = "Ridge",
  scm = TRUE
)

summary(kansas_ridge)
```

What changed:

- the SCM weighting logic remains the starting point
- the ridge outcome model is used to correct bias from imperfect pre-treatment fit
- this may require limited extrapolation away from the donor convex hull

## Step 6: Compare Pre-Treatment Fit Visually

```{r fig.height=5.7}
plot(kansas_scm)
```

```{r fig.height=5.7}
plot(kansas_ridge)
```

Checkpoint:

- Does ridge augmentation visibly improve the pre-treatment path?
- If yes, ask whether that improvement looks modest and disciplined or aggressive and opaque.

## Step 7: Compare The Two Specifications Side By Side

```{r}
model_summaries <- list(
  classical_scm = summary(kansas_scm),
  ridge_ascm = summary(kansas_ridge)
)

model_summaries
```

Use this step to compare:

- pre-treatment fit quality
- implied post-treatment effect direction
- how much additional modeling you had to buy in order to improve fit

## Step 8: Add A Placebo-In-Time Style Stress Test

The advanced design question is not only whether ASCM fits better. It is whether it creates treatment-timed divergence only when treatment is plausibly active.

```{r}
kansas_backdate <- augsynth(
  lngdpcapita ~ treated | revenuepop + emplvlcapita,
  unit = state,
  time = year,
  t_int = 2008,
  data = kansas_annual |>
    mutate(treated = if_else(state == "Kansas" & year >= 2008, 1L, 0L)),
  progfunc = "Ridge",
  scm = TRUE
)

summary(kansas_backdate)
plot(kansas_backdate)
```

How to read it:

- a large placebo-in-time effect would suggest instability or overfitting
- a near-zero placebo effect does not prove the design, but it is useful reassurance when fit is otherwise credible

## Step 9: Write The Design Judgment

Use the outputs above to answer:

1. Does classical SCM fit Kansas well enough to stand on its own?
2. What does ridge augmentation improve?
3. What extrapolation cost are you implicitly accepting when you move from classical SCM to ASCM?
4. Is this a case for classical SCM, augmented SCM, or a different design altogether?

## Final Prompt

Write a short memo with one of these conclusions:

- stay with classical SCM because fit is already good enough
- use augmented SCM because the comparative-case design is still credible but classical fit is too weak
- abandon the SCM family here because even augmentation would be doing too much work

## Next Step

After this lab, go back to the [Synthetic Control slides](../slides/scm.qmd) and [SCM presenter script](../slides/scm_script.qmd) to consolidate the bigger design lesson: SCM credibility comes from donor support, pre-treatment fit, and transparent diagnostics before it comes from any single package call.
