---
title: "Lab: Proposition 99 With tidysynth"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as the first full SCM application lab.

- Keep the [Synthetic Control](../notes/scm/synthetic-control.qmd) overview open in another tab.
- Use the [California Proposition 99 notes](../notes/scm/california-tobacco-synthetic-control-notes.qmd) when you want the paper-level rationale behind each design step.
- Do not rush to the post-treatment plot. The donor pool, predictors, and pre-treatment fit are the actual argument.
- End with a design judgment about whether California's synthetic comparison looks credible enough to interpret.

::: {.callout-note}
The code is shown but not executed when the site is rendered. That keeps the page readable while preserving a runnable workflow for teaching and self-study.
:::

## Training Goal

Turn the canonical Proposition 99 application into a repeatable `tidysynth` workflow:

1. define the treated unit and intervention date
2. build a donor pool that tries to stay plausibly untreated
3. choose predictors and lagged outcomes that discipline pre-treatment fit
4. inspect donor weights, balance, path fit, and placebo-style comparisons
5. explain why this is a design-and-diagnostics workflow rather than just a nice graph

## Dataset At A Glance

This lab uses `tidysynth::smoking`, the bundled state-year panel tied to California's tobacco-control program.

- Treated unit: California
- Intervention anchor: `1988` in the teaching implementation, so `1989` onward is interpreted as the post-treatment period
- Outcome: per-capita cigarette sales (`cigsale`)
- Core predictors: income, cigarette price, beer consumption, age structure, and lagged smoking outcomes
- Main value: this is the cleanest benchmark case for teaching a full end-to-end SCM workflow in a tidy grammar

## What To Hand Back

By the end of the lab, you should be able to report:

- which states were excluded before estimation and why
- which predictors anchor the fit exercise
- which donor states receive positive or near-positive weight
- whether California's pre-treatment fit is strong enough to make the `1989` onward gap meaningful
- whether the placebo diagnostics reinforce the result or only partially reassure you

## Step 1: Load Packages And Inspect The Panel

```{r}
required_packages <- c(
  "tidysynth",
  "dplyr",
  "ggplot2",
  "tibble",
  "purrr"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

data("smoking", package = "tidysynth")

smoking |>
  glimpse()
```

Checkpoint:

- The data are already in tidy long format.
- Every row is a state-year observation, which makes the `tidysynth` pipeline easier to read than the original matrix workflow.

## Step 2: Define The Intervention And Donor Exclusions

Following the canonical Proposition 99 design, exclude states that adopted large tobacco-control programs during the post period, states with very large cigarette-tax increases, and the District of Columbia.

```{r}
treated_state <- "California"
intervention_year <- 1988

excluded_states <- c(
  "Alaska",
  "Hawaii",
  "Maryland",
  "Massachusetts",
  "Michigan",
  "New Jersey",
  "New York",
  "Washington",
  "District of Columbia"
)

analysis_data <- smoking |>
  filter(year <= 2000) |>
  filter(!state %in% excluded_states)

analysis_data |>
  count(state, sort = TRUE)
```

What to discuss:

- donor-pool construction is part of identification, not housekeeping
- the point is not to find the biggest donor pool, but the most defensible untreated comparison set
- if contamination remains plausible even after exclusions, that is a design limitation you should say out loud

## Step 3: Inspect The Raw California Trend

```{r fig.height=5.5}
analysis_data |>
  mutate(group = if_else(state == treated_state, "California", "Donor pool")) |>
  group_by(group, year) |>
  summarise(cigsale = mean(cigsale, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = cigsale, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = intervention_year, linetype = 2, color = "gray40") +
  labs(
    x = NULL,
    y = "Per-capita cigarette sales",
    color = NULL
  ) +
  theme_minimal(base_size = 12)
```

Checkpoint:

- A simple donor-pool average is not the SCM counterfactual.
- This first plot is only a benchmark for why a weighted synthetic comparison is needed.

## Step 4: Build The Synthetic-Control Object

```{r}
smoking_out <- analysis_data |>
  synthetic_control(
    outcome = cigsale,
    unit = state,
    time = year,
    i_unit = treated_state,
    i_time = intervention_year,
    generate_placebos = TRUE
  ) |>
  generate_predictor(
    time_window = 1980:1988,
    lnincome = mean(lnincome, na.rm = TRUE),
    retprice = mean(retprice, na.rm = TRUE),
    age15to24 = mean(age15to24, na.rm = TRUE)
  ) |>
  generate_predictor(
    time_window = 1984:1988,
    beer = mean(beer, na.rm = TRUE)
  ) |>
  generate_predictor(
    time_window = 1975,
    cigsale_1975 = cigsale
  ) |>
  generate_predictor(
    time_window = 1980,
    cigsale_1980 = cigsale
  ) |>
  generate_predictor(
    time_window = 1988,
    cigsale_1988 = cigsale
  ) |>
  generate_weights(
    optimization_window = 1970:1988,
    Margin.ipop = 0.02,
    Sigf.ipop = 7,
    Bound.ipop = 6
  ) |>
  generate_control()
```

This pipeline does the same design work as the classical `Synth` setup, but in a grammar that is easier to teach line by line.

## Step 5: Inspect Balance Before You Look At Effects

```{r}
balance_tbl <- smoking_out |>
  grab_balance_table()

balance_tbl
```

What to look for:

- the synthetic California column should sit close to observed California on the key predictors
- the donor-pool average is the benchmark you are trying to beat
- weak lagged-outcome balance is a warning sign because those lags are carrying a lot of the design discipline

## Step 6: Inspect Donor Weights

```{r}
weights_tbl <- smoking_out |>
  grab_unit_weights() |>
  arrange(desc(weight))

weights_tbl
```

Checkpoint:

- donor weights are part of the result, not an appendix
- if the synthetic control is built from a small set of plausible states, the counterfactual becomes easier to inspect and defend

## Step 7: Plot Observed And Synthetic California

```{r fig.height=5.7}
smoking_out |>
  plot_trends()
```

```{r fig.height=5.7}
smoking_out |>
  plot_differences()
```

What to discuss:

- the trend plot asks whether California and synthetic California line up closely through `1988`
- the difference plot asks whether the post-treatment divergence is both visible and treatment-timed
- a dramatic post-treatment gap is not enough if the pre-treatment fit is loose

## Step 8: Make The Gap Series Inspectable

```{r}
gap_tbl <- smoking_out |>
  grab_synthetic_control() |>
  transmute(
    year = time_unit,
    observed = real_y,
    synthetic = synth_y,
    gap = observed - synthetic
  )

gap_tbl
```

This is a useful discipline step. Never rely only on the plot when you can inspect the actual treated-versus-synthetic series directly.

## Step 9: Add Placebo-By-Unit Diagnostics

```{r fig.height=6}
smoking_out |>
  plot_placebos()
```

```{r}
mspe_tbl <- smoking_out |>
  grab_significance()

mspe_tbl
```

How to read this:

- placebos ask whether California's post-treatment divergence looks unusually large relative to donor states reassigned as if they were treated
- this is diagnostic reassurance, not a substitute for a credible donor pool
- if many placebo units fit terribly before treatment, interpret the ranking with caution

## Step 10: Write The Design Judgment

Use the outputs above to answer:

1. Why is California better suited to SCM than a simple treated-versus-rest-of-US comparison?
2. Which donor states actually build synthetic California?
3. Is the pre-treatment fit strong enough that the `1989` onward gap is worth interpreting?
4. Do the placebo diagnostics strengthen the case, or do they mainly show how much the result depends on fit quality?

## Next Step

Move next to the [augmented SCM lab](synthetic-control-augmentation-lab.qmd), where the main question shifts from "how do I run classical SCM?" to "what should I do when classical fit is visibly weak?"
