---
title: "Lab: Proposition 99 With Transparent SCM"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as the first full SCM application lab.

<!-- lab-downloads:start -->
::: {.resource-links}
[Quarto document](downloads/synthetic-control-proposition-99-lab.qmd){download="synthetic-control-proposition-99-lab.qmd"}
[R Jupyter notebook](notebooks/synthetic-control-proposition-99-lab.ipynb){download="synthetic-control-proposition-99-lab.ipynb"}
[Complete lab ZIP](downloads/synthetic-control-proposition-99-lab.zip){download="synthetic-control-proposition-99-lab.zip"}
[Data and setup](data.qmd#synthetic-control-proposition-99-lab)
:::
<!-- lab-downloads:end -->

- Keep the [Synthetic Control](../notes/scm/synthetic-control.qmd) overview open in another tab.
- Use the [California Proposition 99 notes](../notes/scm/california-tobacco-synthetic-control-notes.qmd) when you want the paper-level rationale behind each design step.
- Do not rush to the post-treatment plot. The donor pool, predictors, and pre-treatment fit are the actual argument.
- End with a design judgment about whether California's synthetic comparison looks credible enough to interpret.

::: {.callout-note}
The code is shown but not executed when the site is rendered. That keeps the page readable while preserving a runnable workflow for teaching and self-study.
:::

## Training Goal

Turn the canonical Proposition 99 application into a repeatable SCM workflow using explicit matrices and base-R optimisation:

1. define the treated unit and intervention date
2. build a donor pool that tries to stay plausibly untreated
3. choose predictors and lagged outcomes that discipline pre-treatment fit
4. inspect donor weights, balance, path fit, and placebo-style comparisons
5. explain why this is a design-and-diagnostics workflow rather than just a nice graph

## Dataset At A Glance

This lab uses a checksummed snapshot of the `smoking` state-year panel tied to California's tobacco-control program. The data are bundled, so the package that originally distributed them is not required at runtime.

- Treated unit: California
- Intervention anchor: `1988` in the teaching implementation, so `1989` onward is interpreted as the post-treatment period
- Outcome: per-capita cigarette sales (`cigsale`)
- Core predictors: income, cigarette price, beer consumption, age structure, and lagged smoking outcomes
- Main value: this is the cleanest benchmark case for teaching a full end-to-end SCM workflow while keeping the optimisation visible

## What To Hand Back

By the end of the lab, you should be able to report:

- which states were excluded before estimation and why
- which predictors anchor the fit exercise
- which donor states receive positive or near-positive weight
- whether California's pre-treatment fit is strong enough to make the `1989` onward gap meaningful
- whether the placebo diagnostics reinforce the result or only partially reassure you

## Step 1: Load Packages And Inspect The Panel

```{r}
data_helpers <- c("data/load-data.R", "../data/load-data.R", "docs/labs/data/load-data.R")
data_helpers <- data_helpers[file.exists(data_helpers)]
if (!length(data_helpers)) stop("Extract the complete lab ZIP, including its data folder, before running.")
source(data_helpers[[1]])
```

```{r}
required_packages <- c(
  "dplyr",
  "ggplot2",
  "tibble",
  "tidyr"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  stop("Install the documented R environment first; missing: ", paste(missing_packages, collapse=", "), call.=FALSE)
}

invisible(lapply(required_packages, library, character.only = TRUE))

smoking <- qed_data("smoking")

smoking |>
  glimpse()
```

Checkpoint:

- The data are already in tidy long format.
- Every row is a state-year observation, which makes the construction of predictor and outcome matrices easy to audit.

## Step 2: Define The Intervention And Donor Exclusions

Following the canonical Proposition 99 design, exclude states that adopted large tobacco-control programs during the post period, states with very large cigarette-tax increases, and the District of Columbia.

```{r}
treated_state <- "California"
intervention_year <- 1988

excluded_states <- c(
  "Alaska",
  "Hawaii",
  "Maryland",
  "Massachusetts",
  "Michigan",
  "New Jersey",
  "New York",
  "Washington",
  "District of Columbia"
)

analysis_data <- smoking |>
  filter(year <= 2000) |>
  filter(!state %in% excluded_states)

analysis_data |>
  count(state, sort = TRUE)
```

What to discuss:

- donor-pool construction is part of identification, not housekeeping
- the point is not to find the biggest donor pool, but the most defensible untreated comparison set
- if contamination remains plausible even after exclusions, that is a design limitation you should say out loud

## Step 3: Inspect The Raw California Trend

```{r fig.height=5.5}
analysis_data |>
  mutate(group = if_else(state == treated_state, "California", "Donor pool")) |>
  group_by(group, year) |>
  summarise(cigsale = mean(cigsale, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = cigsale, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = intervention_year, linetype = 2, color = "gray40") +
  labs(
    x = NULL,
    y = "Per-capita cigarette sales",
    color = NULL
  ) +
  theme_minimal(base_size = 12)
```

Checkpoint:

- A simple donor-pool average is not the SCM counterfactual.
- This first plot is only a benchmark for why a weighted synthetic comparison is needed.

## Step 4: Build The Synthetic-Control Design

```{r}
pre_years <- 1970:1988
plot_years <- 1970:2000
donor_states <- setdiff(unique(analysis_data$state), treated_state)

unit_predictors <- function(state_name, data = analysis_data) {
  unit_data <- data |>
    filter(state == state_name)

  c(
    lnincome = mean(unit_data$lnincome[unit_data$year %in% 1980:1988], na.rm = TRUE),
    retprice = mean(unit_data$retprice[unit_data$year %in% 1980:1988], na.rm = TRUE),
    age15to24 = mean(unit_data$age15to24[unit_data$year %in% 1980:1988], na.rm = TRUE),
    beer = mean(unit_data$beer[unit_data$year %in% 1984:1988], na.rm = TRUE),
    cigsale_1975 = unit_data$cigsale[match(1975, unit_data$year)],
    cigsale_1980 = unit_data$cigsale[match(1980, unit_data$year)],
    cigsale_1988 = unit_data$cigsale[match(1988, unit_data$year)]
  )
}

unit_outcomes <- function(state_name, years, data = analysis_data) {
  unit_data <- data |>
    filter(state == state_name)
  unit_data$cigsale[match(years, unit_data$year)]
}

simplex_weights <- function(theta) c(theta, 1 - sum(theta))

fit_scm <- function(target_state, comparison_states, data = analysis_data) {
  X1 <- unit_predictors(target_state, data)
  X0 <- vapply(comparison_states, unit_predictors, numeric(length(X1)), data = data)
  Z1 <- unit_outcomes(target_state, pre_years, data)
  Z0 <- vapply(
    comparison_states,
    unit_outcomes,
    numeric(length(pre_years)),
    years = pre_years,
    data = data
  )

  design_target <- c(X1, Z1)
  design_donors <- rbind(X0, Z0)
  row_scale <- apply(cbind(design_target, design_donors), 1, sd)
  row_scale[!is.finite(row_scale) | row_scale == 0] <- 1
  scaled_target <- design_target / row_scale
  scaled_donors <- design_donors / row_scale

  loss <- function(theta) {
    weights <- simplex_weights(theta)
    mean((scaled_target - as.numeric(scaled_donors %*% weights))^2)
  }

  gradient <- function(theta) {
    weights <- simplex_weights(theta)
    residual <- scaled_target - as.numeric(scaled_donors %*% weights)
    donor_differences <- scaled_donors[, -ncol(scaled_donors), drop = FALSE] -
      scaled_donors[, ncol(scaled_donors)]
    -2 * colMeans(donor_differences * residual)
  }

  constraint_matrix <- rbind(
    diag(length(comparison_states) - 1),
    rep(-1, length(comparison_states) - 1)
  )
  constraint_boundary <- c(rep(0, length(comparison_states) - 1), -1)

  fit <- constrOptim(
    theta = rep(1 / length(comparison_states), length(comparison_states) - 1),
    f = loss,
    grad = gradient,
    ui = constraint_matrix,
    ci = constraint_boundary,
    method = "BFGS",
    control = list(maxit = 5000, reltol = 1e-11)
  )
  weights <- simplex_weights(fit$par)

  stopifnot(all(weights >= 0), abs(sum(weights) - 1) < 1e-10)
  list(
    weights = weights,
    comparison_states = comparison_states,
    predictors = X1,
    donor_predictors = X0,
    convergence = fit$convergence,
    objective = fit$value
  )
}

california_fit <- fit_scm(treated_state, donor_states)
```

The helper keeps the design work visible. Base R's `constrOptim()` constrains donor weights to be non-negative and sum to one while minimizing standardized predictor and pre-treatment outcome mismatch.

## Step 5: Inspect Balance Before You Look At Effects

```{r}
balance_tbl <- tibble(
  predictor = names(california_fit$predictors),
  California = california_fit$predictors,
  synthetic = as.numeric(california_fit$donor_predictors %*% california_fit$weights),
  donor_average = rowMeans(california_fit$donor_predictors)
)

balance_tbl
```

What to look for:

- the synthetic California column should sit close to observed California on the key predictors
- the donor-pool average is the benchmark you are trying to beat
- weak lagged-outcome balance is a warning sign because those lags are carrying a lot of the design discipline

## Step 6: Inspect Donor Weights

```{r}
weights_tbl <- tibble(
  state = california_fit$comparison_states,
  weight = california_fit$weights
) |>
  arrange(desc(weight))

weights_tbl
```

Checkpoint:

- donor weights are part of the result, not an appendix
- if the synthetic control is built from a small set of plausible states, the counterfactual becomes easier to inspect and defend

## Step 7: Plot Observed And Synthetic California

```{r fig.height=5.7}
treated_path <- unit_outcomes(treated_state, plot_years)
donor_paths <- vapply(
  donor_states,
  unit_outcomes,
  numeric(length(plot_years)),
  years = plot_years
)

gap_tbl <- tibble(
  year = plot_years,
  observed = treated_path,
  synthetic = as.numeric(donor_paths %*% california_fit$weights)
) |>
  mutate(gap = observed - synthetic)

gap_tbl |>
  select(year, observed, synthetic) |>
  pivot_longer(-year, names_to = "series", values_to = "cigsale") |>
  ggplot(aes(year, cigsale, colour = series)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = intervention_year, linetype = 2, colour = "gray40") +
  labs(x = NULL, y = "Per-capita cigarette sales", colour = NULL) +
  theme_minimal(base_size = 12)
```

```{r fig.height=5.7}
ggplot(gap_tbl, aes(year, gap)) +
  geom_hline(yintercept = 0, linetype = 3, colour = "gray50") +
  geom_vline(xintercept = intervention_year, linetype = 2, colour = "gray40") +
  geom_line(linewidth = 1) +
  labs(x = NULL, y = "California minus synthetic California") +
  theme_minimal(base_size = 12)
```

What to discuss:

- the trend plot asks whether California and synthetic California line up closely through `1988`
- the difference plot asks whether the post-treatment divergence is both visible and treatment-timed
- a dramatic post-treatment gap is not enough if the pre-treatment fit is loose

## Step 8: Make The Gap Series Inspectable

```{r}
gap_tbl
```

This is a useful discipline step. Never rely only on the plot when you can inspect the actual treated-versus-synthetic series directly.

## Step 9: Add Placebo-By-Unit Diagnostics

```{r fig.height=6}
placebo_paths <- lapply(donor_states, function(placebo_state) {
  placebo_donors <- setdiff(donor_states, placebo_state)
  placebo_fit <- fit_scm(placebo_state, placebo_donors)
  observed <- unit_outcomes(placebo_state, plot_years)
  donors <- vapply(
    placebo_donors,
    unit_outcomes,
    numeric(length(plot_years)),
    years = plot_years
  )
  tibble(
    state = placebo_state,
    year = plot_years,
    gap = observed - as.numeric(donors %*% placebo_fit$weights)
  )
})

placebo_gap_tbl <- bind_rows(placebo_paths)

ggplot(placebo_gap_tbl, aes(year, gap, group = state)) +
  geom_line(colour = "gray75", linewidth = 0.35) +
  geom_line(
    data = gap_tbl,
    aes(year, gap),
    inherit.aes = FALSE,
    colour = "firebrick",
    linewidth = 1.1
  ) +
  geom_hline(yintercept = 0, colour = "gray50") +
  geom_vline(xintercept = intervention_year, linetype = 2, colour = "gray40") +
  labs(x = NULL, y = "Treated minus synthetic") +
  theme_minimal(base_size = 12)
```

```{r}
mspe_tbl <- bind_rows(
  placebo_gap_tbl,
  gap_tbl |>
    transmute(state = treated_state, year, gap)
) |>
  group_by(state) |>
  summarise(
    pre_mspe = mean(gap[year <= intervention_year]^2),
    post_mspe = mean(gap[year > intervention_year]^2),
    mspe_ratio = post_mspe / pre_mspe,
    .groups = "drop"
  ) |>
  arrange(desc(mspe_ratio))

mspe_tbl
```

How to read this:

- placebos ask whether California's post-treatment divergence looks unusually large relative to donor states reassigned as if they were treated
- this is diagnostic reassurance, not a substitute for a credible donor pool
- if many placebo units fit terribly before treatment, interpret the ranking with caution

## Step 10: Write The Design Judgment

Use the outputs above to answer:

1. Why is California better suited to SCM than a simple treated-versus-rest-of-US comparison?
2. Which donor states actually build synthetic California?
3. Is the pre-treatment fit strong enough that the `1989` onward gap is worth interpreting?
4. Do the placebo diagnostics strengthen the case, or do they mainly show how much the result depends on fit quality?

## Next Step

Move next to the [augmented SCM lab](synthetic-control-augmentation-lab.qmd), where the main question shifts from "how do I run classical SCM?" to "what should I do when classical fit is visibly weak?"
