---
title: "Lab: Synthetic Control Mechanics"
format:
  html: default
execute:
  echo: true
  eval: false
---

## How To Use This Page

Use this page as the first hands-on SCM lab.

- Keep the [Synthetic Control](../notes/scm/synthetic-control.qmd) overview open in another tab.
- Use this lab to understand the grammar of classical SCM before you move to real case studies.
- Read `dataprep()` as part of the design, not as data wrangling boilerplate.
- End by explaining what the optimizer did and whether the fit looks credible enough to learn from.

::: {.callout-note}
The code is shown but not executed when the site is rendered. That keeps the page readable while preserving a runnable workflow for teaching and self-study.
:::

## Training Goal

Use `Synth::synth.data` to make the mechanics of classical SCM visible:

1. define the treated unit and donor pool explicitly
2. encode the design with `dataprep()`
3. estimate the synthetic control with `synth()`
4. inspect donor weights `W` and predictor weights `V`
5. interpret the path plot and gap plot without pretending the toy data prove a substantive claim

## Dataset At A Glance

This lab uses `Synth::synth.data`, the package's toy panel for teaching the original workflow.

- Purpose: mechanics first, not substantive interpretation
- Data shape: a small panel with one treated unit and a compact donor pool
- Main value: it isolates the optimizer, the matrices, and the output objects without the extra complexity of a real policy application
- Main limit: the dataset is too artificial to teach donor contamination, spillovers, or serious placebo logic

## What To Hand Back

By the end of the lab, you should be able to report:

- which unit is treated and which units enter the donor pool
- what `dataprep()` is doing to produce `X1`, `X0`, `Z1`, and `Z0`
- which donors receive positive weight in `solution.w`
- which predictors or lagged outcomes receive the most weight in `solution.v`
- whether the pre-treatment fit looks good enough to trust the toy example as a demonstration of mechanics

## Step 1: Load Packages And Inspect The Toy Panel

```{r}
required_packages <- c(
  "Synth",
  "dplyr",
  "tibble"
)

missing_packages <- required_packages[!vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

data("synth.data", package = "Synth")

synth.data |>
  tibble::as_tibble() |>
  glimpse()
```

Checkpoint:

- Can you see the unit identifier, time variable, predictors, and outcome?
- Which columns are design inputs and which columns are only labels?

## Step 2: Define The Treated Unit And Donor Pool

The point of this step is not software convenience. It is to make the comparison explicit before any optimization happens.

```{r}
treated_unit <- 7

donor_units <- c(29, 2, 13, 17, 32, 38)

unit_lookup <- synth.data |>
  distinct(unit.num, name) |>
  arrange(unit.num)

treated_name <- unit_lookup |>
  filter(unit.num == treated_unit)

donor_lookup <- unit_lookup |>
  filter(unit.num %in% donor_units)

treated_name
donor_lookup
```

What to discuss:

- SCM does not remove design judgment; it makes it legible.
- Even in a toy example, you should know who is eligible to act as a donor before you estimate weights.

## Step 3: Encode The Design With `dataprep()`

`dataprep()` is the design object in the original `Synth` workflow. It decides which predictors matter, which pre-treatment periods define fit, and which periods will be plotted later.

```{r}
dataprep.out <- dataprep(
  foo = synth.data,
  predictors = c("X1", "X2", "X3"),
  predictors.op = "mean",
  dependent = "Y",
  unit.variable = "unit.num",
  time.variable = "year",
  special.predictors = list(
    list("Y", 1991, "mean"),
    list("Y", 1985, "mean"),
    list("Y", 1980, "mean")
  ),
  treatment.identifier = treated_unit,
  controls.identifier = donor_units,
  time.predictors.prior = 1984:1989,
  time.optimize.ssr = 1984:1990,
  unit.names.variable = "name",
  time.plot = 1984:1996
)

names(dataprep.out)
```

## Step 4: Inspect The Objects That `dataprep()` Built

```{r}
dim(dataprep.out$X1)
dim(dataprep.out$X0)
dim(dataprep.out$Z1)
dim(dataprep.out$Z0)

dataprep.out$X1
dataprep.out$X0

dataprep.out$tag
```

Checkpoint:

- `X1` and `X0` contain the predictor targets for the treated unit and donors.
- `Z1` and `Z0` contain the pre-treatment outcome history used to optimize fit.
- `tag` is worth reading because it records the design choices you just made.

## Step 5: Estimate The Synthetic Control

```{r}
synth.out <- synth(dataprep.out)

names(synth.out)
```

This is the estimator step. The constrained optimizer now chooses donor weights `W` and predictor weights `V` given the design encoded above.

## Step 6: Read Donor Weights `W` And Predictor Weights `V`

```{r}
donor_weights <- tibble(
  unit.num = donor_units,
  weight = as.numeric(synth.out$solution.w)
) |>
  left_join(unit_lookup, by = "unit.num") |>
  arrange(desc(weight))

predictor_weights <- tibble(
  predictor = rownames(dataprep.out$X1),
  v_weight = as.numeric(synth.out$solution.v[, 1])
) |>
  arrange(desc(v_weight))

donor_weights
predictor_weights
```

What to look for:

- sparse `W` weights mean only a few donors carry the synthetic comparison
- large `V` weights show which predictors or lagged outcomes mattered most for fit
- neither object is decorative; both are part of the causal argument

## Step 7: Build The Standard Summary Tables

```{r}
synth.tables <- synth.tab(
  dataprep.res = dataprep.out,
  synth.res = synth.out
)

synth.tables$tab.pred
synth.tables$tab.w
synth.tables$tab.v
```

Checkpoint:

- Does the synthetic unit improve predictor balance relative to a simple donor average?
- Are the non-zero donor weights substantively legible or completely opaque?

## Step 8: Plot The Treated And Synthetic Paths

```{r fig.height=5.5}
path.plot(
  synth.res = synth.out,
  dataprep.res = dataprep.out,
  Ylab = "Outcome",
  Xlab = "Year",
  Legend = c("Treated unit", "Synthetic control"),
  Legend.position = "bottomright"
)
abline(v = 1990, lty = 2, col = "gray40")
```

```{r fig.height=5.5}
gaps.plot(
  synth.res = synth.out,
  dataprep.res = dataprep.out,
  Ylab = "Treated minus synthetic",
  Xlab = "Year"
)
abline(v = 1990, lty = 2, col = "gray40")
abline(h = 0, lty = 3, col = "gray50")
```

What to discuss:

- the path plot asks whether the synthetic unit tracks the treated unit before treatment
- the gap plot turns that same information into a treatment-effect style display
- if the pre-treatment gap is wide, the right lesson is usually "design problem" rather than "interesting effect"

## Step 9: Calculate The Gap Series Directly

```{r}
gap_tbl <- tibble(
  year = dataprep.out$tag$time.plot,
  treated = as.numeric(dataprep.out$Y1plot),
  synthetic = as.numeric(dataprep.out$Y0plot %*% synth.out$solution.w)
) |>
  mutate(gap = treated - synthetic)

gap_tbl
```

This step matters because it makes the plotted divergence inspectable as ordinary data rather than as an opaque graphics side effect.

## Final Prompt

Write a short answer to these questions:

1. What part of SCM became clearer once you saw `dataprep()` and `synth()` as separate steps?
2. Which donors actually build the synthetic unit?
3. Does this toy example teach you substantive causal inference, or mainly the structure of the estimator?

## Next Step

Move next to the [Proposition 99 lab](synthetic-control-proposition-99-lab.qmd), where the same workflow becomes a real comparative case study with donor exclusions, placebo logic, and stronger design stakes.
