---
title: "How to Do Interrupted Time Series"
subtitle: "From an intervention timeline to a defensible counterfactual in R"
format: html
execute:
  echo: true
  warning: false
  message: false
---

Interrupted time series (ITS) compares an outcome after an intervention with the
continuation of its earlier pattern. It is useful when a policy affects a whole
population at a known time and a credible untreated comparison group is absent.
The challenge is not drawing two regression lines. It is defending why the
earlier pattern tells us what would have happened without the intervention.

This guide takes an applied evaluator through a monthly series, a segmented
model, serial dependence, effect contrasts, and a reporting decision. Basic
regression and R data frames are assumed. All data come from the repository's
[local data collection](../../labs/data.qmd#seatbelts); the figures and tables
below are computed from that snapshot. For downloadable exercises, start with
the [mechanics lab](../../labs/interrupted-time-series-mechanics-lab.qmd).

## 1. Decide whether the design fits the question

Write a short design statement before choosing a model. Specify the population,
outcome, intervention, implementation date, observation interval, and the effect
horizon that matters for the decision. A programme can change the outcome level,
its rate of change, both, or neither. The expected shape should follow a mechanism,
not whichever specification produces the smallest p-value [@wagner2002segmented;
@lopezbernal2017tutorial].

Check whether the intervention was actually abrupt. An announcement can precede
implementation; compliance may rise gradually; a reporting change may take effect
on another date. A single step indicator is inappropriate when it compresses a
long implementation process into a fictitious overnight change. Record these
dates even when the primary specification uses one interruption.

In this example, the UK's front-seat belt requirement took effect on 31 January
1983. February 1983 is the first treated **monthly** observation. The outcome is
front-seat passengers killed or seriously injured, not deaths alone. It is a
count, not a risk per journey. The question is whether its path changed at the
interruption, and whether that change can plausibly be attributed to the policy
[@rseatbeltsdata; @harvey1986seatbelts].

The core identifying assumption is that, absent the policy, the pre-intervention
outcome process would have continued in the way represented by the model. That
requires more than a good fit: no unaccounted concurrent shock, stable measurement,
and a population/exposure process compatible with the proposed counterfactual.
With one interrupted series, the model cannot distinguish two simultaneous causes
solely because one is the intervention of interest.

## 2. Load, audit, and plot the series

```{r}
source("../../labs/data/load-data.R")
library(ggplot2)
library(nlme)
Seatbelts <- qed_data("Seatbelts")
dat <- as.data.frame(Seatbelts)
dat$date <- seq(as.Date("1969-01-01"), by="month", length.out=nrow(dat))
dat$index <- seq_len(nrow(dat))
first_treated <- which(dat$law == 1)[1]
dat$time <- dat$index - first_treated
dat$post_time <- pmax(0, dat$time)
dat$season_sin <- sin(2*pi*dat$index/12)
dat$season_cos <- cos(2*pi*dat$index/12)
dat$log_front <- log(dat$front)
stopifnot(nrow(dat) == 192, all(dat$front > 0),
          dat$date[first_treated] == as.Date("1983-02-01"),
          sum(dat$law == 0) == 169, sum(dat$law == 1) == 23)
data.frame(start=min(dat$date), end=max(dat$date),
           pre_months=sum(dat$law == 0), post_months=sum(dat$law == 1))
```

```{r}
ggplot(dat, aes(date, front)) +
  geom_line(color="#246c85", linewidth=0.6) +
  geom_vline(xintercept=as.Date("1983-02-01"), linetype=2) +
  labs(x=NULL, y="Front-seat passengers killed or seriously injured") +
  theme_minimal(base_size=12)
```

Look for missing periods, abrupt recording breaks, outliers, seasonal cycles,
and changes in the variance. Reconcile suspicious observations against the data
dictionary and institutional history. Do not delete a large value just because
it makes a line less smooth. Here there are 169 pre-policy months but only 23
post-policy months: estimates of a new slope are informed by less than two years
after the intervention.

The exposure denominator matters. Fewer injuries can reflect less driving rather
than safer travel. An exposure-adjusted model would answer a different question
from a total-count model. Adding contemporaneous kilometres as a covariate is not
automatically a correction: ask whether exposure is a confounder, mediator, or
part of the policy effect you want to measure.

## 3. Encode the interruption explicitly

Use a centered time variable with zero in the first treated month. Let $D_t$ be
one after implementation and $P_t=\max(0,t)$ be months since the interruption.
Our starting model is

$$
\log(Y_t)=\beta_0+\beta_1t+\beta_2D_t+\beta_3P_t
 +\gamma_1\sin(2\pi m_t/12)+\gamma_2\cos(2\pi m_t/12)+\epsilon_t.
$$

Here $\beta_1$ is the pre-intervention log trend, $\beta_2$ the immediate log-level
change at month zero, and $\beta_3$ the change in monthly log slope. The post-policy
slope is $\beta_1+\beta_3$. At horizon $h$, the intervention contrast is
$\beta_2+h\beta_3$. If instead the first treated month had post-time equal to one,
the interpretation of the level coefficient would change. Check a few rows
around the interruption rather than trusting the variable name [@xiao2021parameterization].

```{r}
dat[(first_treated-2):(first_treated+2), c("date", "time", "law", "post_time")]
specification <- log_front ~ time + law + post_time + season_sin + season_cos
ols <- lm(specification, data=dat)
summary(ols)$coefficients
```

The log scale keeps fitted counts positive after back-transformation and makes
the intervention contrast multiplicative. It is not compulsory: a count model
with an exposure offset can be more natural when modelling rates. Choose the
scale to match the scientific question and residual behavior. A single harmonic
captures a smooth annual cycle; it does not establish that all seasonal structure
has been removed.

## 4. Inspect serial dependence and refit

Monthly outcomes are rarely independent. Persistent residuals can make ordinary
standard errors too optimistic and can signal an inadequate trend or seasonal
model. Inspect the residual autocorrelation function rather than applying an AR(1)
model by ritual.

```{r}
acf(residuals(ols), main="OLS residual autocorrelation")
fit <- gls(specification, data=dat,
           correlation=corAR1(form=~index), method="REML")
summary(fit)$tTable
plot(ACF(fit, resType="normalized"), alpha=0.05)
```

AR(1) says that residual association declines geometrically with the number of
months separating observations. It is a working error model, not an identification
strategy. Examine the normalized residuals after fitting: remaining seasonal
peaks, trend, or changing variance warrant investigation. Do not keep adding
parameters until every diagnostic looks reassuring. A finite sample can have
nonzero residual correlations even under an adequate model, while a misspecified
mean can masquerade as serial dependence.

GLS here changes estimation and uncertainty relative to OLS. Its intervals still
condition on the model family, intervention coding, and absence of an omitted
concurrent cause. They do not quantify all the uncertainty in the causal design.

## 5. Estimate effects at decision-relevant horizons

Report an immediate contrast and a small set of substantively chosen horizons,
not only a slope-change p-value. The covariance between the level and slope
coefficients is essential:

$$
\operatorname{Var}(\hat\beta_2+h\hat\beta_3)
=V_{22}+2hV_{23}+h^2V_{33}.
$$

```{r}
b <- coef(fit)
V <- vcov(fit)
horizons <- c(0, 6, 12)
effects <- do.call(rbind, lapply(horizons, function(h) {
  L <- setNames(rep(0, length(b)), names(b))
  L["law"] <- 1
  L["post_time"] <- h
  estimate <- sum(L*b)
  se <- sqrt(drop(t(L) %*% V %*% L))
  data.frame(months=h, log_contrast=estimate, se=se,
             percent=100*expm1(estimate),
             lower=100*expm1(estimate-1.96*se),
             upper=100*expm1(estimate+1.96*se))
}))
knitr::kable(effects, digits=3)
```

These are approximate model-based 95% intervals using a normal critical value.
The percentage transformation describes a contrast on the exponentiated expected
log scale. Without an additional retransformation argument, it is not automatically
the percentage effect on the arithmetic mean number of casualties. Avoid claiming
an exact number of injuries prevented by subtracting two exponentiated log fits.

## 6. Display the counterfactual you actually estimated

```{r}
counterfactual <- dat
counterfactual$law <- 0
counterfactual$post_time <- 0
dat$fitted_log <- as.numeric(predict(fit))
dat$counterfactual_log <- as.numeric(predict(fit, newdata=counterfactual))
ggplot(dat, aes(date, log_front)) +
  geom_line(color="grey65", linewidth=0.4) +
  geom_line(aes(y=fitted_log, color="Segmented fit"), linewidth=0.8) +
  geom_line(aes(y=counterfactual_log, color="No-intervention terms"),
            linewidth=0.8, linetype=2) +
  geom_vline(xintercept=as.Date("1983-02-01"), linetype=3) +
  scale_color_manual(values=c("No-intervention terms"="#177b72", "Segmented fit"="#245ca4")) +
  labs(x=NULL, y="Log front-seat casualties", color=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
```

This counterfactual switches off the intervention terms in a model estimated on
the **whole** series. It is not an out-of-sample forecast from a pre-policy-only
fit. Those two approaches use different information and answer different
validation questions. The [counterfactual-validation lab](../../labs/interrupted-time-series-counterfactual-validation-lab.qmd)
shows how to select forecasting specifications using only untreated periods and
compare them at rolling origins. A good historical forecasting score supports a
model's predictive usefulness, not the claim that the post-policy world contained
no other changes.

## 7. Stress-test the substantive conclusion

Consider alternative effect shapes and seasonal controls because they represent
different claims about how the programme works. Compare like-for-like horizons.
For likelihood comparisons of models with different fixed effects use maximum
likelihood rather than comparing REML likelihoods.

```{r}
dat$month <- factor(format(dat$date, "%m"))
level_only <- gls(log_front ~ time + law + season_sin + season_cos,
                  data=dat, correlation=corAR1(form=~index), method="ML")
level_slope <- update(fit, method="ML")
monthly <- gls(log_front ~ time + law + post_time + month,
               data=dat, correlation=corAR1(form=~index), method="ML")
data.frame(model=c("Level only", "Level and slope", "Month indicators"),
           AIC=c(AIC(level_only), AIC(level_slope), AIC(monthly)))
```

Use the result to expose model dependence, not to select a causal explanation by
AIC alone. Also consider defensible pre-period lengths, a plausible implementation
lag, outlier sensitivity, and whether the conclusion depends on a handful of
post-policy observations. A placebo interruption can reveal an unstable model,
but testing many dates and retaining the least troublesome one is not a valid
robustness strategy.

For this application, concurrent drink-driving enforcement and broader road-safety
changes limit attribution. Rear-seat passengers and van drivers are not automatically
untreated controls: behavior, exposure and spillovers can connect them to the
same intervention. A controlled ITS is useful only when the comparison outcome
has a defensible relationship with the treated counterfactual and no relevant
treatment exposure [@lopezbernal2018controls; @ukparliament1986seatbeltdebate].

## 8. Write a bounded conclusion

A useful report identifies the outcome scale and population, the 169/23 split,
the February 1983 coding, the trend/seasonal/error specifications, and effects with
intervals at the chosen horizons. Show both observed and counterfactual paths,
then state which alternative explanations the design cannot separate.

Do not report a causal effect merely because the interruption coefficient is
statistically significant. Conversely, an imprecise estimate does not demonstrate
that the intervention had no effect. Distinguish a change consistent with the
programme from a uniquely identified programme effect, and explain whether the
remaining uncertainty is tolerable for the decision.

Continue with the [design and diagnostics lab](../../labs/interrupted-time-series-design-diagnostics-lab.qmd)
and the [ITS overview](interrupted-time-series.qmd) for the wider source map.

## References
