---
title: "Interrupted Time Series"
execute:
  echo: true
  warning: false
  message: false
---

# What The Method Is

Interrupted time series (ITS) evaluates an intervention introduced at a known point in a regularly observed series. Instead of comparing one observation before treatment with one after treatment, it estimates the trajectory before the intervention and asks whether the level or trend changed when the intervention began [@wagner2002segmented; @lopezbernal2017tutorial].

The untreated counterfactual is the outcome path expected without the intervention. In a basic ITS that path is projected from the pre-intervention series after accounting for features such as trend and seasonality. The gap between the observed and projected paths is estimated, not observed, so a competing event at the same date can create the same pattern.

::: {.callout-note title="Teach And Practise The Method"}
Use the [Interrupted Time Series slides](../../slides/its.qmd) for the core workshop sequence. Continue with [ITS Foundations With Seatbelts](../../labs/interrupted-time-series-mechanics-lab.qmd), [Seatbelt Law Design and Robustness](../../labs/interrupted-time-series-design-diagnostics-lab.qmd), and [ITS Counterfactual Validation](../../labs/interrupted-time-series-counterfactual-validation-lab.qmd).
:::

# When To Use It

ITS is most useful when:

- an intervention, policy, shock, or service change has a defensible start date;
- the outcome is measured consistently at regular intervals;
- enough pre-intervention observations exist to characterize trend, seasonality, and dependence;
- enough post-intervention observations exist to estimate the expected effect shape; and
- randomization or a stronger comparison-based design is unavailable.

Typical applications include legislation, changes to clinical guidance, tax reforms, transport policies, and service redesigns. ITS is not a defensible label for one observation before and one after. It is also weak when the interruption date was chosen after inspecting the outcome, implementation was too gradual to locate in time, or measurement changed with the intervention [@cochrane2024robinsits].

# The Counterfactual And Estimand

A transparent segmented-regression model is

$$
Y_t = \beta_0 + \beta_1t + \beta_2I_t + \beta_3T^{post}_t
      + \text{seasonality}_t + \epsilon_t,
$$

where:

- $Y_t$ is the outcome at time $t$;
- $t$ is continuous calendar time;
- $I_t$ is zero before the intervention and one from the intervention onward;
- $T^{post}_t$ is zero before and at the first intervention observation, then counts elapsed periods;
- $\beta_2$ is the immediate level change; and
- $\beta_3$ is the change in trend after intervention.

At $h$ periods after intervention, the estimated difference from the no-intervention trajectory is

$$
\text{effect}(h) = \beta_2 + h\beta_3.
$$

Its uncertainty must include the covariance between the level- and trend-change estimates.

::: {.callout-warning title="Code post-intervention time from zero"}
Do not use calendar time multiplied by the intervention indicator as the post-intervention trend term. Use elapsed time since intervention, with zero in the first intervention period, so the level-change coefficient remains interpretable [@xiao2021parameterization].
:::

# Core Causal Conditions

## No Concurrent Outcome-Relevant Shock

Nothing else should change at the intervention date in a way that alters the outcome. This is the central threat in an uncontrolled ITS. Policy bundles, economic shocks, strikes, enforcement changes, and other reforms can all be mistaken for the intervention effect.

## A Credible Untreated Trajectory

The pre-intervention series must contain enough information to extrapolate. A linear trend is a model choice, not a property guaranteed by the design. Curvature, structural breaks, evolving seasonality, or a short baseline can make the projected path implausible.

## Stable Measurement And Composition

Outcome definitions, coding, coverage, reporting incentives, and the population represented should remain comparable over time. A new reporting system introduced with the policy is an interruption in measurement as well as policy.

## Pre-Specified Timing And Effect Shape

The interruption date and expected effect form should come from implementation knowledge. Anticipation, phased delivery, delayed effects, and temporary transition periods should be represented when they are plausible, not added after viewing results.

## A Credible Control, If Used

A controlled ITS needs an untreated series with a stable pre-intervention relationship to the treated outcome. The control must not be affected by the intervention, spillovers, or changes in composition. Correlation alone does not establish those conditions [@lopezbernal2018controls].

# Data And Design Checklist

Before modelling, record:

1. The intervention date, implementation period, and any anticipation window.
2. The unit represented by each observation and whether the same population is measured over time.
3. The outcome definition, denominator, frequency, and measurement process.
4. The pre- and post-intervention spans.
5. Expected baseline trend, seasonality, lag, and duration of effect.
6. Other events near the interruption date.
7. Potential comparison series and negative-control outcomes.
8. The primary estimand and a short prespecified sensitivity set.

There is no universal minimum number of observations. The requirement depends on noise, autocorrelation, seasonality, effect size, and model complexity. The practical question is whether the available series can support the proposed counterfactual [@cochrane2024robinsits].

# Worked Example: Compulsory Front-Seat Belt Wearing

R includes monthly Great Britain road-casualty series from January 1969 through December 1984. Compulsory front-seat belt wearing began on 31 January 1983, and the dataset's monthly `law` indicator begins in February 1983 [@rseatbeltsdata; @ukparliament2012seatbelts]. The original study used structural time-series models to assess the legislation [@harvey1986seatbelts].

The primary outcome here is `front`: front-seat passengers killed or seriously injured. It is a count rather than a fatality-only outcome or an exposure-adjusted rate.

```{r}
#| label: prepare-seatbelts
library(ggplot2)
library(nlme)

data("Seatbelts", package = "datasets")

seatbelts <- as.data.frame(Seatbelts)
seatbelts$time <- seq_len(nrow(seatbelts))
seatbelts$date <- seq(as.Date("1969-01-01"), by = "month", length.out = nrow(seatbelts))
seatbelts$post_time <- ifelse(
  seatbelts$law == 1,
  seatbelts$time - min(seatbelts$time[seatbelts$law == 1]),
  0
)
seatbelts$season_sin <- sin(2 * pi * seatbelts$time / 12)
seatbelts$season_cos <- cos(2 * pi * seatbelts$time / 12)
intervention_date <- min(seatbelts$date[seatbelts$law == 1])

stopifnot(
  nrow(seatbelts) == 192L,
  intervention_date == as.Date("1983-02-01"),
  seatbelts$post_time[seatbelts$date == intervention_date] == 0
)
```

Plot first. The graph reveals the long baseline, strong annual seasonality, an apparent post-law fall, and the much shorter post-law period.

```{r}
#| label: fig-seatbelts-raw
#| fig-cap: "Monthly front-seat passengers killed or seriously injured in Great Britain."
#| fig-height: 4.8
ggplot(seatbelts, aes(date, front)) +
  geom_line(linewidth = 0.6, colour = "#24527a") +
  geom_vline(xintercept = intervention_date, linetype = "dashed", colour = "#a23b3b") +
  labs(
    x = NULL,
    y = "Front-seat passengers killed or seriously injured",
    title = "Road casualties before and after compulsory belt wearing"
  ) +
  theme_minimal(base_size = 12)
```

## Fit The Segmented Models

Use a log outcome so intervention coefficients represent proportional changes. A sine/cosine pair represents the annual cycle. OLS makes the parameterization clear; generalized least squares then allows AR(1) residual dependence.

```{r}
#| label: fit-seatbelts-models
its_formula <- log(front) ~
  time + law + post_time + season_sin + season_cos

ols_fit <- lm(its_formula, data = seatbelts)

gls_fit <- gls(
  its_formula,
  data = seatbelts,
  correlation = corAR1(form = ~ time),
  method = "REML"
)

ols_terms <- coef(summary(ols_fit))[c("law", "post_time"), ]
gls_terms <- summary(gls_fit)$tTable[c("law", "post_time"), ]

model_comparison <- data.frame(
  Model = rep(c("OLS", "GLS with AR(1) errors"), each = 2),
  Term = rep(c("Immediate change", "Monthly trend change"), times = 2),
  Estimate = c(ols_terms[, "Estimate"], gls_terms[, "Value"]),
  `Standard error` = c(ols_terms[, "Std. Error"], gls_terms[, "Std.Error"]),
  check.names = FALSE
)

knitr::kable(model_comparison, digits = 3)
```

AR(1) errors address residual dependence. They do not remove a concurrent shock, choose the right intervention date, or turn an unsuitable comparison outcome into a control.

## Estimate Horizon-Specific Changes

```{r}
#| label: seatbelts-horizons
effect_at_horizon <- function(model, horizon) {
  weights <- c(law = 1, post_time = horizon)
  estimate <- sum(weights * coef(model)[names(weights)])
  covariance <- vcov(model)[names(weights), names(weights)]
  standard_error <- sqrt(drop(t(weights) %*% covariance %*% weights))

  data.frame(
    Horizon = horizon,
    `Percent change` = 100 * (exp(estimate) - 1),
    `Lower 95%` = 100 * (exp(estimate - 1.96 * standard_error) - 1),
    `Upper 95%` = 100 * (exp(estimate + 1.96 * standard_error) - 1),
    check.names = FALSE
  )
}

horizon_results <- do.call(
  rbind,
  lapply(c(0, 6, 12), function(h) effect_at_horizon(gls_fit, h))
)

knitr::kable(horizon_results, digits = 1)
```

The immediate contrast and later contrasts answer different questions. With only 23 post-law observations, long-horizon trend extrapolation deserves particular caution.

## Plot The Estimated Counterfactual

```{r}
#| label: fig-seatbelts-counterfactual
#| fig-cap: "Observed front-seat casualties and estimated trajectories with and without the law terms."
#| fig-height: 5.2
seatbelts$fitted_observed <- exp(predict(gls_fit, newdata = seatbelts))
no_law_data <- transform(seatbelts, law = 0, post_time = 0)
seatbelts$fitted_no_law <- exp(predict(gls_fit, newdata = no_law_data))

ggplot(seatbelts, aes(date)) +
  geom_line(aes(y = front, colour = "Observed"), linewidth = 0.55) +
  geom_line(aes(y = fitted_observed, colour = "Fitted with law"), linewidth = 0.75) +
  geom_line(
    aes(y = fitted_no_law, colour = "Projected no-law path"),
    linewidth = 0.75,
    linetype = "dashed"
  ) +
  geom_vline(xintercept = intervention_date, linetype = "dotted") +
  scale_colour_manual(values = c(
    "Observed" = "#303030",
    "Fitted with law" = "#24527a",
    "Projected no-law path" = "#bf6b21"
  )) +
  labs(x = NULL, y = "Monthly count", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

# Design And Robustness Lessons From Seatbelts

## Outcomes Are Not Interchangeable

`front` and `drivers` count people killed or seriously injured, whereas `DriversKilled` and `VanKilled` count deaths. Rear-seat casualties were not directly targeted by the 1983 front-seat requirement, but rear-seat occupancy may have changed after the law. Labels, exposure, and intervention status must be defended before comparing coefficients.

## Candidate Controls Need A Causal Argument

Rear-seat casualties can reveal whether the targeted series moved differently from a nominally untargeted group. They are not a clean control if passengers moved from front to rear seats or if exposure changed. Van-driver deaths have a weak pre-law relationship with front-seat passenger casualties and fall around the interruption themselves. Controlled ITS should therefore be presented as a diagnostic contrast, not an automatic correction [@lopezbernal2018controls].

## Contextual Variables Need A Role

Distance driven can represent exposure and petrol price can affect travel. Including them changes the estimand and assumes their relationship with casualties remains stable. Variables changed by the law should not be controlled away mechanically. Compare a short, prespecified sensitivity set and explain the assumption behind each row.

## Concurrent Changes Limit Attribution

The 1986 Parliamentary debate recorded contemporaneous changes in drink-driving enforcement and evidential breath testing as competing explanations for part of the 1983 casualty decline [@ukparliament1986seatbeltdebate]. Those processes are not measured in `Seatbelts`. A defensible summary is that targeted casualties fell relative to the projected path and to rear-seat casualties, while this dataset alone cannot isolate compulsory belt wearing from every simultaneous road-safety change.

# Stronger And More Complex Designs

## Controlled Interrupted Time Series

A credible untreated comparison can remove a genuinely shared shock. Estimate the differential level and trend changes using interactions, preserve separate residual processes where appropriate, and show the treated and comparison series before modelling [@lopezbernal2018controls].

## Count And Rate Outcomes

Counts may require Poisson or negative-binomial models, especially when values are small. Rates require a defensible denominator or offset. A log-transformed count model, as used above, is a teaching simplification rather than a universal default.

## Non-Linear, Lagged, Or Gradual Effects

Use ramps, splines, distributed lags, pulses, or temporary transitions when implementation knowledge supports them. Complexity should represent the delivery process rather than search for a favourable curve.

## Counterfactual Forecast Validation

When no comparison series is credible, compare candidate untreated models on held-out pre-intervention origins and the same reporting horizons. Select the primary model without post-intervention outcomes, retain a near-performing comparator, and stop when contextual predictors leave their training support [@bergmeir2018; @vanderweele2019].

# Common Failure Modes

- Calling a two-point before/after comparison ITS.
- Choosing the interruption date after inspecting the outcome.
- Coding post-intervention time from the start of the series rather than zero at intervention.
- Ignoring seasonality or serial dependence.
- Treating any correlated outcome as an unaffected control.
- Adding post-treatment variables as controls.
- Selecting a counterfactual model on post-intervention fit.
- Reporting only the immediate coefficient when the model also contains a trend change.
- Interpreting a prediction interval as causal confidence.
- Claiming that a stable estimate across models eliminates concurrent confounding.

# Reporting Checklist

Report:

1. outcome definition, frequency, population, and denominator;
2. intervention date and implementation evidence;
3. number and span of pre- and post-intervention observations;
4. model terms, seasonality, and dependence structure;
5. immediate and named horizon-specific effects with uncertainty;
6. comparison-series rationale, if used;
7. prespecified sensitivities and placebo checks;
8. concurrent events and measurement changes; and
9. the strongest causal statement the design can support.

# Software

Useful R tools include:

- `lm()` for transparent segmented-regression mechanics;
- `nlme::gls()` for generalized least squares with structures such as AR(1);
- `stats::arima()` for dynamic regression and intervention models;
- `sandwich` for heteroskedasticity- and autocorrelation-consistent covariance estimates; and
- `forecast` or `fable` for forecast-oriented extensions.

# Key Sources

- Segmented-regression foundations: @wagner2002segmented.
- Applied ITS tutorial: @lopezbernal2017tutorial.
- Controls in ITS: @lopezbernal2018controls.
- Correct segmented-regression parameterization: @xiao2021parameterization.
- Seatbelt application: @harvey1986seatbelts.
